---
title: "Alabama Pre-K tutorial: from the raw pipeline to reporting tables"
vignette: >
  %\VignetteIndexEntry{Alabama Pre-K tutorial: from the raw pipeline to reporting tables}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 2
    fig-width: 7
    fig-height: 5
    fig-align: center
    out-width: "85%"
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| eval: true

library(catchmentACS)
library(dplyr)
library(sf)
library(tidyr)

# Force network-free, deterministic builds: no provider/Census calls, no cache.
old_options <- options(
  catchmentACS.progress = "off",
  catchmentACS.rate_first_default = FALSE
)
old_cache <- getOption("catchmentACS.cache_enabled", NULL)
on.exit(options(old_options), add = TRUE)
on.exit(options(catchmentACS.cache_enabled = old_cache), add = TRUE)
cacs_set_cache(FALSE, scope = "session")
```

This is the package's long-form, flagship tutorial. Where
`vignette("getting-started", package = "catchmentACS")` shows the whole pipeline
in one call and `vignette("methodology", package = "catchmentACS")` explains the
math, this article does two things in sequence. First it takes the one call
*apart* — walking the five pipeline functions one stage at a time, reading each
intermediate result, then collapsing them back into a single `cacs_run()` and
confirming the two paths agree to the last digit. Then it carries the recomposed
result all the way to the *deliverables* an equity researcher actually hands
off: a site-by-drive-time rate matrix, a priority ranking, a single-site
profile, and a tidy table ready to write to CSV.

It is written for a Pre-K equity researcher who wants to understand each stage
well enough to describe it in a methods section, and who then needs to turn the
output into a memo, a dissertation chapter, or a journal table. It assumes you
have met `dplyr`, `sf`, isochrones, and margins of error before.

The motivating question comes from Pre-K funding-equity work: *for each First
Class Pre-K site, what are the neighborhood
conditions — poverty, food assistance, employment — within a short drive, and
how much do those conditions shift as the catchment grows from 5 to 15 minutes?*
`catchmentACS` is the reusable engine for that question. It replaces an earlier
set of one-off analysis scripts with a documented, regression-tested package, so
a finding can be reproduced — and a table re-generated — by a colleague who was
not in the room when it was first run. Stable, uncertainty-aware answers feed
directly into how the state reasons about where new Pre-K capacity should go.

::: {.callout-note}
## About the data in this article

Every table, map, and figure here is built entirely from the package's bundled
`inst/extdata` fixtures (`cacs_alabama_sites`, the synthetic `legacy_2025_*`
objects, and `sample_alabama_subset`), so this article renders with **no**
Census, OSRM, ORS, or other network calls. These fixtures preserve the
production calling contract (sites, drive-time polygons, tract-level ACS
variables) but are **synthetic and illustrative** — included only for examples,
tests, and reproducible package checks. They are **not** live Alabama policy
estimates and must not be cited as real ACS results. Any real analysis requires
a fresh live run (see `vignette("providers", package = "catchmentACS")`).
:::

## The data: Alabama Pre-K sites

A production run starts from a small table of program sites. The bundled
`cacs_alabama_sites` shows the shape that table needs: a stable site identifier,
a display name, a county FIPS code, a region label, and a single point geometry
in EPSG:4326. These ten sites are the live-run template — coordinates are
jittered and names are anonymized — so you can swap in your own administrative
site list with the same columns and nothing else changes.

```{r}
#| label: preview-sites
#| eval: true

cacs_alabama_sites |>
  sf::st_drop_geometry() |>
  select(site_id, site_name, county_fips, region_label)
```

```{r}
#| label: preview-sites-geometry
#| eval: true

sf::st_geometry(cacs_alabama_sites)
```

One bookkeeping note. `cacs_alabama_sites` is the live-run *template*; the
executable path in this article instead uses the interlocking `legacy_2025_*`
fixtures, because those carry the precomputed isochrones and the matching frozen
baseline that let the whole document render offline and deterministically. So
that the intermediate tables below stay small enough to read, the walkthrough
uses three legacy sites — one northern, one north-central, one northwest — at
drive times of 5, 10, and 15 minutes. The reasoning is identical for the full
ten-site batch.

```{r}
#| label: choose-sites
#| eval: true

site_ids <- c("AL_SITE_03", "AL_SITE_07", "AL_SITE_08")
```

## The pipeline, one stage at a time

`cacs_run()` orchestrates five functions in sequence. The first two reach the
network; the last three are pure local computation on whatever those two
returned:

> `cacs_acs_prefetch()` (ACS download) → `cacs_isochrone()` (routing) →
> `cacs_intersect_weight()` → `cacs_propagate_moe()` → `cacs_derive_rates()`.

This section runs each stage on its own so you can see the object that moves
between them — and so you can lift any stage out as a standalone manuscript
figure.

### Stages 1 and 2 are network operations (shown, not run)

In a live run you would fetch statewide ACS tracts and draw drive-time
isochrones like this. Both forms are shown for reference and are **not executed**
here, because each would make a network request:

```{r}
#| label: stage1-acs-live
#| eval: false

# Live form (needs a Census API key and network access):
acs <- cacs_acs_prefetch(state = "AL", year = 2023)
```

```{r}
#| label: stage2-iso-live
#| eval: false

# Live form (needs a routing provider; the OSRM public demo needs no key):
iso <- cacs_isochrone(
  sites       = cacs_alabama_sites,
  drive_times = c(5, 10, 15),
  provider    = "osrm",
  osrm_mode   = "demo"
)
```

`cacs_acs_prefetch()` returns an `sf` tibble of tract-level ACS estimates
(`GEOID`, `NAME`, `variable`, `estimate`, `moe`, `geometry`), and
`cacs_isochrone()` returns one drive-time polygon per `(site_id, drive_time_min)`
pair. The two bundled fixtures used below are exactly those return shapes,
frozen: `sample_alabama_subset.rds` stands in for the ACS prefetch, and
`legacy_2025_isochrones.rds` stands in for the isochrones. Loading them lets us
run the rest of the pipeline with no network at all.

```{r}
#| label: load-fixtures
#| eval: true

# These are exactly what stages 1 and 2 would have returned, frozen to disk.
sites <- readRDS(system.file(
  "extdata", "legacy_2025_sites.rds", package = "catchmentACS"
))
acs <- readRDS(system.file(
  "extdata", "sample_alabama_subset.rds", package = "catchmentACS"
))
iso <- readRDS(system.file(
  "extdata", "legacy_2025_isochrones.rds", package = "catchmentACS"
))

# Restrict to our three-site working batch and three drive-time bands.
analysis_sites <- sites |>
  filter(site_id %in% site_ids)
iso <- iso |>
  filter(site_id %in% site_ids, drive_time_min %in% c(5L, 10L, 15L))

list(
  acs_dim         = dim(acs),
  acs_crs         = sf::st_crs(acs)$epsg,
  acs_variables   = length(unique(acs$variable)),
  acs_tracts      = length(unique(acs$GEOID)),
  iso_dim         = dim(iso),
  iso_crs         = sf::st_crs(iso)$epsg
)
```

The ACS fixture carries `r length(unique(readRDS(system.file("extdata", "sample_alabama_subset.rds", package = "catchmentACS"))$variable))`
source variables across the Alabama tracts in EPSG:4269; the isochrone fixture
is a set of drive-time polygons in EPSG:4326. Those are the two coordinate
systems `cacs_intersect_weight()` expects on input.

### Stage 3: spatial intersection and area weighting

This is the algorithmic heart of the package. `cacs_intersect_weight()` overlays
each isochrone on the ACS tracts, computes how much of each tract falls inside
the catchment, and aggregates the source ACS counts by that overlap. It runs
fully offline:

```{r}
#| label: stage3-intersect
#| eval: true

weighted <- cacs_intersect_weight(
  iso_sf        = iso,
  acs_sf        = acs,
  weight_method = "area",
  verbose       = FALSE
)

dim(weighted)
```

```{r}
#| label: stage3-read
#| eval: true

weighted |>
  select(site_id, drive_time_min, variable, estimate, weight_sum, n_tracts) |>
  arrange(site_id, drive_time_min, variable) |>
  head(8)
```

Two columns are worth reading closely. `n_tracts` is how many tracts the
catchment touched — your sample thickness. `weight_sum` is the sum of the tract
area weights that went into the aggregate; each weight is the fraction of a
tract's area lying inside the catchment,
$\text{coverage\_wt}_j = \text{area}(\text{catchment} \cap \text{tract}_j) / \text{area}(\text{tract}_j)$.
Because catchments slice through tract boundaries, `weight_sum` is generally not
a whole number — a catchment covering most of three tracts and the edge of a
fourth might sum to something like 3.4. That is the answer to a reviewer who
asks why the effective tract count is not an integer: tracts enter partially, in
proportion to how much of their area the drive-time area covers.

The "Estimand families and weights" section of `?cacs_intersect_weight` documents
the full weighting scheme. In short, count variables (`spatial_total`) use this
coverage weight directly, while scalar and proxy variables (a median-income
proxy, say) use a separate normalized mean weight that sums to one within a
cell. The function also attaches an internal *carrier* table — the
coverage-weighted numerator and denominator totals — that the next two stages
read.

### Stage 4: margin-of-error propagation

ACS estimates ship with published margins of error (MOE), and those must travel
through the weighting. `cacs_propagate_moe()` attaches an MOE to every row,
choosing the formula appropriate to each variable's estimand family. It is pure
computation:

```{r}
#| label: stage4-moe
#| eval: true

with_moe <- cacs_propagate_moe(weighted, verbose = FALSE)

with_moe |>
  select(site_id, drive_time_min, variable, estimate, moe,
         moe_formula_effective) |>
  arrange(site_id, drive_time_min, variable) |>
  head(8)
```

A margin of error is the half-width of a 90% confidence interval — you report a
value as `estimate ± moe`, exactly as the Census Bureau reports ACS figures.
This stage does not recompute the point estimates; it fills in `moe` and records,
per row, which formula was used (`moe_formula_effective`) and whether any
fallback was needed (`moe_fallback`, `moe_fallback_reason`). We do not re-derive
those formulas here — the four families and the never-silent fallback rule are
documented in `?cacs_propagate_moe` and
`vignette("methodology", package = "catchmentACS")`.

One diagnostic habit: a non-trivial `moe_fallback_reason` is an audit signal, not
an error. It records that a requested formula could not produce a value for a
cell (for example a zero denominator or a missing input MOE) so the run stays
transparent. In this intentionally clean mini-batch no fallback fires:

```{r}
#| label: stage4-fallback-check
#| eval: true

with_moe |>
  filter(moe_fallback_reason != "n/a" | failure_origin != "none") |>
  nrow()
```

### Stage 5: derive the curated rates

Finally, `cacs_derive_rates()` appends the five curated catchment rates as new
`derived_rate` rows beneath the source-variable rows, each as a ratio of
area-weighted ACS totals with its own propagated MOE:

```{r}
#| label: stage5-rates
#| eval: true

final <- suppressWarnings(cacs_derive_rates(with_moe, verbose = FALSE))

dim(final)
```

```{r}
#| label: stage5-read
#| eval: true

final |>
  filter(estimand_family == "derived_rate") |>
  select(site_id, drive_time_min, variable, estimate, moe,
         moe_formula_effective) |>
  arrange(site_id, drive_time_min, variable) |>
  head(10)
```

Each rate maps to an ACS numerator/denominator pair and to a policy reading:

| Rate | ACS source | Policy reading |
|---|---|---|
| `poverty_rate` | `B17001_002 / B17001_001` | Catchment poverty — a demand-side proxy for Pre-K need |
| `snap_rate` | `B22003_002 / B22003_001` | SNAP receipt — an administrative cross-check on household income stress |
| `ssi_rate` | `B19056_002 / B19056_001` | Households with Supplemental Security Income — an equity dimension for disability/elderly-caregiver households |
| `unemp_rate` | `B23025_005 / B23025_003` | Unemployment — parental labor-market attachment |
| `labor_force_participation` | `B23025_002 / B23025_001` | Labor force participation — a supply-side (childcare-demand) proxy |

The MOE formula for each rate is selected by `cacs_derive_rates()`'s
`formula_dispatch` argument. Under the package's baseline policy, `poverty_rate`
and `labor_force_participation` are C1-eligible (the `proportion_subset` form,
valid when the numerator is a true subset of the denominator), while
`snap_rate`, `ssi_rate`, and `unemp_rate` are routed to C2
(`general_ratio_conservative`). A C1 rate whose subset variance comes out
negative falls back to C2 for that cell, with the override recorded in
`moe_fallback`, `moe_fallback_reason`, and `moe_formula_effective`. The full
dispatch contract lives in `?cacs_propagate_moe` and `?cacs_derive_rates`.

## Recompose into one call

The five stages above are exactly what `cacs_run()` does internally. Supplying
both `precomputed_isochrones =` and `acs =` tells it to skip the two network
stages and use the fixtures we loaded, so this one call reproduces the manual
chain offline — no requests, no key:

```{r}
#| label: recompose-run
#| eval: true

result <- cacs_run(
  sites                  = analysis_sites,
  state                  = "AL",
  year                   = 2023,
  drive_times            = c(5, 10, 15),
  variables              = unname(cacs_acs_default_vars),
  provider               = "osrm",
  precomputed_isochrones = iso,
  acs                    = acs,
  weight_method          = "area",
  output                 = "long",
  verbose                = FALSE
)

class(result)
dim(result)
```

The provenance attribute confirms the path taken: both inputs were supplied, so
the orchestrator ran the 3-call path (validation, intersection/weighting, MOE
propagation, rate derivation, output assembly) with no network.

```{r}
#| label: recompose-provenance
#| eval: true

attr(result, "cacs_run_provenance") |>
  as.data.frame() |>
  select(execution_path, provider, state, year,
         weight_method, formula_dispatch, bypass_iso, bypass_acs)
```

And the result is identical to the hand-built chain — not merely close. Comparing
the two estimate columns on the same `(site_id, drive_time_min, variable)` keys:

```{r}
#| label: recompose-equivalence
#| eval: true

manual <- final |>
  arrange(site_id, drive_time_min, variable)
oneshot <- tibble::as_tibble(result) |>
  arrange(site_id, drive_time_min, variable)

c(
  same_nrow      = nrow(manual) == nrow(oneshot),
  max_estimate_d = max(abs(manual$estimate - oneshot$estimate), na.rm = TRUE),
  max_moe_d      = max(abs(manual$moe - oneshot$moe), na.rm = TRUE)
)
```

That zero is the point of the first half of this exercise: the readable
five-stage decomposition and the single production call are the same
computation. In daily work you write the one line; the five stages are what you
cite in a methods section. The rest of the article works from `result`.

## Reporting-ready outputs

This is the part that matters at hand-off. Everything below is a deliverable an
equity researcher can drop into a memo, a dissertation chapter, or a journal
submission — and each maps to a concrete reporting need.

### Select the analyst-facing columns

`result` is one tidy *long* table — one row per site, drive-time band, and
variable — carrying provenance and audit fields that are useful when you share
results and ignorable otherwise. Most daily analysis starts from a handful of
columns, so name them once and reuse the selection. Read `failure_origin` first:
`"none"` is your success flag for that cell.

```{r}
#| label: explore-columns
#| eval: true

result_tbl <- tibble::as_tibble(result, rate_first = FALSE)

report_cols <- c(
  "site_id", "drive_time_min", "variable",
  "estimate", "moe", "n_tracts", "failure_origin"
)

result_tbl |>
  select(all_of(report_cols)) |>
  head(12)
```

This batch is intentionally clean, so every rate succeeds. In a larger live run,
when a derived rate comes back `NA`, `failure_origin` names the stage that
produced the gap — an upstream ACS input, an isochrone failure, an empty
intersection, or a missing aggregation carrier — so you can diagnose rather than
guess. The `estimate` for a rate is a proportion: `0.173` means 17.3%.

### The site x drive-time rate matrix

The most compact deliverable is one row per site and drive time with all five
curated rates side by side. Pivoting the long table wide gives exactly the shape
of a manuscript table:

```{r}
#| label: rate-matrix
#| eval: true

rate_matrix <- result_tbl |>
  filter(
    variable %in% names(cacs_acs_default_rates),
    failure_origin == "none"
  ) |>
  select(site_id, drive_time_min, variable, estimate) |>
  pivot_wider(names_from = variable, values_from = estimate) |>
  arrange(site_id, drive_time_min)

rate_matrix
```

This is the analyst's primary worktable: every site at every catchment band,
with the five rates as columns. For a reader-facing version that pairs each rate
with its uncertainty, `summary()` reshapes the same numbers into one row per site
and drive time, each rate formatted as `estimate ± MOE`:

```{r}
#| label: rate-matrix-moe
#| eval: true

summary(result)$rates_per_site_moe
```

If a `<NA>` cell appears in a larger run, it is not a formatting error: it means
the package could not form a stable numerator/denominator pair for that rate in
that cell. Keep such cells visible in review tables rather than overwriting them
with zero — a missing rate is a diagnostic, not an absence of need.

### A 15-minute priority table

A common policy task is to rank sites by need at the widest catchment. Here we
sort the three sites by their 15-minute poverty rate, highest first — the kind of
ordering that drives a "where to invest next" conversation.

```{r}
#| label: priority-table
#| eval: true

priority_15 <- result_tbl |>
  filter(
    drive_time_min == 15,
    variable == "poverty_rate",
    failure_origin == "none"
  ) |>
  arrange(desc(estimate)) |>
  transmute(
    rank         = row_number(),
    site_id,
    poverty_rate = estimate,
    moe_90       = moe,
    n_tracts_num
  )

priority_15
```

The estimate stays a proportion; multiply by 100 only at the final formatting
step so any downstream arithmetic remains stable. Swap `poverty_rate` for
`snap_rate` or `unemp_rate` to rank on a different indicator, or change
`drive_time_min` to compare priorities at a tighter band.

### A focal-site profile

For a single-site memo the question is different: not "which site ranks highest?"
but "how does this one site change as its catchment grows from 5 to 15 minutes?"
The profile below shows all five curated rates across the three rings for one
focal site, each with its margin of error and the implied confidence interval.

```{r}
#| label: focal-site-profile
#| eval: true

focal_site <- "AL_SITE_03"

result_tbl |>
  filter(
    site_id == focal_site,
    variable %in% names(cacs_acs_default_rates),
    failure_origin == "none"
  ) |>
  arrange(variable, drive_time_min) |>
  transmute(
    variable,
    drive_time_min,
    estimate,
    moe_90  = moe,
    ci_low  = pmax(estimate - moe, 0),
    ci_high = estimate + moe
  )
```

Reporting each rate as `estimate ± moe_90` (or as the `[ci_low, ci_high]`
interval shown here) is the same convention the Census Bureau uses for ACS
figures, so a reviewer can read the catchment numbers exactly as they would read
a published table.

### An exportable report tibble

Finally, reduce the object to a clean, CSV-ready table. Keep the MOE, the failure
flag, and the effective MOE formula alongside the estimate so a reviewer can
trace both uncertainty and any fallback behavior.

```{r}
#| label: export-shape
#| eval: true

export_tbl <- result_tbl |>
  filter(variable %in% names(cacs_acs_default_rates)) |>
  transmute(
    site_id,
    drive_time_min,
    variable,
    estimate,
    moe_90                = moe,
    estimate_pct          = 100 * estimate,
    moe_90_pct            = 100 * moe,
    failure_origin,
    moe_formula_effective,
    moe_fallback
  ) |>
  arrange(site_id, drive_time_min, variable)

head(export_tbl, 15)
```

You would write this out with, for example,
`readr::write_csv(export_tbl, "alabama_prek_rates.csv")`. Before publishing any
narrative claim, filter or flag rows where `failure_origin != "none"`: a missing
rate may still be a useful diagnostic, but it must not be read as zero.

## Reproducibility and the frozen baseline

The package ships a frozen regression baseline, `legacy_2025_golden_output.rds`:
the full ten-site, three-band `cacs_run()` result captured once and committed to
`inst/extdata`. Its role is to anchor reproducibility — the test suite recomputes
the pipeline and verifies it against this baseline at a tight numerical
tolerance, so an accidental change in any stage's arithmetic is caught before
release.

```{r}
#| label: load-golden
#| eval: true

golden <- readRDS(system.file(
  "extdata", "legacy_2025_golden_output.rds", package = "catchmentACS"
))

list(
  class          = class(golden)[1],
  dim            = dim(golden),
  n_sites        = length(unique(golden$site_id)),
  drive_times    = sort(unique(golden$drive_time_min))
)
```

Its provenance records how it was produced — the same `cacs_run()` contract,
captured on the 3-call (fixtures-only) path:

```{r}
#| label: golden-provenance
#| eval: true

attr(golden, "cacs_run_provenance") |>
  as.data.frame() |>
  select(execution_path, state, year, weight_method,
         formula_dispatch, output_format, level)
```

What the guarantee means for you in practice: if you re-run the package on these
bundled inputs and the recomputed estimates match the baseline within tolerance,
your install reproduces the reference numbers exactly. That is the user-facing
form of the regression check the maintainers run in CI — you can port a 2025-era
analysis script onto this API and confirm the numbers did not move. A literal
row-by-row comparison is only meaningful when it runs in a clean, offline-stable
environment matching the one that produced the golden file (the same package
version and spatial-library stack). The three-site batch above is a deliberately
small slice for teaching, not the full-roster object the golden file records, so
this article shows the baseline's shape and provenance rather than asserting a
numeric diff; the faithful comparison belongs in the test suite, which performs
it at the locked tolerance.

## A map

With the optional `leaflet` package installed, you can see one site's 15-minute
catchment over its intersecting tracts. This chunk is gated on the namespace, so
it is simply skipped when `leaflet` is absent, and it calls `leaflet::` rather
than attaching the package.

```{r}
#| label: focal-map
#| eval: !expr requireNamespace("leaflet", quietly = TRUE)

# Focal site point and its 15-minute ring (both EPSG:4326).
focal_pt   <- analysis_sites[analysis_sites$site_id == focal_site, ]
focal_ring <- iso[iso$site_id == focal_site & iso$drive_time_min == 15, ]

# Build a tract-level poverty rate from the source ACS counts, then keep only
# the tracts the 15-minute ring touches. Everything stays offline.
suppressMessages(sf::sf_use_s2(FALSE))

pov <- acs |>
  sf::st_drop_geometry() |>
  filter(variable %in% c("B17001_001", "B17001_002")) |>
  select(GEOID, variable, estimate) |>
  tidyr::pivot_wider(
    id_cols = GEOID, names_from = variable, values_from = estimate
  ) |>
  transmute(GEOID, poverty_rate = B17001_002 / B17001_001)

tract_geom <- acs |>
  filter(variable == "B17001_001") |>
  select(GEOID) |>
  left_join(pov, by = "GEOID") |>
  sf::st_transform(4326)

hit <- tract_geom[
  sf::st_intersects(tract_geom, focal_ring, sparse = FALSE)[, 1],
]

pal <- leaflet::colorNumeric("YlOrRd", domain = hit$poverty_rate)

leaflet::leaflet(hit) |>
  leaflet::addProviderTiles("CartoDB.Positron") |>
  leaflet::addPolygons(
    fillColor = ~pal(poverty_rate), fillOpacity = 0.6,
    weight = 0.5, color = "#666666",
    label = ~sprintf("Tract %s: %.1f%%", GEOID, 100 * poverty_rate)
  ) |>
  leaflet::addPolygons(
    data = focal_ring, fill = FALSE,
    weight = 2, opacity = 0.8, color = "#c05621"
  ) |>
  leaflet::addCircleMarkers(
    data = focal_pt, radius = 5, color = "#2b6cb0",
    fillOpacity = 0.9, stroke = FALSE, popup = ~site_id
  ) |>
  leaflet::addLegend(
    pal = pal, values = ~poverty_rate,
    title = "Tract poverty rate",
    labFormat = leaflet::labelFormat(transform = function(x) 100 * x, suffix = "%")
  )
```

The blue dot is the site, the orange outline is its 15-minute drive-time area,
and the shaded tracts are everything that area touches, colored by tract poverty
rate. This is the per-tract picture that area weighting collapses into a single
catchment estimate. For a deeper map-based walkthrough, see
`vignette("visual-walkthrough", package = "catchmentACS")`.

## Transition to a live run

To move from this offline walkthrough to current Alabama data, drop the two
bypass arguments. The same `cacs_run()` call then draws drive-time areas from the
selected provider and pulls ACS variables through `tidycensus`:

```{r}
#| label: live-transition
#| eval: false

# One-time Census key setup:
tidycensus::census_api_key("YOUR_KEY_HERE", install = TRUE)

live_result <- cacs_run(
  sites         = cacs_alabama_sites,   # the 10 bundled sites, or your own roster
  state         = "AL",
  year          = 2023,
  drive_times   = c(5, 10, 15),
  variables     = unname(cacs_acs_default_vars),
  provider      = "osrm",
  weight_method = "area",
  output        = "long",
  verbose       = TRUE
)
```

A few notes for the live form:

- **Census key and ACS years.** This release supports ACS 5-year end years
  2009–2024 (2024 is the most recent vintage). A Census key is needed when the
  ACS request is a live cache miss or `force_refresh = TRUE`; a warm ACS cache
  hit, or supplying an explicit `acs =` object, needs no key.
- **Weighting.** `weight_method = "area"` is the implemented method in this
  release. `weight_method = "population"` is a reserved name that fails loud at
  the orchestrator boundary before any network or spatial work begins — it is a
  planned future extension, not an active path.
- **Routing providers.** `"osrm"` is implemented (the public demo endpoint needs
  no key but can be rate-limited), `"ors"` is implemented when the optional
  `openrouteservice` package and an API key are present, and `"mapbox"` and
  `"r5r"` are reserved fail-loud provider names.

## Where to next

- **The first run, in ten minutes**:
  `vignette("getting-started", package = "catchmentACS")`.
- **The math** behind area weighting and MOE propagation:
  `vignette("methodology", package = "catchmentACS")`.
- **Routing providers and credentials** — OSRM (default), ORS (optional, needs a
  key), and the reserved Mapbox/r5r surfaces, plus the offline bypass:
  `vignette("providers", package = "catchmentACS")`.
- **Map-based inspection of one site**:
  `vignette("visual-walkthrough", package = "catchmentACS")`.
- **Every `cacs_run()` argument in detail**: `?cacs_run`. The pipeline functions
  are `?cacs_acs_prefetch`, `?cacs_isochrone`, `?cacs_intersect_weight`,
  `?cacs_propagate_moe`, and `?cacs_derive_rates`.

Bug reports and feature requests:
<https://github.com/joonho112/catchmentACS/issues>.
