---
title: "Methodology: the pipeline as an estimator"
vignette: >
  %\VignetteIndexEntry{Methodology: the pipeline as an estimator}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 2
    fig-width: 7
    fig-height: 5
    fig-align: center
    out-width: "85%"
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| eval: true

library(catchmentACS)
library(dplyr)
library(sf)

# Force network-free, deterministic builds: no provider/Census calls, no cache.
old_options <- options(
  catchmentACS.progress = "off",
  catchmentACS.rate_first_default = FALSE
)
old_cache <- getOption("catchmentACS.cache_enabled", NULL)
on.exit(options(old_options), add = TRUE)
on.exit(options(catchmentACS.cache_enabled = old_cache), add = TRUE)
cacs_set_cache(FALSE, scope = "session")
```

This article is the entry point to the package's **method track**. It frames the
whole `catchmentACS` pipeline as a single statistical *estimator*, fixes the
**notation** that the deeper theory articles reuse, names the six estimand
families the package is built around, and shows — in real numbers, on the
bundled fixtures — why a count and an average are aggregated with *different*
weights. Where the applied articles answer "how do I run this?", this one
answers "what, exactly, is being estimated, and by what formula?"

It is deliberately math-light and does **not** re-derive the math: each estimand
family and each margin-of-error (MOE) formula has a full derivation in a
dedicated theory article, and this overview states the result and points there.
The three companion articles are
`vignette("theory-spatial-aggregation", package = "catchmentACS")`,
`vignette("theory-moe-propagation", package = "catchmentACS")`, and
`vignette("theory-derived-rates", package = "catchmentACS")`. By the end you
should be able to read any catchment estimate as
$\text{estimate} \pm \text{MOE}$ and know which of six families produced it.

If you want to *run* the pipeline rather than read its math, start with
`vignette("getting-started", package = "catchmentACS")` and
`vignette("alabama-tutorial", package = "catchmentACS")`.

::: {.callout-note}
## About the data in this article

Every executable cell here is built entirely from the package's bundled
`inst/extdata` fixtures — the synthetic `legacy_2025_*` objects and the
`sample_alabama_subset` ACS extract — so this article renders with **no** Census,
OSRM, ORS, or other network calls. These fixtures preserve the production calling
contract (sites, drive-time polygons, tract-level ACS variables) but are
**synthetic and illustrative**: they are included only for examples, tests, and
reproducible package checks. They are **not** live Alabama policy estimates and
must not be cited as real ACS results. Any real analysis requires a fresh live
run (see `vignette("providers", package = "catchmentACS")`).
:::

## The pipeline is an estimator

It helps to read the whole pipeline as one estimator. `cacs_run()` looks like a
data-processing wrapper, but statistically it is one estimator with five stages.
Two stages reach data sources; the last three are deterministic computation on
whatever those returned:

> `cacs_acs_prefetch()` (ACS download) → `cacs_isochrone()` (routing) →
> `cacs_intersect_weight()` → `cacs_propagate_moe()` → `cacs_derive_rates()`.

The estimand is a neighborhood characteristic *inside a drive-time catchment*.
The ACS publishes its estimates on **census tracts**, which do not line up with
catchments — a drive-time area slices through tract boundaries. So the estimator
re-expresses tract-level published values as catchment-level values by
**area-weighted aggregation**, then carries the published uncertainty through
that aggregation. Concretely, for each site $s$:

1. **Sites → catchments.** Take the drive-time isochrone polygon $I_s$ (drawn
   fresh, or reused from precomputed polygons) and the ACS tract polygons $T_j$.
2. **Catchments → tract overlaps.** Cut the intersections $I_s \cap T_j$ on an
   equal-area projection, and from each intersection build a *weight* describing
   how much of tract $j$ the catchment captures.
3. **Overlaps → estimates.** Aggregate each tract-level estimate $Y_j$ (or
   $X_j$) to the catchment with the weight that matches the variable's *estimand
   family*.
4. **Estimates → margins of error.** Propagate the published tract-level MOEs
   through the same weights.
5. **Estimates → rates.** Form the curated derived rates as ratios of aggregated
   numerators and denominators, each with its own propagated MOE.

The single function `cacs_run()` chains these stages; the five pipeline
functions expose each one on its own. Two choices in steps 2–3 carry the entire
statistical content of the package, and they are the reason the rest of this
article exists: **which weight**, and **which family**. Get either wrong and the
arithmetic still runs — it just estimates the wrong thing, silently.

### Shared notation

The whole method track uses the same symbols. Establishing them once here lets
the three theory articles stay terse.

| Symbol | Meaning |
|---|---|
| $I_s$ | the drive-time isochrone (catchment) polygon for site $s$ (a cumulative ring) |
| $T_j$ | census tract $j$ |
| $I_s \cap T_j$ | the part of tract $j$ that falls inside the catchment |
| $\lvert\,\cdot\,\rvert$ | equal-area measure (area on EPSG:5070, in $\text{m}^2$) |
| $Y_j$ | a tract-level **count** ACS estimate (extensive: persons, households) |
| $X_j$ | a tract-level **scalar / proxy** ACS estimate (intensive: a median, a per-capita figure) |
| $A_j, B_j$ | a ratio's numerator and denominator counts in tract $j$ |
| $M_j$ | the published 90% margin of error for the tract-level estimate |

Two weights do all the work. The **coverage weight** is the fraction of a
tract's *area* captured by the catchment,

$$
w^{cov}_{sj} = \frac{\lvert I_s \cap T_j \rvert}{\lvert T_j \rvert},
$$

and the **normalized mean weight** is the catchment's intersection area for tract
$j$ as a share of *all* its intersection area,

$$
w^{mean}_{sj} = \frac{\lvert I_s \cap T_j \rvert}{\sum_k \lvert I_s \cap T_k \rvert},
\qquad \sum_j w^{mean}_{sj} = 1,
$$

so the mean weights sum to one within a site and drive-time cell. From these,
the three estimator forms are

$$
\widehat{Y}_s = \sum_j w^{cov}_{sj}\, Y_j,
\qquad
\widehat{\bar X}_s = \sum_j w^{mean}_{sj}\, X_j,
\qquad
\widehat{R}_s = \frac{\sum_j w^{cov}_{sj}\, A_j}{\sum_j w^{cov}_{sj}\, B_j}.
$$

A **spatial total** allocates counts into the catchment with the coverage
weight; a **scalar/proxy mean** blends intensive values with the normalized mean
weight; a **derived rate** aggregates the numerator and denominator *each as a
coverage total*, then divides. Throughout, standard errors and margins of error
are interconvertible through the Census multiplier at the 90% level,

$$
M = 1.645\, SE, \qquad SE = \frac{M}{1.645},
$$

and every margin of error in the pipeline is reported at that level: variances
accumulate on the SE scale and convert back at the end.

## The six estimand families at a glance

A variable's *estimand family* decides which weight aggregates it and which MOE
path applies. This is the central methodological point of the package, and the
spatial aggregation algorithm is built around it: a count and a median are
aggregated by **different** weights, and using the wrong one is a silent
statistical error, not a runtime failure.

| Estimand family | Example variable | Point estimate | Weight |
|---|---|---|---|
| `spatial_total` | total population; persons in poverty (`B17001_002`) | $\widehat{Y}_s = \sum_j w^{cov}_{sj} Y_j$ | coverage $w^{cov}_{sj}$ |
| `derived_rate` | poverty rate; SNAP rate | $\widehat{R}_s = \dfrac{\sum_j w^{cov}_{sj} A_j}{\sum_j w^{cov}_{sj} B_j}$ | coverage (numerator and denominator separately) |
| `area_weighted_scalar_proxy` | per-capita income (`B19301_001`) | $\widehat{\bar X}_s = \sum_j w^{mean}_{sj} X_j$ | normalized mean $w^{mean}_{sj}$ |
| `population_weighted_scalar_proxy` *(reserved)* | scalar proxy under population weighting | $\widehat{\bar X}_s = \sum_j w^{mean}_{sj} X_j$ | population-normalized mean |
| `median_proxy` | tract median household income (`B19013_001`) | weighted blend of tract medians, **not** the true catchment median | normalized mean $w^{mean}_{sj}$ |
| `metadata_only` | provenance fields | no numeric aggregation | none |

Read the table by weight. An extensive **count** like *persons in poverty* must
be *summed* into the catchment with the **coverage weight** — half a tract
contributes half its people. An intensive **scalar** like *per-capita income*
must be *averaged* with the **normalized mean weight** — half a tract does not
contribute "half a dollar per person," so the contributions average rather than
accumulate. Derived rates are a special case: numerator and denominator are each
aggregated as coverage-weighted totals, then divided, so the rate inherits the
coverage path. The `median_proxy` family is labelled a *proxy* on purpose — a
weighted blend of tract medians is not the true median of the pooled catchment
population, and the package keeps that caveat attached to the estimate.

Add an intensive variable as if it were a count, or divide a count by a weight
sum as if it were an average, and you get a number with no defensible
interpretation. Nothing errors; the result is just wrong. The family classifier
is what stops that from happening — and the next section shows the two paths
diverging on actual fixture numbers.

Two families are scaffolding rather than active estimators in this release. In
`weight_method = "area"` (the implemented path), the area forms of the weights
above are active. `population_weighted_scalar_proxy` and the population mean
weight are a reserved future surface; `weight_method = "population"` fails loud
at the orchestrator boundary today. `metadata_only` rows carry provenance and
receive no numeric aggregation. The classifier and the per-family aggregation
are documented in the "Estimand families and weights" section of
`?cacs_intersect_weight`.

## A worked example: coverage total vs mean proxy

This is the section to read slowly. We take **one** site and **one** ring from
the fixtures, look at the weights the package builds, and then watch a count and
an average go through their two different aggregation paths — so the abstract
$w^{cov}$-vs-$w^{mean}$ distinction lands as real numbers. The same contract
scales to many sites and drive times.

We use the synthetic site `AL_SITE_03` at a 15-minute drive time, whose
catchment intersects three tracts. First, intersect and keep the per-tract audit
so we can see the geometry ingredients:

```{r}
#| label: we-intersect
#| eval: true

# These fixtures are exactly the shapes stages 1-2 would have returned, frozen.
sites <- readRDS(system.file(
  "extdata", "legacy_2025_sites.rds", package = "catchmentACS"
))
iso <- readRDS(system.file(
  "extdata", "legacy_2025_isochrones.rds", package = "catchmentACS"
))
acs <- readRDS(system.file(
  "extdata", "sample_alabama_subset.rds", package = "catchmentACS"
))

site_id    <- "AL_SITE_03"
drive_time <- 15L

iso_one <- iso[
  iso$site_id == site_id & iso$drive_time_min == drive_time, ,
  drop = FALSE
]

weighted <- cacs_intersect_weight(
  iso_sf           = iso_one,
  acs_sf           = acs,
  weight_method    = "area",
  keep_tract_audit = TRUE,
  verbose          = FALSE
)
```

The tract audit holds, for each tract the catchment touched, the intersection
area and the tract area. From those two columns the package forms both weights.
Computing them side by side makes the difference concrete:

```{r}
#| label: we-weights
#| eval: true

audit <- attr(weighted, "cacs_tract_audit") |>
  transmute(
    GEOID,
    int_area_m2,
    tract_area_m2,
    w_cov  = int_area_m2 / tract_area_m2,            # coverage weight
    w_mean = int_area_m2 / sum(int_area_m2)          # normalized mean weight
  ) |>
  arrange(desc(int_area_m2))

audit

c(
  sum_w_cov  = sum(audit$w_cov),    # NOT constrained to 1
  sum_w_mean = sum(audit$w_mean)    # constrained to 1 by construction
)
```

Read the two sums. The coverage weights $w^{cov}_{sj}$ each measure *how much of
that tract* the catchment captured, so they do **not** sum to one — here they
sum to the effective number of tract-equivalents the catchment covers. The mean
weights $w^{mean}_{sj}$ are renormalized to sum to exactly one. That single
structural difference is what separates a *total* from an *average*.

Now send a count through the coverage path. *Persons in poverty* (`B17001_002`)
is a `spatial_total`, so the package reports
$\widehat{Y}_s = \sum_j w^{cov}_{sj}\,Y_j$ with `weight_basis = "coverage"`:

```{r}
#| label: we-total
#| eval: true

weighted |>
  filter(variable == "B17001_002") |>
  select(variable, estimate, moe, weight_sum, n_tracts,
         estimand_family, weight_basis)
```

The `weight_sum` column is exactly $\sum_j w^{cov}_{sj}$ — the same
not-equal-to-one sum from the audit — and `estimate` is the coverage-weighted
count of people in poverty across the three partial tracts. That is a total: it
scales with how much area the catchment captures.

Now send an *average* through. *Per-capita income* (`B19301_001`) is an
`area_weighted_scalar_proxy`: an intensive, per-person quantity that must be
*blended*, not summed. Its estimator is the normalized-mean-weighted average
$$\widehat{\bar X}_s = \sum_j w^{mean}_{sj}\,X_j,$$
whose weights sum to one within a site and drive-time cell, so the result stays
on the per-person scale and lies between the contributing tracts' own values.
(For the implementation see `?cacs_intersect_weight`; for the full derivation see
`vignette("theory-spatial-aggregation", package = "catchmentACS")`.)

The teaching moment is what happens if you ignore the family and aggregate this
intensive variable *as if it were a count*. Below we pull the three tracts'
published per-capita incomes and apply **both** weights by hand, so the contrast
is self-contained and exact:

```{r}
#| label: we-silent-error
#| eval: true

pc_income <- acs |>
  sf::st_drop_geometry() |>
  filter(GEOID %in% audit$GEOID, variable == "B19301_001") |>
  select(GEOID, X = estimate)

demo <- audit |>
  select(GEOID, w_cov, w_mean) |>
  left_join(pc_income, by = "GEOID")

demo

c(
  coverage_weighted_SUM = sum(demo$w_cov  * demo$X),   # wrong estimand for an average
  mean_weighted_PROXY   = sum(demo$w_mean * demo$X)    # correct: stays on the $/person scale
)
```

The two numbers tell the whole story. The **coverage-weighted sum** is `80169` —
a dollar amount inflated past any single tract's value, because summing an
intensive quantity is meaningless. It is not an income anyone earns. The
**mean-weighted proxy** is `26723`, which stays on the per-person scale, between
the tracts' own values, which is what a catchment "average per-capita income"
should be. Both used identical geometry; only the *family-appropriate weight*
differs. This is the silent estimand error the family classifier exists to
prevent, and it is why `catchmentACS` refuses to apply one weight to everything.

The MOE column moved with the estimate in both family rows above. How it is
computed is the subject of the next section, in brief.

## Margins of error in brief

ACS estimates ship with a published 90% margin of error, and that uncertainty
must travel through the aggregation. `cacs_propagate_moe()` attaches an MOE to
every row by dispatching on the estimand family. There are **four** MOE formula
families; this article only names them and states the conversion.

- **Family A — weighted sum** (for `spatial_total`): the root-sum-square of the
  coverage-weighted component MOEs.
- **Family B — weighted mean** (for the `*_scalar_proxy` and `median_proxy`
  families): the same root-sum-square form with the normalized mean weights,
  which sum to one and are treated as fixed.
- **Family C1 — proportion subset** (for a `derived_rate` whose numerator $A$ is
  a true subset of its denominator $B$): the cross term is *subtracted*.
- **Family C2 — general ratio, conservative** (the package's default for the
  curated derived rates): the cross term is *added*, so the variance is always
  non-negative.

All four sit on the same conversion, $M = 1.645\, SE$: each component MOE is
divided by $1.645$ to a variance, the variances are combined, and the result is
multiplied back by $1.645$.

The one branch worth flagging here is the **C1 → C2 fallback**, and that it is
never silent. When a Family C1 (`proportion_subset`) calculation yields a
negative subset variance, the row is recomputed with the Family C2
(`general_ratio_conservative`) formula so a value is still produced, and the
override is recorded per row: `moe_formula_effective` becomes
`"general_ratio_conservative"`, `moe_fallback` is set to `TRUE`, and
`moe_fallback_reason` records the cause. A single summary warning reports the
counts at the end of a run.

We do **not** derive these formulas here — including why variances accumulate on
the SE scale and why the conservative ratio is the default. The full
Family A/B/C1/C2 algebra, the independence assumptions, and the numerical guards
live in `vignette("theory-moe-propagation", package = "catchmentACS")` and
`?cacs_propagate_moe`.

## Projection and the within-tract assumption

Two modeling choices underpin every number above.

**Equal-area measurement on EPSG:5070.** The coverage weight
$w^{cov}_{sj} = \lvert I_s \cap T_j \rvert / \lvert T_j \rvert$ is a ratio of two
areas, so it is only meaningful if numerator and denominator are measured in the
*same* equal-area metric. The package transforms validated inputs to **EPSG:5070
(NAD83 / CONUS Albers Equal Area)** once before measuring any area, so equal
physical areas carry equal weight regardless of where in the catchment they
fall. This is also why the current release is scoped to the contiguous United
States plus DC, where Albers distortion is negligible.

**Uniform-within-tract allocation.** Area weighting assumes the ACS
characteristic is spread *uniformly* over each tract: a catchment that covers 30%
of a tract's area is credited with 30% of its count. This is exact for area but
only approximate for people, and it is the standard areal-interpolation
assumption that a future `weight_method = "population"` mode is designed to
relax.

Both are stated only briefly here; the equal-area rationale and the
uniform-within-tract assumption are argued in full in
`vignette("theory-spatial-aggregation", package = "catchmentACS")`.

## Where the details live

This article is the map; the derivations are in the three method-track articles
and the function reference, each reusing the notation fixed above.

- **Spatial aggregation** — the coverage vs mean weight derivation, the
  per-site intersection algorithm, the full six-family aggregation table,
  EPSG:5070, and the within-tract assumption:
  `vignette("theory-spatial-aggregation", package = "catchmentACS")` and
  `?cacs_intersect_weight`.
- **MOE propagation** — the four formula families (A weighted sum, B weighted
  mean, C1 proportion subset, C2 conservative general ratio), the SE ↔ MOE
  relation, the never-silent C1 → C2 fallback, and the numerical guards:
  `vignette("theory-moe-propagation", package = "catchmentACS")` and
  `?cacs_propagate_moe`.
- **Derived rates** — the five curated rates, their ACS numerator/denominator
  pairs, the baseline C1/C2 policy, and the carrier-lookup mechanics:
  `vignette("theory-derived-rates", package = "catchmentACS")` and
  `?cacs_derive_rates`.

For the record, each of the five curated rates maps to a fixed ACS numerator and
denominator. `ssi_rate` is `B19056_002 / B19056_001` — households with
Supplemental Security Income over the `B19056` household universe — and under the
package's baseline policy `poverty_rate` and `labor_force_participation` are
C1-eligible (`proportion_subset`), while `snap_rate`, `ssi_rate`, and
`unemp_rate` route to C2 (`general_ratio_conservative`). The full catalogue and
dispatch rationale are in
`vignette("theory-derived-rates", package = "catchmentACS")`.

For the applied, run-it path that produces reporting tables from these same
fixtures, see `vignette("getting-started", package = "catchmentACS")`,
`vignette("alabama-tutorial", package = "catchmentACS")`, and
`vignette("providers", package = "catchmentACS")`.

## Limitations

The estimator's scope is deliberately narrow in this release.

- **Area weighting only.** `weight_method = "area"` is implemented; the
  population-weighted scalar-proxy family and `weight_method = "population"` are
  a reserved fail-loud future surface. The uniform-within-tract assumption is
  exact for area, approximate for population.
- **Median is a proxy.** A `median_proxy` value is a weighted blend of tract
  medians, not the true median of the pooled catchment population; it is labelled
  as a proxy throughout.
- **Conservative ratio MOE by default.** Derived-rate MOEs default to the Family
  C2 general-ratio form; the tighter C1 subset form is opt-in for the C1-eligible
  rates and falls back to C2 (never silently) when its variance goes negative.
- **A fixed rate catalogue.** The five curated rates are locked; arbitrary
  user-supplied ratios are deferred to a future release.
- **CONUS + DC.** EPSG:5070 is the authoritative equal-area default for the
  contiguous United States and DC; the data path is ACS 5-year tract data for one
  state at a time, and other regions await a future CRS policy.
- **Fixtures.** The bundled fixtures are synthetic and regression-oriented —
  useful for examples and package checks, not a substitute for a fresh live run
  when producing final analysis.

The interpretive caution that frames all of the above: an isochrone's polygon
area is provider- and snapshot-sensitive, so for policy reporting the tract
membership, the weighted estimates, their margins of error, and the recorded
provenance are more informative audit surfaces than raw polygon area alone. And,
as the callout at the top stresses, the fixtures used here are synthetic — the
numbers above illustrate the *method*, not Alabama.
