---
title: "Porting catchmentACS workflows to v0.5"
vignette: >
  %\VignetteIndexEntry{Porting catchmentACS workflows to v0.5}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 2
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| eval: true

library(catchmentACS)
library(dplyr)
library(sf)

# Force network-free, deterministic builds: no provider/Census calls, no cache.
old_options <- options(
  catchmentACS.progress = "off",
  catchmentACS.rate_first_default = FALSE
)
old_cache <- getOption("catchmentACS.cache_enabled", NULL)
on.exit(options(old_options), add = TRUE)
on.exit(options(catchmentACS.cache_enabled = old_cache), add = TRUE)
cacs_set_cache(FALSE, scope = "session")
```

This vignette collects the five changes you need to know when moving a v0.4
script to v0.5. **Good news first**: v0.5 is a beta-readiness release of
safety and clarity refinements, not breaking changes. Almost every v0.4 script
runs unmodified. The `cacs_run()` contract, the five sanctioned rates, the
canonical output schema, and the area-weighting math are all unchanged; the
changes below mostly make existing behavior more complete or more honest about
its limits.

For a full applied walkthrough see `vignette("getting-started")` and
`vignette("alabama-tutorial")`; for routing-provider details see
`vignette("providers")`; and for the prior step see
`vignette("porting-v03-to-v04")`.

## Quick map

| Area | v0.4 behavior | v0.5 behavior |
|---|---|---|
| `$isochrone` list-column fill (`output = "list_column"` / `"both"`) | populated on bypass paths only | populated on **all** paths, including computed 5-call / 4-call-B |
| `cacs_isochrone()` / `cacs_run()` site input | `sf` points or tibble | also accepts base `data.frame` |
| `cacs_acs_prefetch()` year handling | broader, fractional years truncated | pinned to 2009-2024; fractional years rejected |
| ACS key requirement | key checked before cache lookup | cache checked **first**; key needed only for live misses / `force_refresh` |
| Provider / weighting status | some surfaces returned placeholders | OSRM + ORS implemented; others fail loud |
| `weight_method = "population"` | aborts during the run | aborts **immediately**, before any work |

## Change 1 -- list-column output is now fully populated

In v0.4, `cacs_run(output = "list_column")` (and `output = "both"`) filled the
`$isochrone` list-column only on the precomputed bypass paths. In v0.5 the same
list-column cell is filled from the resolved isochrone object on **all**
supported execution paths, including the computed 5-call path and the 4-call-B
path (`acs` supplied, isochrones computed). Each `$isochrone` cell holds the
one-row `sf` for its `(site_id, drive_time_min)` pair, so the result is
self-contained for mapping whether or not you passed `precomputed_isochrones`.

```{r}
#| eval: false

# v0.5: $isochrone is populated even when isochrones are computed live.
run_lc <- cacs_run(
  sites       = sites_sf,
  state       = "AL",
  drive_times = c(5, 10),
  provider    = "osrm",
  output      = "list_column"
)
print(run_lc$isochrone[[1]])   # an sf row, not NULL
```

This is strictly more data than v0.4: no v0.4 script relied on these cells being
`NULL`. The once-per-session fill notice you may have seen on bypass paths in
v0.4 now also fires on the computed paths.

## Change 2 -- base `data.frame` site inputs

`cacs_isochrone()` now accepts a base `data.frame` of sites in addition to
tibbles and `sf` POINT objects, and `cacs_run()` was aligned to the same input
contract. A non-`sf` input must still carry `site_id`, `lon`, and `lat`
columns; the function promotes it to `sf` (EPSG:4326) for you.

```{r}
#| eval: false

sites_df <- data.frame(
  site_id = c("S1", "S2"),
  lon     = c(-86.81, -87.55),
  lat     = c(33.52, 34.73)
)

# v0.5: a plain data.frame works directly, no manual sf::st_as_sf() needed.
iso <- cacs_isochrone(sites_df, drive_times = c(5, 10), provider = "osrm")
out <- cacs_run(sites_df, state = "AL", drive_times = c(5, 10))
```

Tibble and `sf` inputs behave exactly as before, so existing scripts need no
change. See `?cacs_isochrone` and `?cacs_run` for the full input contract.

## Change 3 -- ACS preflight is stricter and safer

`cacs_acs_prefetch()` tightened three behaviors:

- **End years pinned to 2009-2024.** This release verifies ACS5 support only
  over that window. A `year` outside it aborts with a clear range message
  instead of attempting an unverified Census endpoint.
- **Fractional years are rejected.** A non-integer `year` (e.g. `2023.5`) now
  aborts asking for a whole-number year, rather than silently truncating it.
- **The cache is checked before a Census key is required.** A cache hit (or an
  explicit `acs =` bypass through `cacs_run()`) is served with **no**
  `CENSUS_API_KEY`. Only a live cache *miss* or `force_refresh = TRUE` requires
  a key before any tidycensus metadata or data call.

```{r}
#| eval: false

# Cache hit or explicit `acs =` bypass: no Census key needed.
acs_cached <- cacs_acs_prefetch(state = "AL", year = 2023)   # served from cache

# Live miss or force_refresh = TRUE: a CENSUS_API_KEY is required first.
acs_fresh <- cacs_acs_prefetch(state = "AL", year = 2023, force_refresh = TRUE)

# Rejected in v0.5 (these abort rather than guess):
# cacs_acs_prefetch(state = "AL", year = 2025)     # outside 2009-2024
# cacs_acs_prefetch(state = "AL", year = 2023.5)   # fractional year
```

If a v0.4 script passed an out-of-window or fractional `year`, round it to a
supported whole-number year. See `?cacs_acs_prefetch` for the cache and
live-access sections.

## Change 4 -- provider and weighting status is explicit and fail-loud

v0.5 makes the implemented surface explicit across the documentation and the
code paths. OSRM and ORS are the implemented routing providers. Mapbox, r5r,
population weighting, and custom-rate catalogues are reserved or deferred for a
future release, and selecting them now **fails loud** with a future-release
message instead of returning a partial or placeholder result.

```{r}
#| eval: false

# Implemented today:
iso_osrm <- cacs_isochrone(sites_df, provider = "osrm")
iso_ors  <- cacs_isochrone(sites_df, provider = "ors")   # needs an ORS key

# Reserved -- these abort with a future-release message in v0.5:
# cacs_isochrone(sites_df, provider = "mapbox")
# cacs_isochrone(sites_df, provider = "r5r")
```

If your script only used OSRM or ORS (the documented routing providers), nothing
changes. See `vignette("providers")` for provider setup and the implemented
matrix.

## Change 5 -- `weight_method = "population"` fails fast

`weight_method = "population"` is still reserved for a future release. In v0.5 it
now aborts **immediately at the orchestrator boundary**, before any ACS,
provider, or spatial work begins, so you no longer wait on a live routing or
Census request for a path that cannot complete. Area weighting remains the only
implemented method.

```{r}
#| eval: false

# v0.5: aborts at once, before any network or spatial work.
# cacs_run(sites_df, state = "AL", weight_method = "population")

# Use the implemented method (the default):
out <- cacs_run(sites_df, state = "AL", weight_method = "area")
```

The reserved `bg_pop_sf` argument is likewise ignored under area weighting. If a
v0.4 script already used `weight_method = "area"` (the default), it is unaffected.

## What did not change

- The `cacs_run()` contract: the 5-call / 3-call / 4-call execution-path
  semantics and the 21-argument signature.
- The five sanctioned rates and their frozen formulas.
- The canonical long output schema and its column order.
- The area-weighting math and the per-tract area-weight definitions.
- All v0.4 classed conditions still emit; v0.5 only broadens wording and adds
  fail-loud guards where placeholders used to be.

## Further reading

- `?cacs_run` -- list-column fill, provider/weighting status, fast-fail boundary.
- `?cacs_isochrone` -- accepted site inputs and provider status.
- `?cacs_acs_prefetch` -- year window, cache behavior, and live-access sections.
- `vignette("getting-started")` -- the end-to-end workflow.
- `vignette("alabama-tutorial")` -- a full applied walkthrough.
- `vignette("providers")` -- routing-provider setup and the implemented matrix.
- `vignette("porting-v03-to-v04")` -- the previous migration guide.
