---
title: "Porting catchmentACS workflows to v0.3"
vignette: >
  %\VignetteIndexEntry{Porting catchmentACS workflows to v0.3}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 2
execute:
  warning: false
  message: false
---

This guide collects, in one place, the changes you are most likely to hit when
moving a v0.1 or v0.2 workflow to v0.3. If a validator message pointed you to
`vignette("porting-v01-to-v03", package = "catchmentACS")`, work through the
sections below in order.

Four changes account for most of the migration work: isochrone topology is now
fixed to the cumulative form, the OSRM `res` default is stated explicitly as
`70L`, cache identity is protected by a fingerprint, and the SSI rate formula is
corrected to `B19056_002 / B19056_001`.

```{r}
#| label: setup
#| eval: true

library(catchmentACS)
library(dplyr)
library(sf)

cacs_resource <- function(dir, file) {
  installed <- system.file(dir, file, package = "catchmentACS")
  if (nzchar(installed) && file.exists(installed)) {
    return(installed)
  }
  for (candidate in c(file.path("inst", dir, file),
                      file.path("..", "inst", dir, file))) {
    if (file.exists(candidate)) {
      return(normalizePath(candidate, winslash = "/", mustWork = TRUE))
    }
  }
  stop("Cannot find bundled catchmentACS resource: ", file.path(dir, file))
}

fx031 <- readRDS(cacs_resource("testdata", "fixture_031_3site_anchors.rds"))
fx032 <- readRDS(cacs_resource("testdata", "fixture_032_3site_fresh.rds"))
annulus_fixture <- readRDS(cacs_resource("testdata", "fixture_annulus_input.rds"))
legacy_iso <- readRDS(cacs_resource("extdata", "legacy_2025_isochrones.rds"))
```

::: {.callout-note}
## About the data in this article

Every table and check here is built entirely from the package's bundled
`inst/testdata` and `inst/extdata` fixtures (the synthetic `fixture_*` and
`legacy_2025_*` objects), so this article renders with **no** Census, OSRM, ORS,
or other network calls. These fixtures preserve the production calling contract
(sites, drive-time polygons, tract-level ACS variables) but are **synthetic and
illustrative** — included only for examples, tests, and reproducible package
checks. They are **not** live Alabama policy estimates and must not be cited as
real ACS results. Any real analysis requires a fresh live run (see
`vignette("providers", package = "catchmentACS")`).
:::

## Quick map

| Area | v0.1 / v0.2 pattern | v0.3 contract |
|---|---|---|
| Isochrone topology | annulus and cumulative rows could be mixed | pipeline ingestion accepts cumulative rings only |
| Annulus conversion | implicit interpretation was easy to miss | convert explicitly with `cacs_rings_to_cumulative()` |
| OSRM `res` | omitting `res` followed older behavior | omitting `res` resolves to `70L` with a once-per-session notice |
| Cache identity | legacy entries had weaker compatibility checks | persisted entries use sha256 fingerprint sidecars |
| SSI rate | the old denominator used the wrong household universe | `ssi_rate = B19056_002 / B19056_001` |
| Diagnostics | run-time errors were terse | validators include issue tables, fix hints, and examples |
| Summary output | wide tibble print could wrap awkwardly | use `cacs_summary_as_markdown()` for Quarto/GitHub tables |

## Eight script-side transitions from the 2025 replication

### S-01: Use a deliberate planar regime for archived polygons

Older archived polygons can be repaired and compared after moving to a metric
CRS. The package handles this internally; in custom replication scripts, make
the same choice explicit.

```{r}
#| label: s01-planar-regime
#| eval: true

old_s2 <- sf::sf_use_s2(FALSE)
iso_5070 <- fx031$iso_sf_v2 |>
  sf::st_make_valid() |>
  sf::st_transform(5070)
self_intersects <- diag(sf::st_intersects(iso_5070, iso_5070, sparse = FALSE))
sf::sf_use_s2(old_s2)

tibble::tibble(all_self_intersect = all(self_intersects))
```

### S-02: Filter cumulative rings by both lower and upper bounds

The 2025 script initially matched `isomax` alone, which can mix annulus and
cumulative rows. v0.3 makes topology explicit.

```{r}
#| label: s02-ring-filter
#| eval: true

mixed <- rbind(
  annulus_fixture[annulus_fixture$isomax == 10, ],
  transform(
    annulus_fixture[annulus_fixture$isomax == 10, ],
    isomin = 0L,
    ring_topology = "cumulative"
  )
)

tibble::tibble(
  rows_by_isomax_only = nrow(dplyr::filter(sf::st_drop_geometry(mixed),
                                           isomax == 10L)),
  rows_by_cumulative_filter = nrow(dplyr::filter(sf::st_drop_geometry(mixed),
                                                 isomin == 0L,
                                                 isomax == 10L,
                                                 ring_topology == "cumulative"))
)
```

### S-03: Use the canonical coordinate source

When you reconcile administrative and geocoded coordinates, keep the decision
as data, not a hand-typed value, so each site records which source it used and
why. The illustrative table below flags one anchor where the two sources
differed by more than 600 m (synthetic example; real administrative
coordinates are not shipped with the package).

```{r}
#| label: s03-coordinate-audit
#| eval: true

tibble::tibble(
  site_id      = c("AL_ANCHOR_01", "AL_ANCHOR_02", "AL_ANCHOR_03"),
  used_in_2025 = c("administrative", "administrative", "geocoded"),
  geo_diff_m   = c(0, 621, 0),
  verdict      = c("match", "mismatch (>600 m; used administrative)", "match")
)
```

### S-04: Prefer scalar-safe extraction

`mapply()` can return a list when the extraction function is not scalar-stable.
Use `vapply()` or dplyr joins for deterministic numeric vectors.

```{r}
#| label: s04-scalar-extraction
#| eval: true

acs_032 <- fx032$acs_sf |> sf::st_drop_geometry()
queries <- acs_032 |>
  dplyr::filter(variable == "B01003_001") |>
  dplyr::slice_head(n = 3)

extract_one <- function(data, geoid, variable) {
  data$estimate[data$GEOID == geoid & data$variable == variable][[1]]
}

vapply(seq_len(nrow(queries)), function(i) {
  extract_one(acs_032, queries$GEOID[[i]], queries$variable[[i]])
}, numeric(1))
```

### S-05: Map friendly 2025 wide names back to raw ACS codes

Legacy 2025 exports sometimes used friendly stems such as `total_pop_E`. The
v0.3 package surface expects raw ACS variable codes in long form.

```{r}
#| label: s05-friendly-names
#| eval: true

wide_2025 <- queries |>
  dplyr::transmute(GEOID, total_pop_E = estimate)
code_to_stem <- c(B01003_001 = "total_pop")

tibble::tibble(
  code = "B01003_001",
  legacy_column = paste0(unname(code_to_stem[["B01003_001"]]), "_E"),
  first_value = wide_2025[[paste0(unname(code_to_stem[["B01003_001"]]), "_E")]][[1]]
)
```

### S-06: Preflight CRS before intersection

`cacs_intersect_weight()` accepts isochrones in EPSG:4326 and ACS tracts in
EPSG:4269, then reprojects internally. Use `cacs_validate_iso()` to catch CRS
mistakes before the heavy path.

```{r}
#| label: s06-crs-preflight
#| eval: true

bad_crs <- sf::st_transform(legacy_iso[1:2, ], 4269)
cacs_validate_iso(bad_crs) |>
  dplyr::filter(check == "iso_crs") |>
  dplyr::select(check, actual, expected, example)
```

### S-07: Keep provider metadata typed

`provider_downgrade` is logical, not character. v0.3 diagnostics now point to
the exact column and an example.

```{r}
#| label: s07-provider-downgrade-type
#| eval: true

bad_type <- legacy_iso[1, ]
bad_type$provider_downgrade <- NA_character_
cacs_validate_iso(bad_type) |>
  dplyr::filter(col == "provider_downgrade") |>
  dplyr::select(col, actual, expected, example)
```

### S-08: Treat weighted seams as tibbles with audit attributes

The weighted seam is a canonical long tibble, not an `sf` object and not a
tract-level geometry table. Tract and carrier details live in explicit columns
or attributes.

```{r}
#| label: s08-weighted-seam-location
#| eval: true

weighted <- fx031$weighted_seam_v2
tibble::tibble(
  is_sf = inherits(weighted, "sf"),
  has_GEOID = "GEOID" %in% names(weighted),
  has_weight_sum = "weight_sum" %in% names(weighted),
  carrier_rows = nrow(attr(weighted, "cacs_aggregation_carriers"))
)
```

## v0.2 to v0.3 package changes

| Group | User-facing change |
|---|---|
| Silent data fixes | ACS test cache isolation, fingerprint sidecars, corrected SSI formula, and carrier-aware rate MOE |
| Architecture | cumulative-only rings, OSRM omitted-`res = 70L`, cache kept on by default but made verifiable |
| UX | classed conditions, progress messages, visual helper input forms, schema preflight |
| Performance | fast OSRM path, warm-cache counters, and regression guards |
| Design | tract audit attribute, urbanicity divergence docs, internal sf noise suppression, Markdown summary helper |

## Cumulative-rings migration

`ring_topology = "annulus"` is not accepted by v0.3 ingestion. Convert complete
annulus chains explicitly, and reject incomplete chains during preflight.

```{r}
#| label: cumulative-ring-positive-example
#| eval: true

make_square <- function(xmin, ymin, xmax, ymax) {
  sf::st_polygon(list(rbind(
    c(xmin, ymin), c(xmax, ymin), c(xmax, ymax),
    c(xmin, ymax), c(xmin, ymin)
  )))
}

toy_annulus <- sf::st_sf(
  tibble::tibble(
    site_id = "S01",
    drive_time_min = c(5L, 10L, 15L),
    isomin = c(0L, 5L, 10L),
    isomax = c(5L, 10L, 15L),
    ring_topology = "annulus",
    provider = "osrm",
    profile = "car",
    generated_at = Sys.time()
  ),
  geometry = sf::st_sfc(
    make_square(-87.10, 33.00, -87.00, 33.10),
    make_square(-87.15, 32.95, -86.95, 33.15),
    make_square(-87.20, 32.90, -86.90, 33.20),
    crs = 4326
  )
)

cacs_rings_to_cumulative(toy_annulus) |>
  sf::st_drop_geometry() |>
  dplyr::select(site_id, isomin, isomax, drive_time_min, ring_topology)
```

## Cache migration

The first v0.3 run may recompute old cache entries. That is expected: entries
without a valid fingerprint sidecar, corrupt entries, malformed sidecars, or
digest mismatches are treated as misses.

```{r}
#| label: cache-state
#| eval: true

state <- cacs_get_cache_state()
state |>
  dplyr::select(enabled, namespace_mode, fingerprint_algorithm)
```

For reproducible scripts, temporarily disable cache reads or set
`options(catchmentACS.cache_dir = "...")` before the package cache path is
first resolved:

```{r}
#| label: cache-controls
#| eval: true

old_cache_enabled <- cacs_get_cache_state()$enabled[[1]]
cacs_set_cache(FALSE, scope = "session")
cacs_get_cache_state() |>
  dplyr::select(enabled, cache_dir) |>
  dplyr::mutate(cache_dir = "<user-cache-dir>/catchmentACS")
cacs_set_cache(old_cache_enabled, scope = "session")
```

## API surface deltas

| Helper | Purpose |
|---|---|
| `cacs_validate_iso()` | non-aborting isochrone preflight issue table |
| `cacs_rings_to_cumulative()` | explicit annulus-to-cumulative conversion |
| `cacs_set_cache()` / `cacs_get_cache_state()` | cache controls and hit/miss diagnostics |
| `cacs_capture_conditions()` | classed condition capture for audit logs |
| `cacs_describe()` | provenance-oriented object description |
| `cacs_summary_as_markdown()` | Markdown rendering for summary tables |

```{r}
#| label: api-summary-markdown
#| eval: true

summary_tbl <- tibble::tibble(
  variable = c("poverty_rate", "ssi_rate"),
  mean = c(0.123456, 0.045678),
  sd = c(0.0101, 0.0042),
  n_NA = c(0L, 0L)
)
cacs_summary_as_markdown(summary_tbl)
```

## What did not change

`cacs_run()` remains the main entry point. Cache remains enabled by default,
because v0.3 makes cache entries safer rather than removing the speed benefit.
Explicit `res = 50L` remains supported when reproducing v0.2-style OSRM
geometry. In v0.3, omitting `res` resolves to `70L`; in v0.5.0, public-demo OSRM
calls default to `30L` unless demo-budget protection is disabled, while
docker and self-hosted paths keep the `70L` default.

## Further reading

- `vignette("getting-started", package = "catchmentACS")`
- `vignette("alabama-tutorial", package = "catchmentACS")`
- `vignette("providers", package = "catchmentACS")`
- `vignette("porting-v03-to-v04", package = "catchmentACS")` — v0.3 → v0.4/v0.5
- `?cacs_run`
- `?cacs_validate_iso`
- `?cacs_rings_to_cumulative`
- `?cacs_describe`
- `?cacs_summary_as_markdown`
