---
title: "Providers, Credentials, and Offline Paths"
vignette: >
  %\VignetteIndexEntry{Providers, Credentials, and Offline Paths}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 2
execute:
  warning: false
  message: false
---

This article is the v0.5.0 status page for live services and no-network paths in `catchmentACS`. It answers three practical questions:

1. Which services are implemented now?
2. Which credentials are required?
3. How can examples, tests, and reports run offline without accidentally calling Census or a routing provider?

```{r}
#| label: setup
#| eval: true

library(catchmentACS)
library(dplyr)
library(sf)

old_options <- options(
  catchmentACS.progress = "off",
  catchmentACS.rate_first_default = FALSE
)
old_cache <- getOption("catchmentACS.cache_enabled", NULL)
on.exit(options(old_options), add = TRUE)
on.exit(options(catchmentACS.cache_enabled = old_cache), add = TRUE)
cacs_set_cache(FALSE, scope = "session")
```

::: {.callout-note}
## About the data in this article

Every table and check here is built entirely from the package's bundled
`inst/extdata` fixtures (the synthetic `legacy_2025_*` objects and
`sample_alabama_subset`), so this article renders with **no** Census, OSRM, ORS,
or other network calls. These fixtures preserve the production calling contract
(sites, drive-time polygons, tract-level ACS variables) but are **synthetic and
illustrative** — included only for examples, tests, and reproducible package
checks. They are **not** live Alabama policy estimates and must not be cited as
real ACS results. The live code blocks below are shown for reference and marked
non-evaluated; the one provider probe that does run points at an unreachable
localhost port so it stays offline and deterministic. Any real analysis requires
a fresh live run.
:::

## Status matrix

```{r}
#| label: status-matrix
#| eval: true

tibble::tribble(
  ~surface, ~status, ~credential, ~notes,
  "Census ACS5 tracts", "implemented", "Census API key required for live metadata/data",
  "Release-verified for 2009-2024. Cache hits and `acs =` bypass do not need a key.",
  "OSRM", "implemented", "none for public demo",
  "Default provider. Public demo is convenient but can be slow or rate-limited.",
  "OSRM docker/self-hosted", "implemented", "local server URL",
  "Use `osrm_mode = \"docker\"` or provider args for a local/custom server.",
  "OpenRouteService (ORS)", "implemented", "`ors_api_key` or `ORS_API_KEY`",
  "Requires optional openrouteservice package and a non-empty API key.",
  "Mapbox", "reserved fail-loud", "`MAPBOX_TOKEN` not used yet",
  "Provider name is retained, but selecting it aborts in this release.",
  "r5r", "reserved fail-loud", "r5r core not used yet",
  "Provider name is retained, but selecting it aborts in this release.",
  "Area weighting", "implemented", "none",
  "The supported v0.5.0 aggregation method.",
  "Population weighting", "reserved fail-loud", "future population surface",
  "`weight_method = \"population\"` aborts before network or spatial work."
)
```

The short version: use `provider = "osrm"` for the easiest live route, use `provider = "ors"` when you have an ORS key, and use `precomputed_isochrones` plus `acs` whenever you need a completely offline run.

## Which provider should I use?

OSRM and ORS are the two live routing paths in v0.5.0; Mapbox and r5r are reserved names that abort (see "Fail-loud reserved providers"). The decision is mostly about volume and whether you can stand up a local server:

| You want to... | Use | Setup / key |
|---|---|---|
| Try the package or run a handful of sites | `provider = "osrm"`, `osrm_mode = "demo"` | No key, no setup; shared public demo, can be slow or HTTP 429 rate-limited |
| Run a production batch (many sites, high `res`) | `provider = "osrm"`, `osrm_mode = "docker"` | A local/self-hosted OSRM server (default `http://0.0.0.0:5000/`); no per-request quota |
| Avoid hosting OSRM but still route remotely | `provider = "ors"` | Optional `openrouteservice` package **and** a non-empty `ors_api_key` / `ORS_API_KEY` |
| Render docs, tests, or reproducible reports | none — supply `precomputed_isochrones =` and `acs =` | Fully offline; no key, no network (this article's path) |

Rules of thumb: the OSRM public demo is for convenience and low volume, not large jobs; a local OSRM docker endpoint is the production default once batches grow; ORS is the remote alternative when you have a key but cannot self-host. Mapbox and r5r are roadmap names only. The rest of this article covers credentials and each path in detail.

## Credentials

### Census API

ACS downloads are handled by `tidycensus`. Register a Census key once in an interactive R session:

```{r}
#| label: census-key
#| eval: false

tidycensus::census_api_key("YOUR_KEY_HERE", install = TRUE)
```

After restart, `cacs_run()` can call `cacs_acs_prefetch()` automatically in the full live path. A key is checked only after local validation and an ACS cache miss, before live `tidycensus::load_variables()` or `tidycensus::get_acs()` calls. If you pass `acs = <sf>` or hit a warm ACS cache entry, the live ACS stage is skipped and no Census key is required. v0.5.0 supports ACS5 end years 2009-2024 (2024 is the most recent vintage). See `?cacs_acs_prefetch` for the prefetch contract and `?cacs_acs_validate` for checking your own ACS table.

### OSRM

OSRM is the default routing provider and needs no API key when using the public demo endpoint:

```{r}
#| label: osrm-live-demo
#| eval: false

iso <- cacs_isochrone(
  sites = cacs_alabama_sites[1:3, ],
  drive_times = c(5, 10, 15),
  provider = "osrm",
  osrm_mode = "demo",
  res = 30L,
  verbose = TRUE
)
```

For heavier batch work, prefer a local or self-hosted OSRM endpoint. The public demo endpoint is shared infrastructure and can return HTTP 429 under load.

```{r}
#| label: osrm-docker
#| eval: false

iso <- cacs_isochrone(
  sites = cacs_alabama_sites[1:3, ],
  drive_times = c(5, 10, 15),
  provider = "osrm",
  osrm_mode = "docker",
  `osrm.server` = "http://0.0.0.0:5000/",
  verbose = TRUE
)
```

When `res` is omitted on the public demo path, v0.5.0 protects the demo budget by resolving to a lower default than the self-hosted path. Pass `res` explicitly when reproducibility of polygon detail matters.

Before a long OSRM run, you can probe the endpoint with `cacs_validate_osrm_endpoint()`. It is a probe, not a gate: it never throws, returning a one-row tibble whose `quota_ok` flag and `http_status` you inspect (a known-saturated demo also emits a classed `catchmentACS_warning_provider_quota_exhausted` warning). The rendered vignette uses a known-unreachable localhost port, so the chunk stays offline and deterministic; the connection simply fails and `quota_ok` comes back `FALSE`.

```{r}
#| label: osrm-endpoint-preflight-offline
#| eval: true

cacs_validate_osrm_endpoint(
  server = "http://localhost:9",
  timeout = 1
)
```

For a real live preflight, omit `server` or point it at your self-hosted endpoint:

```{r}
#| label: osrm-endpoint-preflight-live
#| eval: false

cacs_validate_osrm_endpoint()
cacs_validate_osrm_endpoint(server = "http://localhost:5000", timeout = 2)
```

### ORS

ORS is implemented, but it requires the optional `openrouteservice` package and a non-empty API key supplied through `ors_api_key` or `ORS_API_KEY`.

```{r}
#| label: ors-live
#| eval: false

Sys.getenv("ORS_API_KEY")

iso <- cacs_isochrone(
  sites = cacs_alabama_sites[1:3, ],
  drive_times = c(5, 10, 15),
  provider = "ors",
  ors_api_key = Sys.getenv("ORS_API_KEY"),
  verbose = TRUE
)
```

Do not put API keys directly in scripts or vignettes. Use `.Renviron`, an environment manager, or CI secrets.

## Offline bypass paths

`cacs_run()` selects its execution path from the presence or absence of two optional inputs:

```{r}
#| label: execution-path-table
#| eval: true

tibble::tribble(
  ~execution_path, ~precomputed_isochrones, ~acs, ~network_behavior,
  "5-call", "not supplied", "not supplied",
  "Calls both routing provider and tidycensus.",
  "4-call-A", "supplied", "not supplied",
  "Skips routing provider; still calls tidycensus.",
  "4-call-B", "not supplied", "supplied",
  "Calls routing provider; skips tidycensus.",
  "3-call", "supplied", "supplied",
  "Skips both routing provider and tidycensus."
)
```

The package vignettes use the `3-call` path. The `provider` column still records the provenance of the precomputed isochrone, but no live provider is called while the article renders. The bundled legacy isochrones are synthetic buffered-circle polygons; `provider = "osrm"` records OSRM-style provenance metadata and does not imply a live OSRM request during rendering.

```{r}
#| label: fixture-provenance
#| eval: true

legacy_iso <- readRDS(system.file(
  "extdata", "legacy_2025_isochrones.rds", package = "catchmentACS"
))

legacy_iso |>
  sf::st_drop_geometry() |>
  count(provider, provider_requested, profile,
        osm_snapshot_date, ring_topology)
```

Here is the smallest complete offline `cacs_run()` provider check:

```{r}
#| label: offline-provider-check
#| eval: true

legacy_sites <- readRDS(system.file(
  "extdata", "legacy_2025_sites.rds", package = "catchmentACS"
))
acs <- readRDS(system.file(
  "extdata", "sample_alabama_subset.rds", package = "catchmentACS"
))

site_id <- "AL_SITE_03"
iso_one <- legacy_iso |>
  filter(site_id == !!site_id, drive_time_min == 5L)
site_one <- legacy_sites |>
  filter(site_id == !!site_id)

out <- cacs_run(
  sites = site_one,
  state = "AL",
  year = 2023,
  drive_times = 5L,
  variables = unname(cacs_acs_default_vars),
  provider = "osrm",
  precomputed_isochrones = iso_one,
  acs = acs,
  weight_method = "area",
  output = "long",
  verbose = FALSE
)

attr(out, "cacs_run_provenance") |>
  as.data.frame() |>
  select(execution_path, provider, bypass_iso, bypass_acs,
         weight_method, output_format)
```

Use the same pattern in tests and reproducible reports: ship or load the exact `sf` ACS and isochrone inputs, pass them into `cacs_run()`, and record the resulting provenance attributes. For the same offline call worked into a full reporting table, see `vignette("alabama-tutorial", package = "catchmentACS")`.

## Fail-loud reserved providers

Mapbox and r5r are in the provider enum for compatibility and roadmap visibility, but they are intentionally not live routing paths in v0.5.0. Selecting either one raises a structured `catchmentACS_error_credential` condition rather than silently returning a placeholder. The check below catches that condition and reports only its class, so the chunk stays offline and prints no raw provider message.

```{r}
#| label: fail-loud-check
#| eval: true

demo_site <- tibble::tibble(
  site_id = "S01",
  lon = -86.80902,
  lat = 33.52203
)

condition_class <- function(expr) {
  tryCatch(
    {
      force(expr)
      "no_error"
    },
    catchmentACS_error = function(e) class(e)[[1L]],
    error = function(e) class(e)[[1L]]
  )
}

fake_r5r_core <- structure(list(), class = c("r5r_core", "list"))

tibble::tibble(
  provider = c("mapbox", "r5r"),
  condition = c(
    condition_class(cacs_isochrone(
      sites = demo_site,
      provider = "mapbox",
      drive_times = 5L,
      mapbox_token = "FAKE_TOKEN_FOR_DOCS",
      verbose = FALSE
    )),
    condition_class(cacs_isochrone(
      sites = demo_site,
      provider = "r5r",
      drive_times = 5L,
      r5r_core = fake_r5r_core,
      verbose = FALSE
    ))
  )
)
```

The same release-status principle applies to `weight_method = "population"`: the argument is reserved, but the implementation is not active in this release. It fails loud at the orchestrator boundary, before any network or spatial work begins.

```{r}
#| label: population-fail-loud
#| eval: true

condition_class(cacs_run(
  sites = demo_site,
  state = "AL",
  drive_times = 5L,
  provider = "osrm",
  weight_method = "population",
  verbose = FALSE
))
```

## Cache behavior

The persistent cache is useful for live work, especially provider and ACS calls. Vignettes disable it for deterministic no-network rendering.

```{r}
#| label: cache-state
#| eval: true

cacs_get_cache_state() |>
  select(enabled, cache_dir, namespace_mode, counters) |>
  mutate(cache_dir = "<user-cache-dir>/catchmentACS")
```

For interactive live analysis, the default cache can save repeated requests. A warm ACS cache entry returns without calling Census and without requiring `CENSUS_API_KEY`; `force_refresh = TRUE` deliberately bypasses that cache and therefore requires a key. If a live ACS or isochrone input changes and you want a fresh pull, clear the relevant namespace or pass provider/fetch arguments that change the cache key.

```{r}
#| label: cache-maintenance
#| eval: false

cacs_cache_status()
cacs_clear_cache("isochrone", confirm = FALSE)
cacs_clear_cache("acs", confirm = FALSE)
```

## Live full-run template

The full live path is concise. It is non-evaluated here because it calls both a routing provider and Census.

```{r}
#| label: live-run-template
#| eval: false

tidycensus::census_api_key("YOUR_KEY_HERE", install = TRUE)

result <- cacs_run(
  sites = cacs_alabama_sites,
  state = "AL",
  year = 2023,
  drive_times = c(5, 10, 15),
  variables = unname(cacs_acs_default_vars),
  provider = "osrm",
  weight_method = "area",
  output = "long",
  verbose = TRUE
)

summary(result)$rates_per_site_moe
attr(result, "cacs_run_provenance")
```

## Where to next

- **Your first run in ten minutes** — the whole pipeline in one offline call:
  `vignette("getting-started", package = "catchmentACS")`.
- **A full reporting workflow** on Alabama sites, taken from the raw pipeline to
  export-ready tables:
  `vignette("alabama-tutorial", package = "catchmentACS")`.
- **The math** behind area weighting and MOE propagation:
  `vignette("methodology", package = "catchmentACS")`.
- **Function reference** — the orchestrator, the isochrone builder, the ACS
  prefetch, and the OSRM preflight probe: `?cacs_run`, `?cacs_isochrone`,
  `?cacs_acs_prefetch`, and `?cacs_validate_osrm_endpoint`.

Bug reports and feature requests:
<https://github.com/joonho112/catchmentACS/issues>.
