---
title: "Area weighting of tract estimates"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    math_method: mathml
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{Area weighting of tract estimates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-options
#| include: false

knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  out.width = "85%"
)
```

```{r}
#| label: setup
#| eval: true

library(catchmentACS)
library(dplyr)
library(sf)  # needed to subset the bundled sf objects with [
```

```{r}
#| label: setup-cache
#| include: false

# Compute every result in this article instead of reading saved ones; the
# option is restored at the end of the article.
old_options <- options(catchmentACS.cache_enabled = FALSE)
```

A drive-time area is drawn from travel times, and its edge cuts across census
tracts: the area around a site takes in some tracts whole and parts of others.
catchmentACS estimates characteristics of the people and households in the
area, such as the number below the poverty level, from the American Community
Survey (ACS) estimates for these tracts. The weight of each tract is computed
from the area it shares with the drive-time area. The margins of error of these
estimates, half-widths of confidence intervals at the 90 percent level by
default, are derived in
`vignette("theory-moe-propagation", package = "catchmentACS")`, and the rates
are described in `vignette("theory-derived-rates", package = "catchmentACS")`.

## Area weighting and its assumption

As in `vignette("methodology", package = "catchmentACS")`, we write $I_s$ for
the drive-time area of site $s$ at one drive time and $T_j$ for census tract
$j$. The part of the tract inside the drive-time area is $I_s \cap T_j$, and
$\lvert\,\cdot\,\rvert$ denotes area. An ACS estimate describes a whole tract,
while the drive-time area may contain only part of it.

Area weighting is a simple and widely used way to move counts from one set of
areas to another. It assumes that whatever a variable counts is spread evenly
over the tract's area, which rarely holds in practice [@comber2019spatial,
p. 8]. Let $\rho_j$ be the number per unit area, in tract $T_j$, of whatever
the variable counts: people, households, or a subgroup such as the people below
the poverty level. If $\rho_j$ is the same everywhere in the tract, the number
in any part $S$ of the tract is proportional to the area of that part:

$$
\text{number in } S = \rho_j \, \lvert S \rvert
= \big(\text{number in } T_j\big) \cdot \frac{\lvert S \rvert}{\lvert T_j \rvert}.
$$

With $S = I_s \cap T_j$, the part of a tract count that lies in the drive-time
area is the count multiplied by the share of the tract's area inside the
drive-time area. This share is the coverage weight of the next section.

Counts, such as the number of people below the poverty level or the number of
households, can be split this way. Medians and per-person values, such
as median household income (`B19013_001`) and per capita income
(`B19301_001`), do not add up over area: half of a tract does not have half of
its per capita income. For these, the package averages the tract values with
weights proportional to the area each tract shares with the drive-time area.
This average stands in for the median or per-person value of the drive-time
area, but its weights follow area and ignore how many people live in each
tract. It can therefore be far from that value when the overlapping tracts
differ in population density, even if people are spread evenly within each
tract; the next section gives an example. The margin of error reported with
such an average describes the sampling error of the average, not the
difference between it and the drive-time area's own value, and that difference
does not shrink as the ACS margins of error do. A rate, such as the poverty rate, is
the ratio of two counts, and each of the two counts is split by area.

The assumption fails where the people of a tract live on a small part of its
area. Suppose that 80 percent of a tract is uninhabited woodland and that its
residents live on the remaining 20 percent. A drive-time area that covers all
of the woodland and none of the settled part gives the tract a coverage weight
of 0.8, so it is credited with 80 percent of the tract's residents, although
none of them live inside. The error grows with how unevenly people are spread
within the tracts, as in a large rural tract that contains a small town.
Methods that use other data, such as land cover, to place people within tracts
relax the assumption [@comber2019spatial, pp. 3–5]. catchmentACS uses area
weighting only: `weight_method = "population"` is not implemented yet, and
using it gives an error.

## Coverage weights and area shares

Both weights are computed from the overlap area $\lvert I_s \cap T_j \rvert$
and differ in what it is divided by. The coverage weight of tract $j$ divides
it by the area of the tract,

$$
w^{cov}_{sj} = \frac{\lvert I_s \cap T_j \rvert}{\lvert T_j \rvert}
\;\in\; [0, 1],
$$

which is 1 when the tract lies wholly inside the drive-time area and 0 when the
two do not overlap. The coverage weights of the overlapping tracts need not sum
to one, since each is a share of a different tract. Their sum adds up the
covered fractions of the tracts: two whole tracts and half of a third give 2.5,
and so do five tracts that are each half covered. The column `weight_sum` of
the result reports the sum for each count, median, and per-person value.

The area share of tract $j$ divides the overlap area by the total overlap area
of the tracts,

$$
w^{mean}_{sj} = \frac{\lvert I_s \cap T_j \rvert}{\sum_k \lvert I_s \cap T_k \rvert},
$$

so the area shares sum to one. These sums, like all sums over tracts below,
run over the tracts kept for the site, drive time, and variable. A tract is
kept if its coverage weight is above the argument `min_weight` (see the section
on the calculation) and the ACS data have a row for it and the variable. The
area shares are proportional to the coverage weights multiplied by the tract
areas, $w^{mean}_{sj} \propto w^{cov}_{sj}\, \lvert T_j \rvert$. When the
tracts have the same area, the area shares are therefore the coverage weights
divided by their sum.

A count is estimated by the sum of the tract estimates $Y_j$ weighted by
coverage weights,

$$
\widehat{Y}_s = \sum_j w^{cov}_{sj}\, Y_j ,
$$

which grows with the part of each tract inside the drive-time area. With area
shares in place of coverage weights, the result would be an average of the
tract counts, on the scale of one tract.

A median or a per-person value is estimated by the average of the tract
estimates $X_j$ weighted by area shares,

$$
\widehat{\bar X}_s = \sum_j w^{mean}_{sj}\, X_j .
$$

Because the area shares are nonnegative and sum to one, $\widehat{\bar X}_s$ is
at least the smallest and at most the largest of the tract values. A sum of the
same values weighted by coverage weights, whose sum can be above or below one,
can fall outside that range.

The area shares weight each tract by its overlap area and ignore how many
people live there. Two made-up tracts show how much this can matter. Tract 1
has an area of 1 km², 1,000 residents, and a per capita income of 10,000
dollars, and lies wholly inside the drive-time area. Tract 2 has an area of
100 km², 1,000 residents, and a per capita income of 50,000 dollars, and 10 km²
of it lie inside the drive-time area. The code below applies the package's
formula for a per-person value to these numbers by hand. For comparison, it
also computes the per capita income of the people inside, on the assumption
that within each tract both the residents and their income per person are the
same everywhere.

```{r}
#| label: two-tracts
#| eval: true

tracts <- data.frame(
  tract      = c(1, 2),
  area_km2   = c(1, 100),       # area of the tract
  inside_km2 = c(1, 10),        # area of its part inside the drive-time area
  residents  = c(1000, 1000),
  income_pc  = c(10000, 50000)  # per capita income, dollars
)
tracts$w_cov  <- tracts$inside_km2 / tracts$area_km2         # coverage weight
tracts$w_mean <- tracts$inside_km2 / sum(tracts$inside_km2)  # area share
tracts

# The package's formula: the average of the tract values weighted by area shares
area_share_average <- sum(tracts$w_mean * tracts$income_pc)

# The people inside and their per capita income (both assumed even within each tract)
residents_inside <- sum(tracts$w_cov * tracts$residents)
income_inside    <- sum(tracts$w_cov * tracts$residents * tracts$income_pc) /
  residents_inside

c(area_share_average = area_share_average,
  residents_inside   = residents_inside,
  income_inside      = income_inside)
```

The package's formula gives
`r format(round(area_share_average), big.mark = ",")` dollars, while the
`r format(residents_inside, big.mark = ",")` people inside have a per capita
income of `r format(round(income_inside), big.mark = ",")` dollars. Most of the
overlap area lies in tract 2, but most of the people inside live in tract 1.
The per capita income of the people inside is their total income divided by
their number, a ratio of two counts. The area-share average equals it when the
overlapping tracts have the same number of residents per unit area.

A rate, such as the poverty rate, is estimated by the ratio of two
coverage-weighted sums of tract estimates,
$\widehat{A}_s = \sum_j w^{cov}_{sj}\, A_j$ for its numerator and
$\widehat{B}_s = \sum_j w^{cov}_{sj}\, B_j$ for its denominator.
`vignette("theory-derived-rates", package = "catchmentACS")` shows when this
ratio is an average of the tract rates, weighted by the part of each tract's
denominator inside the drive-time area, and describes the margins of error of
rates.

## How the package chooses the weights

Each row of the results records the kind of estimate in the column
`estimand_family` and the weights used in `weight_basis`: coverage weights for
counts and rates, and area shares for medians and per-person values. The kind
is read from the ACS code alone, as the help page of `cacs_intersect_weight()`
and `vignette("methodology", package = "catchmentACS")` describe. Median age
(`B01002_001`), for instance, is a median from a table that the package does
not list, so it is added up like a count, without a warning.

For a median, the area-share average is a looser stand-in than for a
per-person value. Even when the tracts have the same population density, an
average of tract medians is in general not the median of the combined
population, which depends on the income distribution in each tract. The ACS
publishes its 5-year [detailed tables](https://www.census.gov/programs-surveys/acs/data/data-tables.html)
for all areas down to block groups,
census tracts included, and one of these tables, `B19001`, gives the number of
households in 16 income brackets. The package does not use these brackets: the
area-share average of the tract medians stands in for the median of the
drive-time area. The brackets themselves can be aggregated, because they are
counts: asking for `B19001_017` in `variables` gives the number of households
in that bracket for the drive-time area, with a coverage-weighted sum like any
other count. What the package has no function for is turning brackets into a
median.

## Measuring the areas

`cacs_intersect_weight()` computes the overlaps and their areas on a flat map.
Before measuring any area, it projects the drive-time areas and the tracts to
EPSG:5070 (NAD83 / Conus Albers) and measures all areas there, in square
meters. This Albers projection is equal-area: the area of a shape on the
projected map is its area on the reference ellipsoid. Every area in
a run is measured on this one map.

A coverage weight is the ratio of two areas within the same tract, so it
changes little with the way area is measured. EPSG defines this projection for
the contiguous 48 states. The package checks its own box for that scope, drawn
around the contiguous United States and the District of Columbia with a quarter
of a degree to spare. The bounding boxes of both inputs must lie between 24.14
and 49.64 degrees north and between -125.25 and -66.68 degrees east. Data for
Alaska, Hawaii, or Puerto Rico give an error.

## The calculation for each site

Before it combines anything, `cacs_intersect_weight()` repairs invalid
geometries in both inputs (see below) and skips, with a warning, tracts whose
area is zero or not finite. The skipped tracts are listed in the
`skipped_geoids` attribute of the result. Tracts with a coverage weight at or
below `min_weight` (by default `1e-6`), such as tracts that only touch the edge
of the area, are dropped before the area shares and the estimates are
computed.

The help page of `cacs_intersect_weight()` lists the steps of the calculation.
It also describes the single row of `NA` values, with
`failure_origin = "intersection"`, that a site and drive time gets when no
tract is left or the calculation fails.

Invalid geometries, such as polygons whose edges cross, are repaired with
`sf::st_make_valid()` or, if some remain invalid, with a buffer of zero width
(`sf::st_buffer()`), each time with a warning. This is done for both inputs
and again for the overlaps of each pair. Repair can change a shape and its
area. A geometry that is still invalid after both attempts is replaced by an
empty one. An empty tract has zero area and is skipped, an empty drive-time
area gives its pair the row described above, and an empty overlap adds
nothing. If every geometry of an input is still invalid, the function stops
with an error.

## Recomputing the weights for one site

The weights and one count for one site can be recomputed from the two areas
that `cacs_intersect_weight()` records for each tract when
`keep_tract_audit = TRUE`. The data for this example ship with the package. In
these data, squares of nearly equal size, scattered with gaps between them and
carrying random ACS values, take the place of census tracts. Each site sits at
the center of one of the squares, and its drive-time areas are circles around
it. The ACS values are made up, so the numbers below show the arithmetic and
not a real place. At a drive time of 10 minutes, the circle of site
`AL_SITE_17` contains the site's own square and cuts two other squares at its
edge. Because of the gaps, the squares cover only a small part of the circle;
real census tracts tile a state, so for an area of this size the sum of the
coverage weights there would be far larger.

```{r}
#| label: vf-intersect
#| eval: true

# Data bundled with the package
iso <- readRDS(system.file(
  "extdata", "legacy_2025_isochrones.rds", package = "catchmentACS"
))
acs <- readRDS(system.file(
  "extdata", "sample_alabama_subset.rds", package = "catchmentACS"
))

site_id    <- "AL_SITE_17"
drive_time <- 10L

iso_one <- iso[
  iso$site_id == site_id & iso$drive_time_min == drive_time, ,
  drop = FALSE
]

weighted <- cacs_intersect_weight(
  iso_sf           = iso_one,
  acs_sf           = acs,
  weight_method    = "area",
  keep_tract_audit = TRUE,
  verbose          = FALSE
)
```

The `cacs_tract_audit` attribute of the result lists, for each tract that the
drive-time area overlaps, the area of the overlap (`int_area_m2`) and the area
of the tract (`tract_area_m2`), in square meters on EPSG:5070:

```{r}
#| label: vf-audit
#| eval: true

areas <- attr(weighted, "cacs_tract_audit") |>
  select(GEOID, int_area_m2, tract_area_m2) |>
  arrange(desc(int_area_m2))

areas
```

The coverage weights divide the overlap area by the tract area, and the area
shares divide it by the total overlap area:

```{r}
#| label: vf-weights
#| eval: true

hand <- areas |>
  mutate(
    w_cov  = int_area_m2 / tract_area_m2,        # |I_s n T_j| / |T_j|
    w_mean = int_area_m2 / sum(int_area_m2)      # |I_s n T_j| / sum_k |I_s n T_k|
  )

hand |> select(GEOID, w_cov, w_mean)

c(
  sum_w_cov  = sum(hand$w_cov),    # not 1 in general: each has its own denominator
  sum_w_mean = sum(hand$w_mean)    # 1 by construction
)
```

The coverage weights are `r format(round(hand$w_cov[1], 3))`,
`r format(round(hand$w_cov[2], 3))`, and `r format(round(hand$w_cov[3], 3))`:
the site's own square lies wholly inside the circle, and the other two only
partly. Their sum, `r format(round(sum(hand$w_cov), 3))`, is the overlap area
counted in squares, because these three squares have the same area; across the
whole data set the squares differ in area by about five percent.

The number of people below the poverty level (`B17001_002`) is a count, so the
package estimates it by $\widehat{Y}_s = \sum_j w^{cov}_{sj}\, Y_j$. The code
below takes the three tract estimates from the ACS data and forms this sum:

```{r}
#| label: vf-total
#| eval: true

counts <- acs |>
  sf::st_drop_geometry() |>
  filter(GEOID %in% hand$GEOID, variable == "B17001_002") |>
  select(GEOID, Y = estimate)

hand_total <- hand |>
  select(GEOID, w_cov) |>
  left_join(counts, by = "GEOID")

hand_total

hand_Y_hat       <- sum(hand_total$w_cov * hand_total$Y)   # sum_j w_cov * Y_j
hand_weight_sum  <- sum(hand_total$w_cov)                  # sum_j w_cov

c(hand_Y_hat = hand_Y_hat, hand_weight_sum = hand_weight_sum)
```

The row of the package's result for the same variable has
`estimand_family = "spatial_total"` and `weight_basis = "coverage"`:

```{r}
#| label: vf-pkg-row
#| eval: true

pkg_row <- weighted |>
  filter(variable == "B17001_002") |>
  select(variable, estimate, weight_sum, n_tracts,
         estimand_family, weight_basis)

pkg_row
```

The hand-computed sum, `r format(round(hand_Y_hat, 1), big.mark = ",")`
people, is the `estimate` in the row above, and the sum of the coverage
weights, `r format(round(hand_weight_sum, 3))`, is its `weight_sum`. The same
two areas are recorded for every tract of every site and drive time in the
result, so any row can be recomputed this way.

## References

```{r}
#| label: restore-options
#| include: false

options(old_options)
rm(old_options)
```

```{=html}
<script>
// Chrome can leave out a hat that sits over a bar, such as the hat of the
// estimated average in the formulas above. Written as a spacing character,
// the hat is always drawn.
document.querySelectorAll("math mover > mover + mo").forEach(function (mo) {
  if (mo.textContent === "\u0302") {
    mo.textContent = "\u02C6";
  }
});
</script>
```
