---
title: "Derived rates: definitions and ratio MOEs"
vignette: >
  %\VignetteIndexEntry{Derived rates: definitions and ratio MOEs}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 2
    fig-width: 7
    fig-height: 5
    fig-align: center
    out-width: "85%"
    html-math-method: katex
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| eval: true

library(catchmentACS)
library(dplyr)
library(sf)

# Force network-free, deterministic builds: no provider/Census calls, no cache.
old_options <- options(
  catchmentACS.progress = "off",
  catchmentACS.rate_first_default = FALSE
)
old_cache <- getOption("catchmentACS.cache_enabled", NULL)
on.exit(options(old_options), add = TRUE)
on.exit(options(catchmentACS.cache_enabled = old_cache), add = TRUE)
cacs_set_cache(FALSE, scope = "session")
```

This article is the **method-track** companion to
`vignette("methodology", package = "catchmentACS")` that fixes the *rates*
layer of the estimator — the last of the three theory articles. The overview
*states* that a derived rate is a ratio of two coverage-weighted totals and
that some rates route to a tighter margin formula than others; this article
makes the catalogue precise — the five sanctioned rates, their fixed ACS
numerator and denominator codes, and the baseline policy that decides which
margin-of-error (MOE) formula each one uses — and then does the thing the
overview cannot: it **reproduces a derived rate and its full MOE routing by
hand**, on a bundled fixture, so a methodologist can confirm that the rates the
package reports are the ratios this page describes, computed by the formula this
page assigns. The reference page `?cacs_derive_rates` is the implementation
anchor; this article supplies the rate catalogue and the routing logic behind
its "Rate definitions" and `formula_dispatch` sections.

It is written for a methodologist or a reviewer who wants the rate model on the
page: why a catchment rate is a *ratio of aggregates* rather than a weighted
average of tract rates, which two of the five rates are eligible for the subset
MOE formula and which three are not, why one rate whose numerator genuinely *is*
a subset still routes to the conservative formula by policy, and exactly what
the package records when the tighter formula is requested but its variance
breaks down. We reuse the notation fixed in the overview without re-motivating
it: the isochrone $I_s$, the tract $T_j$, the coverage weight $w^{cov}_{sj}$
(the fraction of tract $T_j$'s area the catchment at site $s$ captured), and a
rate's tract-level numerator and denominator counts $A_j, B_j$. The MOE algebra
itself — the C1 subset formula, the C2 conservative formula, the
$M = 1.645\,SE$ conversion, and the never-silent C1 → C2 fallback — is **not**
re-derived here; it is the subject of
`vignette("theory-moe-propagation", package = "catchmentACS")` and
`?cacs_propagate_moe`, and this article *routes* rates into those formulas
rather than restating them. The construction of $w^{cov}_{sj}$ is likewise
derived in `vignette("theory-spatial-aggregation", package = "catchmentACS")`;
this article takes both as given and concentrates on the rate definitions and
the dispatch.

::: {.callout-note}
## About the data in this article

Every executable cell here is built entirely from the package's bundled
`inst/extdata` fixtures — the synthetic `legacy_2025_isochrones`,
`legacy_2025_sites`, and `sample_alabama_subset` objects — so this article
renders with **no** Census, OSRM, ORS, or other network calls. These fixtures
preserve the production calling contract (sites, drive-time polygons,
tract-level ACS variables) but are **synthetic and illustrative**: they are
included only for examples, tests, and reproducible package checks. They are
**not** live Alabama policy estimates and must not be cited as real ACS results.
Any real analysis requires a fresh live run (see
`vignette("providers", package = "catchmentACS")`).
:::

## A derived rate is a ratio of two coverage-weighted totals

A derived rate is not aggregated the way a count is. It is a *ratio* of two
quantities, each of which is itself a coverage-weighted catchment total. By the
time `cacs_derive_rates()` runs, the spatial stage has already aggregated each
ACS count to the catchment as a coverage-weighted total, and the MOE stage has
attached a margin to each of those totals; a derived rate does **not** re-touch
the geometry. It takes two of those already-aggregated totals — a numerator and
a denominator — and divides. For a rate with tract-level numerator counts $A_j$
and denominator counts $B_j$, the catchment rate at site $s$ is

$$
\widehat{R}_s
= \frac{\displaystyle\sum_j w^{cov}_{sj}\, A_j}{\displaystyle\sum_j w^{cov}_{sj}\, B_j}
= \frac{\widehat{A}_s}{\widehat{B}_s},
\qquad
\widehat{A}_s = \sum_j w^{cov}_{sj}\, A_j,
\quad
\widehat{B}_s = \sum_j w^{cov}_{sj}\, B_j,
$$

where $\widehat{A}_s$ and $\widehat{B}_s$ are exactly the Family-A spatial
totals of the numerator and denominator ACS cells, each carrying the coverage
weight $w^{cov}_{sj}$ derived in
`vignette("theory-spatial-aggregation", package = "catchmentACS")`. The rate
inherits the *coverage* path because both of its ingredients are extensive
counts: half a tract contributes half its poor persons **and** half its
population, and the halves cancel in the ratio only when each is allocated by
the same area fraction.

The order of operations is the whole point, and it is the reason a rate is its
own estimand family: the package **aggregates the numerator and the denominator
separately, each as its own `spatial_total`, and only then divides**. It forms a
ratio of two aggregates, $(\sum_j w^{cov}_{sj} A_j) / (\sum_j w^{cov}_{sj} B_j)$
— it does **not** form a per-tract rate $A_j / B_j$ and average those,
$\sum_j w_j (A_j / B_j)$. The two are not equal, and the ratio-of-aggregates
form is the correct one: a tract that is barely clipped by the catchment should
influence the catchment rate in proportion to how many of its people the
catchment actually captures, which is what weighting the *counts* before
dividing achieves. Averaging tract rates would instead give a sparsely-captured
tract the same say as a fully-enclosed one, and would silently mis-estimate the
catchment rate whenever the tracts differ in size or in their denominator.
Aggregating two totals and dividing once is the areally correct estimator, and
it is the same $\widehat{R}_s$ the overview lists for the `derived_rate` family.

Mechanically, this means both totals must survive intact until the division. The
upstream stages preserve them: `cacs_intersect_weight()` accumulates each ACS
code's coverage-weighted total and its propagated variance into an attribute
(the *aggregation carriers*), `cacs_propagate_moe()` carries that attribute
forward while it computes per-variable margins, and `cacs_derive_rates()` reads
the numerator total $\widehat{A}_s$ and denominator total $\widehat{B}_s$ back
out of it, divides to get the point estimate, and routes the *pair* of totals
(with their variances) into a ratio-MOE formula. Because $\widehat{A}_s$ and
$\widehat{B}_s$ are each random estimates carrying their own propagated variance,
the rate's MOE is a *ratio-variance* problem, and the package offers the two
delta-method forms the Census Bureau's handbook gives — the subset form (C1) and
the conservative general-ratio form (C2). Which form a given rate uses is the
subject of the routing policy below; the algebra of both forms is derived in
`vignette("theory-moe-propagation", package = "catchmentACS")`. The
"preserve both totals, then divide" rule is what makes the rate's margin of
error computable at all.

## The five-rate catalogue

The package computes exactly **five** curated rates. The set is fixed for this
release: `cacs_derive_rates()` accepts the rate list only if it is *identical*
to the sanctioned catalogue, and any addition, deletion, or code substitution
aborts. The catalogue is surfaced publicly as the package object
`cacs_acs_default_rates` (see `?cacs_acs_default_rates`) so the five
`c(num, den)` pairs can be inspected without reaching into package internals.
The motivation for the whitelist is reproducibility and auditability: a single
frozen origin for every published rate, with the numerator/denominator pairing,
the universe, and the subset relationship vetted once per rate rather than
trusted from arbitrary user input. Arbitrary ratios are a separate, future
surface (covered at the end of this article).

We read the catalogue straight off the shipped object so the table below is the
catalogue the code uses, not a transcription of it:

```{r}
#| label: catalogue-object
#| eval: true

# The catalogue is a public object; each entry is a c(num, den) ACS-code pair.
cacs_acs_default_rates
```

Each rate divides one ACS table cell (the numerator) by another (the
denominator). The denominator is the rate's **universe** — the population or
household base the rate is "out of." Reading the universe column is the quickest
way to see what a rate actually measures, because a rate is only interpretable
relative to its base.

| Rate name | Numerator code | Denominator code | What it measures | Universe (denominator) |
|---|---|---|---|---|
| `poverty_rate` | `B17001_002` | `B17001_001` | persons with income below the poverty level $\div$ persons for whom poverty status is determined | population for whom poverty status is determined (`B17001` total) |
| `snap_rate` | `B22003_002` | `B22003_001` | households receiving SNAP / Food Stamps $\div$ all households | households (`B22003` total) |
| `ssi_rate` | `B19056_002` | `B19056_001` | households with Supplemental Security Income $\div$ all households | households (`B19056` total-household universe) |
| `unemp_rate` | `B23025_005` | `B23025_003` | unemployed persons in the civilian labor force $\div$ civilian labor force | civilian labor force (`B23025_003`) |
| `labor_force_participation` | `B23025_002` | `B23025_001` | persons in the labor force $\div$ population 16 and over | population 16 years and over (`B23025_001`) |

Two features of this table do all the work in the next section. First, the
**universe** column is not decoration: it is the denominator's definition, and
whether the numerator is a true subset *of that universe* is what determines MOE
eligibility. Persons below poverty are a subset of the persons for whom poverty
status is determined; persons in the labor force are a subset of the population
16 and over. Second, two rates (`unemp_rate`, `labor_force_participation`) share
the `B23025` employment-status table but divide by *different* cells: the
unemployment rate is out of the **civilian labor force** (`B23025_003`), while
labor-force participation is out of the whole **population 16 and over**
(`B23025_001`). They are different rates with different universes, and — as the
next section shows — they are routed to different MOE formulas under the per-rate
dispatch.

## Subset structure and MOE routing

Every rate above is a ratio, so its margin of error is a *ratio* margin — and the
MOE companion derives that there are two ratio-variance formulas, differing by a
single sign. This is the section to read slowly, because it is where a rate's
*definition* becomes a *formula choice*. We restate the two formulas only enough
to name them as the routing target; the derivation, the guards, and the by-hand
verification all live in
`vignette("theory-moe-propagation", package = "catchmentACS")`.

- **Family C1 — `proportion_subset`.** The Census Bureau's proportion formula,
  which *subtracts* a cross term. It is valid **only when the numerator
  $\widehat{A}_s$ is a true subset of the denominator $\widehat{B}_s$** (every
  unit counted in the numerator is also counted in the denominator), because the
  subtraction credits the positive correlation a subset relationship guarantees.
  It is the **tighter** of the two estimates when it is valid; its fragility is
  that the subtracted bracket can go negative.
- **Family C2 — `general_ratio_conservative`.** The general-ratio formula, which
  *adds* the same cross term. It is valid for any ratio, is **always
  non-negative**, and is therefore the **conservative** (slightly wider)
  estimate. It is the package default.

### The subset relation versus the routing policy

Whether a rate's numerator is a mathematical *subset* of its denominator is a
fact about ACS table structure. Whether the package *routes* that rate to the
tighter C1 formula is a separate **baseline policy** decision. The two are not
the same, and conflating them is the error this section exists to prevent. The
package records the policy as a per-rate C1-eligibility flag and a per-rate
default formula. The public table below mirrors that locked routing policy
without requiring users to reach into unexported package internals:

```{r}
#| label: routing-tables
#| eval: true

tibble::tibble(
  rate = c(
    "poverty_rate",
    "snap_rate",
    "ssi_rate",
    "unemp_rate",
    "labor_force_participation"
  ),
  c1_eligible = c(TRUE, FALSE, FALSE, FALSE, TRUE),
  default_formula = c(
    "proportion_subset",
    "general_ratio_conservative",
    "general_ratio_conservative",
    "general_ratio_conservative",
    "proportion_subset"
  )
)
```

Read the two columns together. The five rates fall into **three** routing
classes, and it is worth naming each precisely because the middle one is easy to
get wrong:

**C1-eligible (the subset proportions): `poverty_rate` and
`labor_force_participation`.** For both, the numerator is a textbook subset of
the denominator's universe — poor persons within the poverty-status universe,
labor force within the 16-and-over population. These are the only two rates the
package will compute with the tighter Family C1 formula, and they are the two
whose `auto` default formula is `proportion_subset`.

**C2 by household structure: `snap_rate` and `ssi_rate`.** Both are household
rates — households receiving a benefit divided by all households. The package
does not assert a subset *proportion* relationship for them at this baseline, so
they route to the conservative Family C2 formula. Their numerator is genuinely a
subset of "all households," but the curated policy treats household-benefit rates
as general ratios in this release; they are **C2 by policy**.

**Subset numerator, but C2 by policy: `unemp_rate`.** This is the case to read
carefully. The unemployment rate's numerator, the unemployed (`B23025_005`),
genuinely **is** a subset of its denominator, the civilian labor force
(`B23025_003`) — every unemployed person is in the labor force. The subset
relation holds. It is therefore **not a "non-subset" rate** — calling it that
would misstate the ACS table structure. The baseline policy for this release
nonetheless routes `unemp_rate` to the **conservative** Family C2 formula rather
than to C1. So `unemp_rate` sits deliberately in its own class: *its numerator is
a subset, but its baseline MOE formula is the conservative general ratio.* The
correct description is "subset, but C2 by baseline policy," **not** "non-subset."
A future-release transition of this rate to C1 is one of the package's open
questions, and the reference page flags it explicitly.

### `formula_dispatch`: how the routing is selected

`cacs_derive_rates()` takes a `formula_dispatch` argument that selects how the
five rates are assigned to formulas. It has three modes:

| `formula_dispatch` | Effect on the five rates |
|---|---|
| `"general_ratio_conservative"` *(default)* | Broadcast Family C2 to **every** rate, including the two C1-eligible ones. Under it, no rate uses the subtracting formula, so no rate can produce a negative subset variance, and every published margin is the always-defined conservative one — the most conservative, fully reproducible setting. |
| `"proportion_subset"` | Request Family C1 for the rates that are C1-eligible (`poverty_rate`, `labor_force_participation`); the three non-eligible rates (`snap_rate`, `ssi_rate`, `unemp_rate`) are **downgraded to C2** by policy, not silently — the downgrade is recorded in the run's rate-provenance attribute. |
| `"auto"` | Per-rate dispatch from the catalogue: C1 for the two eligible rates, C2 for the other three. The "use each rate's policy default" setting, and the mode that reproduces the catalogue's intended formula for each rate in a single call. |

Note that `"proportion_subset"` and `"auto"` produce the *same* routing for the
five curated rates, because the only rates `"proportion_subset"` upgrades to C1
are exactly the two the catalogue already marks C1 — and the three non-eligible
rates are pinned to C2 under both. The difference between the two modes is in
their *intent* for a future open catalogue, not in their effect on this release's
five rates.

### The C1 → C2 fallback (routing's safety valve)

The dispatch above decides what each rate *requests*. Requesting C1 for an
eligible rate is a *request*, not a guarantee, and there is one more layer below
it — the reason C1 is safe to request at all. The Family C1 bracket
$\mathrm{Var}(A) - (A/B)^2\,\mathrm{Var}(B)$ can come out **negative** on real
aggregated inputs, in which case its square root is undefined. When that happens,
the package does not emit a missing value or a silent C1 number: it
**recomputes that cell's margin with the always-non-negative Family C2 formula**
and records the substitution on the row — `moe_formula_requested` stays
`"proportion_subset"` (the request is preserved), `moe_formula_effective` becomes
`"general_ratio_conservative"` (C2 produced the value), `moe_fallback` is set to
`TRUE`, and `moe_fallback_reason` is `"negative_variance"`. The fallback is never
silent, and a single end-of-run summary reports the counts.

We do **not** re-derive that fallback here — its mechanics, the status-column
contract, and the other degenerate outcomes (zero denominator, missing input
margin) are derived and verified in
`vignette("theory-moe-propagation", package = "catchmentACS")` and documented in
`?cacs_propagate_moe`. What matters for *this* article is that the fallback is
the reason a C1-eligible rate can still legitimately report a C2 margin: routing
proposes C1, and the variance guard disposes. The numerical section below shows
exactly this happening to one rate while a second C1-eligible rate stays on C1,
in the same catchment.

## SSI rate: the current definition

The SSI rate deserves a one-line clarification because its universe is easy to
misread. The current definition is

$$
\widehat{R}^{\text{ssi}}_s
= \frac{\widehat{A}_s\,(\texttt{B19056\_002})}{\widehat{B}_s\,(\texttt{B19056\_001})},
$$

that is, **households with Supplemental Security Income (`B19056_002`) divided by
the `B19056` total-household universe (`B19056_001`)**. The base is *households*,
not persons and not households-with-some-other-income — it is the full household
count of the `B19056` table. We surface it explicitly only because the `B19056`
table also publishes person-level and income-bracket cells that a reader might
reach for; the curated catalogue fixes the household-over-household pairing so
the rate's universe is unambiguous. Because the numerator and denominator are
household-defined cells rather than a person-level subset, the SSI rate routes to
the conservative Family C2 formula by default (it is not C1-eligible). This is
the definition the package computes for this release.

## Bivariate tract exposure: `n_tracts_num` and `n_tracts_den`

A `spatial_total` row carries an `n_tracts` count — the number of tracts that
contributed to that one aggregation. A rate row cannot use that single slot
honestly, because a rate is built from *two* aggregations whose tract sets need
not match. The numerator and denominator of a curated rate come from different
ACS table cells, and the Census Bureau suppresses cells independently: a
catchment can have a valid numerator contribution from a tract that has a
suppressed denominator there, or the reverse. So a rate row carries **two** counts
instead of one, populated independently:

- **`n_tracts_num`** — the number of tracts that contributed a valid count to the
  rate's **numerator** total $\widehat{A}_s$.
- **`n_tracts_den`** — the number of tracts that contributed a valid count to the
  rate's **denominator** total $\widehat{B}_s$.

These two counts are reported **separately** rather than collapsed to a single
`min(num, den)` precisely so a reviewer can *see* an asymmetry directly. If
`n_tracts_num` and `n_tracts_den` disagree for a rate, that is a visible signal
that the numerator and denominator were aggregated over different tract sets —
exactly the kind of suppression asymmetry that a collapsed count would hide.

The `n_tracts` column itself is **`NA_integer_` on every rate row**. This is not
a gap to be filled — it is the contract. `n_tracts` describes a single
aggregation, and a rate is *derived* from two, not aggregated as one; the
meaningful per-tract counts for a rate live in the two `_num` / `_den` slots, so
the single slot is deliberately left empty. Symmetrically, every inherited
**source-variable** row (the spatial totals and proxies passed through from the
MOE stage) carries `n_tracts_num = NA_integer_` and `n_tracts_den = NA_integer_`,
because those two columns are reserved for rates. In the carrier-missing edge case
— when a rate's numerator or denominator carrier is absent for a cell — both
`_num` and `_den` collapse to `NA_integer_` as well, and the diagnostic surface
for that condition is the row's `failure_origin`, not the count columns. The full
population contract is in the "Policy on n_tracts_num and n_tracts_den" section of
`?cacs_derive_rates`.

## Deferred: custom and multi-numerator rates

The five-rate catalogue is intentionally closed for this release, and two
extensions are explicitly reserved for a future (v1.1) surface rather than
smuggled into the current one.

- **Arbitrary user-supplied ratios** are reserved for a separate function,
  `cacs_derive_rates_custom()`, that will accept user-defined numerator/
  denominator pairs with an explicit per-rate formula assignment. Any ratio
  outside the five sanctioned pairs is rejected at the argument boundary today;
  keeping arbitrary ratios out of the curated path preserves the single-origin
  audit trail, so every rate the current function emits is one of the five vetted
  definitions.
- **Multi-numerator rates** — rates whose numerator is a *sum* of several ACS
  cells rather than a single cell — are also deferred. The canonical example is a
  true child poverty rate, which must sum several age-by-sex cells of the
  `B17001` table rather than divide a single cell. A single-cell shortcut for such
  a rate is a silent statistical error, so it is excluded from the curated
  catalogue rather than approximated.

Both are noted here so a reader knows the current function's scope is the *exact*
five-rate catalogue, and that the path for anything beyond it already has a
reserved home. The catalogue object carries this stability promise in its own
documentation: see `?cacs_acs_default_rates`.

## Numerical verification: the rates match the package

This is the section that earns a methodologist's trust. We compute the five rates
with the *real* pipeline on a bundled fixture, offline, and then do four things
the prose above only asserts. We **reproduce a C2 rate and its margin by hand**
and assert equality with the package to machine precision. We **reproduce a
C1-eligible rate whose subset formula succeeds**, confirm its margin is the
tighter subset value (not the conservative one), and match it by hand. On the
same catchment we watch a **second C1-eligible rate fall back to C2** because its
subset variance goes negative, and reproduce *that* margin by hand too. And we
print the **`moe_formula_effective` for all five rates** so the C1-vs-C2 routing
is visible in one table. Finally we force the **carrier-missing** branch and show
it surfaces as `failure_origin = "carrier"`, distinct from any MOE fallback. If
the catalogue or the routing on this page were not the ones the code runs, one of
these checks would fail.

We build the catchment from the fixtures with `cacs_intersect_weight()`, then run
`cacs_propagate_moe()` and `cacs_derive_rates()` — the exact pipeline path of a
real run, on frozen inputs. We use the synthetic site `AL_SITE_03` at a 15-minute
drive time, whose catchment intersects three tracts, because it happens to
exercise all three routing outcomes at once:

```{r}
#| label: vf-pipeline
#| eval: true

# These fixtures are exactly the shapes pipeline stages 1-2 would have returned.
iso <- readRDS(system.file(
  "extdata", "legacy_2025_isochrones.rds", package = "catchmentACS"
))
acs <- readRDS(system.file(
  "extdata", "sample_alabama_subset.rds", package = "catchmentACS"
))

site_id    <- "AL_SITE_03"
drive_time <- 15L

iso_one <- iso[
  iso$site_id == site_id & iso$drive_time_min == drive_time, ,
  drop = FALSE
]

# keep_tract_audit preserves the per-tract carriers we recompute the rate from.
weighted <- cacs_intersect_weight(
  iso_sf           = iso_one,
  acs_sf           = acs,
  weight_method    = "area",
  keep_tract_audit = TRUE,
  verbose          = FALSE
)
propagated <- cacs_propagate_moe(weighted, verbose = FALSE)

# formula_dispatch = "auto" routes each rate to its catalogue default (C1 for
# the two eligible rates, C2 for the other three).
rates <- suppressWarnings(
  cacs_derive_rates(propagated, formula_dispatch = "auto", verbose = FALSE)
)
```

The numerator and denominator totals the rate divides are the *aggregation
carriers* — the Family-A coverage totals $\widehat{A}_s$, $\widehat{B}_s$ and
their accumulated variances that `cacs_intersect_weight()` preserves as an
attribute for exactly this purpose. We pull them so the by-hand checks read from
the same numbers the package divides:

```{r}
#| label: vf-carriers
#| eval: true

z    <- 1.645
carr <- attr(weighted, "cacs_aggregation_carriers")

# A tiny accessor: the coverage total and its accumulated variance for one cell.
get <- function(code, col) carr[[col]][carr$variable == code]

carr |>
  filter(variable %in% c(
    "B22003_002", "B22003_001",   # snap_rate  num / den
    "B23025_002", "B23025_001",   # labor_force_participation num / den
    "B17001_002", "B17001_001"    # poverty_rate num / den
  )) |>
  select(variable, est_total, var_total_raw, weight_sum, n_tracts)
```

### Check 1 — a C2 rate by hand (`snap_rate`)

`snap_rate` is a household-benefit ratio, so it routes to Family C2 under every
dispatch mode. Its point estimate is the ratio of the two coverage totals
$\widehat{R}_s = \widehat{A}_s / \widehat{B}_s$, and its margin is the C2 ratio
form, recovered from the carriers' accumulated variances
$\mathrm{Var}(A) = (M_A/z)^2$ and $\mathrm{Var}(B) = (M_B/z)^2$ as
$M = z\sqrt{\mathrm{Var}(A) + (A/B)^2\,\mathrm{Var}(B)}\,/\,|B|$:

```{r}
#| label: vf-c2-hand
#| eval: true

A_snap    <- get("B22003_002", "est_total")
B_snap    <- get("B22003_001", "est_total")
VarA_snap <- get("B22003_002", "var_total_raw")
VarB_snap <- get("B22003_001", "var_total_raw")

p_snap        <- A_snap / B_snap                                     # the ratio A / B
hand_est_snap <- p_snap
hand_moe_snap <- z * sqrt(VarA_snap + p_snap^2 * VarB_snap) / abs(B_snap)

pkg_snap <- rates |>
  filter(variable == "snap_rate") |>
  select(estimate, moe, moe_formula_effective, moe_fallback)

list(
  A = A_snap, B = B_snap,
  hand_estimate = hand_est_snap,
  hand_moe_C2   = hand_moe_snap,
  pkg           = pkg_snap
)
```

Both reproduce exactly. The package routed `snap_rate` through
`moe_formula_effective = "general_ratio_conservative"` with no fallback; its
reported estimate (about `0.2506`) is the by-hand ratio $A/B$, and its reported
margin (about `0.0345`) is the by-hand C2 root:

```{r}
#| label: vf-c2-assert
#| eval: true

c(
  estimate_match = isTRUE(all.equal(hand_est_snap, pkg_snap$estimate)),
  moe_match      = isTRUE(all.equal(hand_moe_snap, pkg_snap$moe)),
  moe_abs_diff   = abs(hand_moe_snap - pkg_snap$moe)
)
```

The rate on this page and the rate in the code are the same arithmetic.

### Check 2 — a C1-eligible rate whose subset formula succeeds (`labor_force_participation`)

`labor_force_participation` is C1-eligible (labor force $\subseteq$ population
16+), so under `auto` it is computed with the **tighter** Family C1 formula
$M = z\sqrt{\mathrm{Var}(A) - (A/B)^2\,\mathrm{Var}(B)}\,/\,|B|$ — *provided the
bracket stays non-negative.* We compute the bracket by hand, confirm it is
positive, and form the C1 margin:

```{r}
#| label: vf-c1-hand
#| eval: true

A_lfp    <- get("B23025_002", "est_total")
B_lfp    <- get("B23025_001", "est_total")
VarA_lfp <- get("B23025_002", "var_total_raw")
VarB_lfp <- get("B23025_001", "var_total_raw")

p_lfp       <- A_lfp / B_lfp
inside_C1   <- VarA_lfp - p_lfp^2 * VarB_lfp           # C1 bracket; must be >= 0
hand_moe_C1 <- z * sqrt(inside_C1) / abs(B_lfp)
hand_moe_C2 <- z * sqrt(VarA_lfp + p_lfp^2 * VarB_lfp) / abs(B_lfp)  # for contrast

list(
  C1_inside_bracket = inside_C1,          # > 0, so C1 is admissible
  hand_moe_C1       = hand_moe_C1,
  hand_moe_C2       = hand_moe_C2          # the wider value C1 avoids
)
```

The bracket is positive, so C1 is admissible, and the C1 margin (about `0.0240`)
is markedly **tighter** than the conservative C2 value (about `0.0739`) on the
identical inputs — that gap is exactly the cross term C1 credits and C2 does not.
The package agrees, and reports `proportion_subset` with no fallback:

```{r}
#| label: vf-c1-assert
#| eval: true

pkg_lfp <- rates |>
  filter(variable == "labor_force_participation") |>
  select(estimate, moe, moe_formula_effective, moe_fallback, moe_fallback_reason)

pkg_lfp

c(
  moe_match   = isTRUE(all.equal(hand_moe_C1, pkg_lfp$moe)),
  is_C1       = pkg_lfp$moe_formula_effective == "proportion_subset",
  no_fallback = isTRUE(!pkg_lfp$moe_fallback)
)
```

The package's margin equals the hand-computed **subset** root, its effective
formula is `proportion_subset`, and it did not fall back. The tighter formula was
requested, was admissible, and was used.

### Check 3 — a C1-eligible rate that falls back to C2 (`poverty_rate`)

On this *same* catchment, the other C1-eligible rate behaves differently, and
that is the point of choosing it. `poverty_rate`'s numerator is *persons in
poverty* (`B17001_002`) and its denominator is *persons for whom poverty status
is determined* (`B17001_001`); the numerator is a true subset of the denominator,
so it is also requested as `proportion_subset` under `auto`. But its C1 bracket
comes out **negative**, so the subset formula is inadmissible and the package
falls back to C2:

```{r}
#| label: vf-fallback-hand
#| eval: true

A_pov    <- get("B17001_002", "est_total")
B_pov    <- get("B17001_001", "est_total")
VarA_pov <- get("B17001_002", "var_total_raw")
VarB_pov <- get("B17001_001", "var_total_raw")

p_pov       <- A_pov / B_pov
inside_C1   <- VarA_pov - p_pov^2 * VarB_pov           # C1 bracket; < 0 here
hand_moe_C2 <- z * sqrt(VarA_pov + p_pov^2 * VarB_pov) / abs(B_pov)

list(
  C1_inside_bracket = inside_C1,    # < 0, so C1 is inadmissible -> fall back
  hand_moe_C2       = hand_moe_C2   # the value the fallback should report
)
```

The bracket is negative, so C1 cannot produce a real margin and the routing's
safety valve fires. The package records the substitution exactly as the
never-silent contract requires, and the reported margin is the **C2** root, not a
missing value:

```{r}
#| label: vf-fallback-assert
#| eval: true

pkg_pov <- rates |>
  filter(variable == "poverty_rate") |>
  select(estimate, moe,
         moe_formula_requested, moe_formula_effective,
         moe_fallback, moe_fallback_reason)

pkg_pov

c(
  moe_match        = isTRUE(all.equal(hand_moe_C2, pkg_pov$moe)),
  requested_is_C1  = pkg_pov$moe_formula_requested == "proportion_subset",
  effective_is_C2  = pkg_pov$moe_formula_effective == "general_ratio_conservative",
  fallback_flagged = isTRUE(pkg_pov$moe_fallback),
  reason_recorded  = pkg_pov$moe_fallback_reason == "negative_variance"
)
```

Every flag lands: the row asked for `proportion_subset`, its subset variance went
negative, and the conservative formula produced the value, with
`moe_fallback = TRUE` and `moe_fallback_reason = "negative_variance"` recorded so
the path is auditable. The fallback margin reproduces the hand-computed C2 root
exactly. Two C1-eligible rates, the same catchment, opposite outcomes — and both
fully recorded.

### The routing made visible: `moe_formula_effective` across all five rates

Finally, the single table the whole routing section describes — every rate, its
requested formula, the formula whose value was actually written, and the
bivariate tract exposure on each rate row:

```{r}
#| label: vf-routing-table
#| eval: true

rates |>
  filter(estimand_family == "derived_rate") |>
  select(variable, estimate, moe,
         moe_formula_requested, moe_formula_effective,
         moe_fallback, moe_fallback_reason,
         n_tracts, n_tracts_num, n_tracts_den)
```

Read the `moe_formula_effective` column straight down. `snap_rate`, `ssi_rate`,
and `unemp_rate` are `general_ratio_conservative` — the three C2 rates, exactly as
the policy assigns (`unemp_rate` among them, its subset numerator
notwithstanding). `labor_force_participation` is `proportion_subset` — a
C1-eligible rate whose subset formula was admissible. And `poverty_rate` reads
`general_ratio_conservative` with `moe_fallback = TRUE` — a C1-eligible rate that
fell back. The bivariate exposure is visible too: every rate row carries
`n_tracts = NA` while `n_tracts_num` and `n_tracts_den` hold the per-side tract
counts.

### Carrier-missing routing

One last routing outcome is not a formula choice at all: a rate whose numerator
or denominator **carrier is absent** for a cell cannot be computed, and the
package records that as an *upstream carrier failure*, distinct from any MOE
fallback. The default ACS catalogue supplies every carrier the five rates need,
so this arises only in direct pipelines that drop a variable. We force it by
removing the SSI numerator cell `B19056_002` from the ACS extract and rerunning:

```{r}
#| label: vf-carrier-missing
#| eval: true

acs_drop <- acs[sf::st_drop_geometry(acs)$variable != "B19056_002", ]

weighted_d   <- cacs_intersect_weight(
  iso_sf = iso_one, acs_sf = acs_drop,
  weight_method = "area", verbose = FALSE
)
propagated_d <- cacs_propagate_moe(weighted_d, verbose = FALSE)
rates_d      <- suppressWarnings(cacs_derive_rates(propagated_d, verbose = FALSE))

rates_d |>
  filter(estimand_family == "derived_rate") |>
  select(variable, estimate, moe,
         moe_fallback_reason, failure_origin, n_tracts_num, n_tracts_den)
```

`ssi_rate` now reports `estimate = NA`, `moe = NA`, `failure_origin = "carrier"`,
and both `n_tracts_num` and `n_tracts_den` collapse to `NA` — while the other four
rates compute normally. Critically, its `moe_fallback_reason` stays `"n/a"`, not a
fallback reason: a missing carrier is an upstream failure surfaced through
`failure_origin`, **not** an MOE-formula fallback. The two diagnostic surfaces
stay distinct, exactly as the schema promises.

To summarize the check: a C2 rate (`snap_rate`) and its conservative margin
reproduce by hand to the floating-point floor; a C1-eligible rate
(`labor_force_participation`) reports the *tighter* subset margin, also matched by
hand, and is recorded as `proportion_subset` with no fallback; a second
C1-eligible rate (`poverty_rate`) on the same catchment falls back to C2 because
its subset variance goes negative, with the substitution recorded as
`moe_formula_effective = "general_ratio_conservative"`, `moe_fallback = TRUE`, and
`moe_fallback_reason = "negative_variance"`, its margin again matched by hand; the
`moe_formula_effective` column makes the C1-vs-C2 routing of all five rates
visible at once; and a dropped carrier surfaces as `failure_origin = "carrier"`
rather than an MOE fallback. The catalogue and the routing on this page are the
catalogue and the routing the code runs.

## Where this sits

This article fixed the rate catalogue and routed each rate into the MOE
machinery. Its companions cover the geometry and the uncertainty algebra it
depends on, and the overview ties all three together; each reuses the notation
fixed in the overview.

- **The estimator overview**, with the shared notation, the six-family table that
  lists the `derived_rate` form, and the place of derived rates in the pipeline:
  `vignette("methodology", package = "catchmentACS")`.
- **Spatial aggregation** — how the coverage weight $w^{cov}_{sj}$ is built and
  how each of a rate's two totals $\widehat{A}_s$, $\widehat{B}_s$ is aggregated
  as a `spatial_total`:
  `vignette("theory-spatial-aggregation", package = "catchmentACS")`.
- **MOE propagation** — the derivation of the two ratio-variance formulas (the
  subtracting C1 and the adding C2), the SE ↔ MOE relation, the numerical guards,
  and the never-silent C1 → C2 negative-variance fallback this article routes
  rates into: `vignette("theory-moe-propagation", package = "catchmentACS")` and
  `?cacs_propagate_moe`.
- **Function reference** — the `formula_dispatch` modes, the per-rate downgrade
  recording, the `n_tracts_num` / `n_tracts_den` contract, and the
  carrier-missing behavior are documented in `?cacs_derive_rates`; the frozen
  five-rate catalogue object and its stability promise are in
  `?cacs_acs_default_rates`.

For the applied, run-it path that produces reporting tables with these rates and
their margins of error from the same fixtures, see
`vignette("getting-started", package = "catchmentACS")` and
`vignette("alabama-tutorial", package = "catchmentACS")`.
