---
title: "MOE propagation: margins of error"
vignette: >
  %\VignetteIndexEntry{MOE propagation: margins of error}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 2
    fig-width: 7
    fig-height: 5
    fig-align: center
    out-width: "85%"
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| eval: true

library(catchmentACS)
library(dplyr)
library(sf)

# Force network-free, deterministic builds: no provider/Census calls, no cache.
old_options <- options(
  catchmentACS.progress = "off",
  catchmentACS.rate_first_default = FALSE
)
old_cache <- getOption("catchmentACS.cache_enabled", NULL)
on.exit(options(old_options), add = TRUE)
on.exit(options(catchmentACS.cache_enabled = old_cache), add = TRUE)
cacs_set_cache(FALSE, scope = "session")
```

This article is the **method-track** companion to
`vignette("methodology", package = "catchmentACS")` that develops the
*uncertainty* half of the estimator. The overview *names* the four
margin-of-error (MOE) formula families and states the $M = 1.645\,SE$
conversion; this article *derives* them — the SE/MOE relation, the weighted
root-sum-square for sums and means, the two ratio-variance forms, the
never-silent C1 → C2 fallback, and the numerical guards — and then does the
thing the overview cannot: it **reproduces the package's MOE arithmetic by
hand** on a bundled fixture and **triggers the fallback on purpose**, so a
methodologist can confirm that the formulas on this page are the formulas the
code runs. The reference page `?cacs_propagate_moe` is the implementation
anchor; this article supplies the mathematics behind its "MOE formula families"
section.

It is written for a reader who wants the propagation model on the page: why
variances must accumulate on the standard-error scale, why a weighted sum and a
weighted mean share one combination rule, why a ratio has two variance formulas
and which one is the default, and exactly what the package records when the
tighter formula breaks down. We reuse the notation fixed in the overview without
re-motivating it: the isochrone $I_s$, the tract $T_j$, the coverage weight
$w^{cov}_{sj}$ and the normalized mean weight $w^{mean}_{sj}$, a tract-level
count $Y_j$, a scalar $X_j$, a ratio's numerator and denominator counts
$A_j, B_j$, and the published 90% tract margin of error $M_j$. The *spatial*
half — how those weights are built and how the point estimates
$\widehat{Y}_s$, $\widehat{\bar X}_s$, $\widehat{R}_s$ aggregate — is derived in
`vignette("theory-spatial-aggregation", package = "catchmentACS")`; this article
takes the weights as given and carries the published uncertainty through them.

::: {.callout-note}
## About the data in this article

Every executable cell here is built entirely from the package's bundled
`inst/extdata` fixtures — the synthetic `legacy_2025_isochrones`,
`legacy_2025_sites`, and `sample_alabama_subset` objects — so this article
renders with **no** Census, OSRM, ORS, or other network calls. These fixtures
preserve the production calling contract (sites, drive-time polygons,
tract-level ACS variables) but are **synthetic and illustrative**: they are
included only for examples, tests, and reproducible package checks. They are
**not** live Alabama policy estimates and must not be cited as real ACS results.
Any real analysis requires a fresh live run (see
`vignette("providers", package = "catchmentACS")`).
:::

## Why MOE propagation needs its own algebra

Every ACS estimate ships with a published margin of error, and a catchment
estimate is a *function of many* such estimates — a weighted sum of tract counts,
a weighted blend of tract scalars, a ratio of two aggregated totals. The
published MOEs describe the uncertainty of the *inputs*; the question this
article answers is how that uncertainty travels to the *output*. Dropping it is
not an option: a coverage-weighted count of persons in poverty with no margin of
error invites a false precision the underlying survey never claimed. Carrying it
naively is just as wrong, because margins of error do not add — a sum of two
estimates each with margin $M$ does not have margin $2M$.

The propagation rule is the standard one for a linear (or locally linear)
combination of approximately independent estimates: **variances add, margins of
error do not.** So the whole apparatus reduces to three moves repeated in four
guises — convert each component margin to a variance, combine the variances under
the appropriate model, and convert the combined variance back to a margin. The
four families differ only in *how* the variances combine: a weighted sum
(Family A), a weighted mean (Family B), a subset proportion (Family C1), and a
general ratio (Family C2). The U.S. Census Bureau's published guidance on
approximating the margin of error of derived estimates (its ACS handbook for data
users) gives all four; the package reimplements them so the formula, its guards,
and its audit trail live in one place.

Two modeling assumptions sit under everything below, and both are inherited from
the spatial-aggregation model rather than introduced here. First, the tract-level
ACS estimates entering a catchment aggregate are treated as **approximately
independent**, which is what lets their variances add without covariance terms;
this is the same assumption the Census Bureau's own sum-of-estimates
approximation makes, and it is generally slightly conservative for positively
correlated neighbors. Second, the **weights are treated as fixed** design
quantities, not random — the coverage and mean weights are geometric ratios of
measured areas, so they carry no sampling variance of their own. Both are stated
plainly so a reader knows exactly what the margins below do and do not capture.

## Standard errors, margins of error, and the 90% convention

The U.S. Census Bureau publishes ACS uncertainty as a **margin of error** (MOE):
the half-width of a confidence interval, reported at the **90%** level. An
estimate $\widehat{\theta}$ ships as $\widehat{\theta} \pm M$, and the interval
$[\widehat{\theta} - M,\; \widehat{\theta} + M]$ is the Bureau's 90% interval.
The MOE is a *reported* quantity, but it is not the natural quantity for
*combining* uncertainties — variances are. A margin of error and a standard
error (SE) are interconvertible through a single confidence multiplier $z$,

$$
M = z\, SE, \qquad SE = \frac{M}{z},
$$

and at the 90% level the Census multiplier is

$$
z = 1.645,
\qquad\text{so}\qquad
M = 1.645\, SE, \qquad SE = \frac{M}{1.645}.
$$

The constant $1.645$ is the Bureau's official rounded value of the standard
normal 95th percentile; the exact quantile $\Phi^{-1}(0.95) \approx 1.6448536$
differs only in the fourth decimal place, and the package adopts the rounded
$1.645$ at the default level so its output is byte-compatible with hand
calculations done the Bureau's way. (The multiplier is configurable through the
`level` argument; any other level resolves $z$ from the exact symmetric
two-tailed normal quantile instead. Everything below is at the default
$z = 1.645$.)

**Why the SE scale matters.** Variances of independent quantities add; margins
of error do not. If two independent estimates have margins $M_1$ and $M_2$, the
margin of their sum is **not** $M_1 + M_2$. The correct route is to drop to the
SE scale, where variances are additive, combine, and convert back:

$$
M_{\text{sum}}
= z\, SE_{\text{sum}}
= z \sqrt{SE_1^2 + SE_2^2}
= z \sqrt{\left(\tfrac{M_1}{z}\right)^2 + \left(\tfrac{M_2}{z}\right)^2}.
$$

Every formula in this article is an instance of that pattern: **divide each
component MOE by $z$ to a standard error, combine the standard errors (or
variances) under an independence assumption, multiply the combined SE back by
$z$.** The package exposes the two conversions directly as `cacs_se_to_moe()`
and `cacs_moe_to_se()` (exact inverses at a given level), and both are NA-safe:
a missing input returns `NA`, never a silent zero. Internally the same constant
divides MOEs into variances at the start of each family's calculation and
multiplies back at the end.

```{r}
#| label: se-moe-roundtrip
#| eval: true

# M = 1.645 * SE and its exact inverse, on the same vector. NA passes through.
se  <- c(50, 100, NA)
moe <- cacs_se_to_moe(se, level = 0.90)
c(moe = moe, recovered_se = cacs_moe_to_se(moe, level = 0.90))
```

The first SE of `50` becomes a margin of `82.25`, the second `100` becomes
`164.5`, and `cacs_moe_to_se()` recovers the originals exactly; the `NA` survives
both directions. This round-trip is the backbone of every combination below.

## Families A and B: the weighted root-sum-square

The first two families propagate margins through a *linear* combination of tract
estimates, and they share one closed form. They differ only in the weight that
multiplies each component margin — and that weight is exactly the one the
matching estimand family uses for its point estimate, so the uncertainty travels
through the *same* geometry as the estimate.

### Family A — spatial total (weighted sum of counts)

A `spatial_total` is the coverage-weighted sum of tract counts,
$\widehat{Y}_s = \sum_j w^{cov}_{sj}\, Y_j$, with the coverage weight derived in
the spatial-aggregation article. Each tract contributes $w^{cov}_{sj}\, Y_j$, an
estimate whose standard error is $w^{cov}_{sj}\, (M_j / z)$, because a fixed
weight scales an SE linearly. Under approximate independence the variances add,
so the variance of the catchment total is the sum of the squared
coverage-weighted component SEs, and the margin of error is that combined SE
lifted back by $z$:

$$
M = z \sqrt{\sum_j \left(\frac{w^{cov}_{sj}\, M_j}{z}\right)^2}.
$$

This is a **root-sum-square of coverage-weighted component MOEs.** Read the
structure: each published margin $M_j$ is divided by $z$ to a standard error,
scaled by that tract's coverage weight (the fraction of the tract the catchment
captured), squared and summed into a variance, and the result is square-rooted
and multiplied by $z$ back to a margin. A tract captured whole
($w^{cov}_{sj} = 1$) contributes its full margin; a tract clipped to a tenth of
its area contributes a tenth of it. The root-sum-square — rather than a plain sum
— is the arithmetic statement that independent errors partially cancel: the
combined margin grows like the square root of the number of comparable terms, not
linearly.

### Family B — weighted scalar / proxy mean

A scalar proxy — per-capita income, a median-income proxy — is the
normalized-mean-weighted blend $\widehat{\bar X}_s = \sum_j w^{mean}_{sj}\, X_j$,
with the mean weights that sum to one within a site and drive-time cell. Because
the mean weights are also fixed design quantities, the propagation is
structurally identical to Family A; only the weight changes from $w^{cov}$ to
$w^{mean}$:

$$
M = z \sqrt{\sum_j \left(\frac{w^{mean}_{sj}\, M_j}{z}\right)^2}.
$$

The single difference from Family A is the weight. The coverage weight allocates a
*fraction* of each tract's amount and is unbounded in sum; the mean weight is a
*share* that is pre-normalized to sum to one. So a Family-A margin scales with how
much area the catchment captured, while a Family-B margin is a weighted average of
component margins on the same per-unit scale as the proxy itself — consistent with
the point estimate, which is a convex combination of the tract values. In both
families the weight that appears in the margin is the *same* weight that appears
in the point estimate: coverage for a total, normalized mean for a proxy. This is
the propagation analogue of the central spatial-aggregation point — a count and an
average are aggregated by different weights — and it is why a single
root-sum-square form serves both with only the weight swapped. (The implementation
carries a raw accumulated variance per component and an explicit `weight_basis` so
the two paths never cross; the per-cell normalization for Family B is the
package's responsibility, documented in `?cacs_intersect_weight`. A
`metadata_only` row carries no estimate and therefore no margin.)

The numerical check below verifies Family A by hand against the package on a
catchment with genuinely partial coverage, so the weights actually bite.

## Families C1 and C2: ratio variance, subset versus conservative

A `derived_rate` is not linear in its inputs — it is a *ratio*
$\widehat{R}_s = A/B$ of two aggregated coverage totals, where both the numerator
total $A = \sum_j w^{cov}_{sj} A_j$ and the denominator total
$B = \sum_j w^{cov}_{sj} B_j$ are random estimates carrying their own propagated
variance from Family A. A ratio has no exact finite-sample variance, so
propagation uses the standard first-order (delta-method) linearization, and the
package offers the two forms the Census Bureau's handbook gives. Both are stated
on the variance scale; the reported margin is recovered as
$M = z\sqrt{\widehat{\mathrm{Var}}(A/B)}$ at the end. They differ by exactly one
sign, and that sign is the hinge of both their statistical meaning and their
numerical behavior.

### Family C1 — proportion subset

When the numerator $A$ is a true **subset** of the denominator $B$ — every person
counted in $A$ is also counted in $B$, as persons in poverty are a subset of the
population for whom poverty status is determined — the numerator and denominator
errors are positively correlated, and the Bureau's proportion formula *subtracts*
a cross term to credit that correlation:

$$
\widehat{\mathrm{Var}}(A/B)
\approx \frac{\mathrm{Var}(A) - (A/B)^2\, \mathrm{Var}(B)}{B^2}.
$$

Subtracting the cross term makes this the *tighter* (smaller) of the two estimates
when it is valid, which is the whole reason to use it: it does not double-count
the shared sampling uncertainty the subset relationship guarantees. But the
subtraction is also its fragility — the bracketed expression can go **negative**
when the denominator's scaled variance $(A/B)^2\,\mathrm{Var}(B)$ exceeds the
numerator's $\mathrm{Var}(A)$, in which case its square root is undefined and the
formula returns no admissible variance. That edge case is handled in the next
section, never silently.

### Family C2 — general ratio (conservative)

When the subset relationship is *not* guaranteed — an arbitrary ratio of two
counts, or a derived rate where conservative reproducibility is preferred — the
correct general form treats the numerator and denominator as approximately
independent and *adds* the cross term:

$$
\widehat{\mathrm{Var}}(A/B)
\approx \frac{\mathrm{Var}(A) + (A/B)^2\, \mathrm{Var}(B)}{B^2}.
$$

Note the **+** sign. Two consequences follow from that single change. First, this
variance is **always non-negative**: a sum of two squared, non-negative terms can
never go below zero, so Family C2 always yields an admissible margin. Second, it
is the **conservative** estimate — by adding rather than subtracting the cross
term, it never understates the uncertainty, at the cost of being slightly wide
when the numerator really is a subset. This is exactly the trade the package makes
for its curated derived rates: **Family C2 (`general_ratio_conservative`) is the
default**, and the tighter Family C1 (`proportion_subset`) is applied only when a
true subset relationship is asserted for a rate eligible for it. Among the five
curated rates, the poverty rate and labor-force participation are the
subset-eligible ones; the SNAP, SSI, and unemployment rates route to C2 by
default. The full per-rate policy is derived in
`vignette("theory-derived-rates", package = "catchmentACS")`.

The sign is the entire distinction. A single character — the minus in C1, the
plus in C2 — separates the tighter, subset-only formula from the conservative,
always-defined one. Whether C1 is *allowed* to break down gracefully is the
subject of the fallback.

## The C1 → C2 fallback and the status columns

The two ratio formulas are not independent choices made once at dispatch — they
are connected by a guard. When a row is dispatched to Family C1 but its subset
variance comes out **negative**, the package does not return a missing value, a
silent C1 number, or a `NaN` a caller might miss. It **recomputes the margin with
the always-non-negative Family C2 formula** so a defensible value is still
produced, and — this is the design commitment — it **records the substitution per
row** rather than hiding it. The fallback is never silent.

`cacs_propagate_moe()` returns, for every row, both the formula the dispatch
*asked for* and the formula it *actually used*:

| Column | Meaning |
|---|---|
| `moe_formula_requested` | the formula the dispatch resolved for this row (here, `proportion_subset`) |
| `moe_formula_effective` | the formula whose value was actually written (after any fallback) |
| `moe_fallback` | `TRUE` when the effective formula overrode the request |
| `moe_fallback_reason` | the cause: `negative_variance`, `zero_denominator`, `missing_moe`, or `n/a` |

For the C1 → C2 case specifically, a row whose subset variance went negative is
recorded as

- `moe_formula_requested = "proportion_subset"` — the request is preserved,
- `moe_formula_effective = "general_ratio_conservative"` — C2 produced the value,
- `moe_fallback = TRUE`,
- `moe_fallback_reason = "negative_variance"`.

A clean C1 row that stayed positive keeps `moe_formula_effective =
"proportion_subset"`, `moe_fallback = FALSE`, and `moe_fallback_reason = "n/a"`,
and its requested and effective formulas agree. At the end of a run a **single
summary** reports the total counts by reason, and a run-level provenance attribute
records how many rows took each path, so the fallback is auditable in aggregate as
well as per row. The point of keeping both `requested` and `effective` is that the
row reads as a self-documenting audit record: *this row asked for the subset
formula, its variance went negative, so the conservative formula produced the
value.* An auditor reading the output five years later can reconstruct exactly
which rows took which path without rerunning anything; no information is lost and
nothing is repaired behind the analyst's back.

The same `moe_fallback`/`moe_fallback_reason` mechanism also records the two
*degenerate* outcomes that are not formula substitutions but value losses, covered
next: a zero denominator and a missing input margin both yield `NA` with their own
reason strings.

## Numerical guards

Five guards protect the formulas from the inputs real ACS data presents. They are
implemented once, in the package's MOE helpers, so every family inherits the same
behavior, and each leaves a visible trace rather than a silent number.

- **Denominator near zero.** Families B and C divide by a weight sum or a
  denominator estimate. When that divisor is within a small epsilon of zero (the
  package uses the square root of machine epsilon, about $1.5 \times 10^{-8}$),
  the result is set to `NA` rather than allowed to explode. For a ratio row this
  is recorded as `moe_fallback_reason = "zero_denominator"` — the structural
  reason an empty or near-empty catchment does not masquerade as a precise rate.
- **Negative variance in a proportion.** The Family C1 subtraction can go below
  zero; this is the trigger for the C1 → C2 fallback of the previous section,
  recorded as `moe_fallback_reason = "negative_variance"`.
- **Missing-margin propagation.** A missing component margin propagates to a
  missing result rather than being silently treated as zero, preserving ACS
  suppression as a signal. For a ratio row a missing numerator or denominator
  margin yields `NA` with `moe_fallback_reason = "missing_moe"`.
- **Non-negative C2 variance.** Because C2 adds rather than subtracts, its bracket
  is provably $\ge 0$; it has no negative-variance branch and always yields a real
  margin. This is what makes C2 a safe fallback target for C1.
- **The multiplier is a fixed constant.** At the default 90% level $z = 1.645$ is
  the Bureau's rounded value, not the exact normal quantile; this is a deliberate
  choice for byte-level reproducibility against ACS-published margins, and the
  roughly $5 \times 10^{-5}$ gap to $\Phi^{-1}(0.95)$ is absorbed within the
  package's regression tolerance. Other confidence levels resolve $z$ at full
  precision.

The common thread is that none of these guards repairs silently. A zero
denominator, a negative subset variance, and a missing margin each leave a typed
trace in `moe_fallback_reason` and feed the end-of-run summary, so a degenerate
input is always distinguishable in the output from a clean computation.

## Numerical verification: the formulas match the package

This is the section that earns a methodologist's trust. We do two checks against
the *real* `cacs_propagate_moe()`, both offline on bundled fixtures, on a single
catchment. First we reproduce a **Family A weighted root-sum-square** entirely by
hand and assert it equals the package's reported margin to machine precision.
Then, on the *same* catchment, we deliberately construct a derived-rate case whose
**C1 subset variance is negative** and watch the package fall back to C2 —
recording `moe_fallback = TRUE` and `moe_fallback_reason = "negative_variance"` —
and we reproduce the C2 value by hand as well. If the formulas on this page were
not the formulas the code runs, either check would fail.

We build the catchment from the fixtures with `cacs_intersect_weight()`, exactly
the shapes pipeline stages 1–2 would have produced. We pick a catchment whose
coverage is *partial and uneven* so the weighted root-sum-square is genuinely
exercised: `AL_SITE_17` at a 10-minute drive intersects three tracts — one fully
enclosed and two clipped to about a ninth of their area.

```{r}
#| label: vf-intersect
#| eval: true

# These fixtures are exactly the shapes stages 1-2 would have returned, frozen.
iso <- readRDS(system.file(
  "extdata", "legacy_2025_isochrones.rds", package = "catchmentACS"
))
acs <- readRDS(system.file(
  "extdata", "sample_alabama_subset.rds", package = "catchmentACS"
))

site_id    <- "AL_SITE_17"
drive_time <- 10L

iso_one <- iso[
  iso$site_id == site_id & iso$drive_time_min == drive_time, ,
  drop = FALSE
]

# keep_tract_audit exposes the per-tract coverage weights we recompute by hand.
weighted <- cacs_intersect_weight(
  iso_sf           = iso_one,
  acs_sf           = acs,
  weight_method    = "area",
  keep_tract_audit = TRUE,
  verbose          = FALSE
)

propagated <- cacs_propagate_moe(weighted, verbose = FALSE)
```

### Step 1 — reproduce a Family A weighted RSS by hand

*Persons in poverty* (`B17001_002`) is a `spatial_total`, so its margin comes
from Family A. The per-tract audit gives the coverage weight $w^{cov}_{sj}$
(`area_wt`); we join each tract's **published** margin $M_j$ from the ACS fixture
and assemble the inputs:

```{r}
#| label: vf-A-inputs
#| eval: true

VAR <- "B17001_002"

audit <- attr(weighted, "cacs_tract_audit") |>
  transmute(GEOID, w_cov = area_wt)

published <- acs |>
  sf::st_drop_geometry() |>
  filter(GEOID %in% audit$GEOID, variable == VAR) |>
  select(GEOID, M_j = moe)

hand_A <- audit |>
  left_join(published, by = "GEOID") |>
  arrange(desc(w_cov))

hand_A
```

Now the Family A formula, term by term: scale each published margin by its
coverage weight, divide by $z$, square, sum, square-root, and multiply back by
$z$ — $M = z\sqrt{\sum_j (w^{cov}_{sj} M_j / z)^2}$:

```{r}
#| label: vf-A-compute
#| eval: true

z <- 1.645

per_tract_var <- (hand_A$w_cov * hand_A$M_j / z)^2  # (w * M / z)^2 per tract
hand_moe_A    <- z * sqrt(sum(per_tract_var))       # z * sqrt(sum(...))

list(
  per_tract_var = round(per_tract_var, 4),
  summed_var    = sum(per_tract_var),
  hand_moe_A    = hand_moe_A
)
```

The fully-enclosed tract dominates the combined variance; the two clipped tracts
enter with small coverage weights (about a ninth) and contribute little even
though their *raw* published margins are not small — exactly what the weighted
form predicts, since the weight, not the raw margin, governs the contribution.
Now the package's own row for the same variable:

```{r}
#| label: vf-A-compare
#| eval: true

pkg_A <- propagated |>
  filter(variable == VAR) |>
  select(variable, estimate, moe, moe_formula_effective, moe_fallback)

pkg_A

c(
  moe_match    = isTRUE(all.equal(hand_moe_A, pkg_A$moe)),
  moe_abs_diff = abs(hand_moe_A - pkg_A$moe)
)
```

The package routed this count through `moe_formula_effective = "weighted_sum"`
(Family A) with no fallback, and its reported `moe` — about `202.2` — **equals**
the hand-computed weighted root-sum-square exactly, with an absolute difference
of zero. The margin on this page and the margin in the code are the same
arithmetic.

### Step 2 — trigger the C1 → C2 fallback

Now the fallback. The poverty rate's numerator `B17001_002` is a true subset of
its denominator `B17001_001`, so it is C1-eligible; we will request the
`proportion_subset` formula. To make the subset variance go **negative** in a
fully controlled way, we take the real catchment and lower the numerator's
accumulated variance well below the denominator's scaled variance
$(A/B)^2\,\mathrm{Var}(B)$ — the configuration that drives C1's bracket below
zero. The aggregation carriers (an attribute of the intersect output) hold each
component's coverage total and accumulated variance; we copy them, shrink the
numerator's variance, and attach a matching carrier and long row for the rate:

```{r}
#| label: vf-fb-build
#| eval: true

carr <- attr(weighted, "cacs_aggregation_carriers")

num_code <- "B17001_002"  # poverty_rate numerator (subset)
den_code <- "B17001_001"  # poverty_rate denominator (universe)

A    <- carr$est_total[carr$variable == num_code]      # numerator total
B    <- carr$est_total[carr$variable == den_code]      # denominator total
VarB <- carr$var_total_raw[carr$variable == den_code]  # denominator variance
p    <- A / B                                          # the proportion A / B

# Force C1's bracket Var(A) - p^2 Var(B) below zero by setting the numerator's
# accumulated variance to a quarter of the denominator's scaled variance.
VarA_neg <- 0.25 * p^2 * VarB

carr_fb <- carr
carr_fb$var_total_raw[carr_fb$variable == num_code] <- VarA_neg

# A carrier row keyed on the rate name so the long rate row passes the
# carrier-key check; C1/C2 read the numerator/denominator ACS-code rows, not
# this placeholder, so its numeric fields are inert.
rate_carrier <- carr_fb[carr_fb$variable == num_code, , drop = FALSE][1, ]
rate_carrier$variable        <- "poverty_rate"
rate_carrier$estimand_family <- "derived_rate"
rate_carrier$est_total       <- NA_real_
rate_carrier$var_total_raw   <- 0
rate_carrier$est_mean        <- NA_real_
rate_carrier$var_mean_raw    <- 0
carr_fb <- bind_rows(carr_fb, rate_carrier)

# The long-format derived_rate row the formula will act on (built from a
# spatial_total template so all canonical columns and types are present).
rate_row <- weighted[weighted$variable == num_code, , drop = FALSE][1, ]
rate_row$variable              <- "poverty_rate"
rate_row$estimand_family       <- "derived_rate"
rate_row$estimate              <- NA_real_
rate_row$moe                   <- NA_real_
rate_row$weight_basis          <- "coverage"
rate_row$moe_formula_requested <- NA_character_
rate_row$moe_formula_effective <- NA_character_
rate_row$moe_fallback          <- NA
rate_row$moe_fallback_reason   <- "n/a"

weighted_fb <- bind_rows(weighted, rate_row)
attr(weighted_fb, "cacs_aggregation_carriers")  <- carr_fb
attr(weighted_fb, "cacs_schema_version")         <- "1.0"
attr(weighted_fb, "cacs_aggregation_provenance") <-
  attr(weighted, "cacs_aggregation_provenance")

# Confirm we have genuinely created a negative subset variance:
c(
  Var_A_set         = VarA_neg,
  scaled_Var_B      = p^2 * VarB,
  C1_inside_bracket = VarA_neg - p^2 * VarB   # < 0 by construction
)
```

The bracket $\mathrm{Var}(A) - (A/B)^2\,\mathrm{Var}(B)$ is negative, so C1 cannot
produce a real margin. We now request `proportion_subset` for this rate through
the locked `formula` argument and let the package decide:

```{r}
#| label: vf-fb-run
#| eval: true

propagated_fb <- cacs_propagate_moe(
  weighted_fb,
  formula = list(poverty_rate = "proportion_subset"),
  verbose = FALSE
)

fb_row <- propagated_fb |>
  filter(variable == "poverty_rate") |>
  select(variable, estimate, moe,
         moe_formula_requested, moe_formula_effective,
         moe_fallback, moe_fallback_reason)

fb_row
```

The status columns tell the whole story: `moe_formula_requested` is still
`"proportion_subset"` (the request is preserved), but `moe_formula_effective` is
`"general_ratio_conservative"`, `moe_fallback` is `TRUE`, and
`moe_fallback_reason` is `"negative_variance"`. The fallback fired, and it is
fully recorded — exactly the never-silent contract.

### Step 3 — reproduce the C2 fallback value by hand

The recorded margin should be the **C2** formula evaluated on the same inputs,
$M = z\sqrt{\mathrm{Var}(A) + (A/B)^2\,\mathrm{Var}(B)}\,/\,|B|$, and the estimate
should be the proportion $p = A/B$:

```{r}
#| label: vf-fb-verify
#| eval: true

hand_moe_C2 <- z * sqrt(VarA_neg + p^2 * VarB) / abs(B)

c(
  estimate_match = isTRUE(all.equal(p,           fb_row$estimate)),
  moe_match      = isTRUE(all.equal(hand_moe_C2, fb_row$moe)),
  moe_abs_diff   = abs(hand_moe_C2 - fb_row$moe)
)
```

Both match exactly. The package's fallback margin equals the hand-computed C2
root, and its estimate equals the proportion $A/B$. The run-level provenance
attribute also counts the event:

```{r}
#| label: vf-fb-provenance
#| eval: true

attr(propagated_fb, "cacs_moe_provenance")$n_rows_fallback_c1_to_c2
```

One row took the C1 → C2 path, matching the single fallback we constructed.

To summarize the check: a Family A weighted root-sum-square reproduced by hand
equals the package's reported `spatial_total` margin to the floating-point floor
on a partially-covered catchment; a deliberately negative C1 subset variance on
the same catchment triggers the package's never-silent C1 → C2 fallback, recorded
as `moe_formula_requested = "proportion_subset"`,
`moe_formula_effective = "general_ratio_conservative"`, `moe_fallback = TRUE`,
and `moe_fallback_reason = "negative_variance"`; and the resulting C2 margin and
the $A/B$ estimate both reproduce by hand exactly. The formulas in this article
and the arithmetic in the code are the same arithmetic, fallback included.

## Where this sits

This article derived and verified the uncertainty half of the estimator. Its
companions cover the spatial half and the rate catalogue, and the methodology
overview ties all three together; each reuses the notation fixed in the overview.

- **The estimator overview**, with the shared notation, the six-family table,
  and the four-family MOE summary this article expands:
  `vignette("methodology", package = "catchmentACS")`. This article is the
  derivation behind its margins-of-error section.
- **Spatial aggregation** — the coverage and normalized mean weights this article
  carries margins through, the per-site intersection algorithm, and the
  point-estimate forms $\widehat{Y}_s$, $\widehat{\bar X}_s$, $\widehat{R}_s$:
  `vignette("theory-spatial-aggregation", package = "catchmentACS")` and
  `?cacs_intersect_weight`.
- **Derived rates** — which of the curated rates are C1-eligible versus C2 by
  baseline policy, their ACS numerator/denominator pairs, and the carrier-lookup
  mechanics behind the ratio margins:
  `vignette("theory-derived-rates", package = "catchmentACS")` and
  `?cacs_derive_rates`.
- **Function reference** — the four families, the C1 → C2 fallback, the full
  status-column contract, the five numerical guards, and the `formula` and
  `level` arguments are documented in `?cacs_propagate_moe`; the SE ↔ MOE
  converters are `?cacs_se_to_moe` and `?cacs_moe_to_se`.

For the applied, run-it path that produces reporting tables with margins of error
from these same fixtures, see
`vignette("getting-started", package = "catchmentACS")` and
`vignette("alabama-tutorial", package = "catchmentACS")`.
