---
title: "Spatial aggregation: area-weighted estimation"
vignette: >
  %\VignetteIndexEntry{Spatial aggregation: area-weighted estimation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 2
    fig-width: 7
    fig-height: 5
    fig-align: center
    out-width: "85%"
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| eval: true

library(catchmentACS)
library(dplyr)
library(sf)

# Force network-free, deterministic builds: no provider/Census calls, no cache.
old_options <- options(
  catchmentACS.progress = "off",
  catchmentACS.rate_first_default = FALSE
)
old_cache <- getOption("catchmentACS.cache_enabled", NULL)
on.exit(options(old_options), add = TRUE)
on.exit(options(catchmentACS.cache_enabled = old_cache), add = TRUE)
cacs_set_cache(FALSE, scope = "session")
```

This article is a **method-track** companion to
`vignette("methodology", package = "catchmentACS")`. The overview frames the
whole pipeline as one estimator and *states* that a count and an average are
aggregated by two different weights; this article *derives* that result and then
does something the overview cannot — it **reproduces the package's arithmetic by
hand**, on a bundled fixture, so a methodologist can confirm that the formulas on
this page are the formulas the code runs. It is written for a reader who wants
the areal-interpolation model on the page: the uniform-within-tract assumption
stated precisely, the coverage weight and the normalized mean weight built from
first principles, the six estimand families laid out with the exact aggregation
formula and the weight each one uses, and a numerical check that earns trust in
the implementation. The reference page `?cacs_intersect_weight` is the
implementation anchor; this article supplies the mathematics behind its
"Estimand families and weights" section.

We reuse the notation fixed in the overview without redefining its motivation:
the isochrone $I_s$, the tract $T_j$, the intersection $I_s \cap T_j$, the
equal-area measure $\lvert\,\cdot\,\rvert$, a count estimate $Y_j$, a scalar
estimate $X_j$, and a rate's numerator and denominator counts $A_j, B_j$. The
two weights $w^{cov}_{sj}$ and $w^{mean}_{sj}$ and the three estimator forms
$\widehat{Y}_s$, $\widehat{\bar X}_s$, $\widehat{R}_s$ are restated where they
are derived. The margin-of-error algebra is **not** developed here — it is the
subject of `vignette("theory-moe-propagation", package = "catchmentACS")`.

::: {.callout-note}
## About the data in this article

The single numerical check below is built entirely from the package's bundled
`inst/extdata` fixtures — the synthetic `legacy_2025_isochrones`,
`legacy_2025_sites`, and `sample_alabama_subset` objects — so this article
renders with **no** Census, OSRM, ORS, or other network calls. These fixtures
preserve the production calling contract (sites, drive-time polygons,
tract-level ACS variables) but are **synthetic and illustrative**: they are
included only for examples, tests, and reproducible package checks. They are
**not** live Alabama policy estimates and must not be cited as real ACS
results. Any real analysis requires a fresh live run (see
`vignette("providers", package = "catchmentACS")`).
:::

## The areal-interpolation model

The estimand is a neighborhood characteristic of the population that lives inside
a drive-time catchment $I_s$: the count of people in poverty within a 10-minute
drive of a site, the per-capita income of that catchment, its SNAP rate. The ACS
publishes its estimates on census tracts $T_j$, whose boundaries were drawn for
enumeration, not for travel time. A drive-time area slices straight through tract
edges, capturing some tracts whole and others only in part — a sliver of one, half
of another, the whole of a third. Areal interpolation is the model that
re-expresses a tract-level published value as a catchment-level value across that
misalignment.

The model rests on a single assumption.

> **Uniform-within-tract assumption.** The characteristic of interest is spread
> *uniformly* over the area of each tract $T_j$. A catchment covering 40% of a
> tract's area contains 40% of whatever the tract counts.

Stated as a density, let $\rho_j$ be the quantity of the characteristic per unit
area in tract $T_j$. Uniformity means $\rho_j$ is constant across $T_j$, so the
amount of the characteristic contained in any sub-region $S \subseteq T_j$ is
proportional to that sub-region's area,

$$
\text{amount in } S = \rho_j \, \lvert S \rvert
= \big(\text{tract amount}\big)\cdot\frac{\lvert S \rvert}{\lvert T_j \rvert}.
$$

Taking $S = I_s \cap T_j$ gives the amount of the characteristic the catchment
captures from tract $j$: the tract's published amount, scaled by the fraction of
the tract's area inside the catchment. That fraction is the coverage weight of
the next section, and the whole estimator is built from this one identity.

The assumption's reach depends on the *type* of variable, and the distinction is
the hinge of everything that follows.

- An **extensive** (count) variable, $Y_j$ — persons, households, persons in
  poverty — is an *amount* that lives in the tract. Under uniformity it splits in
  proportion to area: half the tract's area carries half its people. The model is
  *exact for area* and *approximate for the count*, the approximation being only
  as good as the uniformity of the underlying density.
- An **intensive** (scalar) variable, $X_j$ — a median, a per-capita figure, a
  rate expressed as a level — is a *per-unit* quantity, not an amount. It does not
  accumulate over area: half a tract does not hold "half a dollar of per-capita
  income." Splitting an intensive value by area is meaningless. What the model can
  do for an intensive variable is treat its tract value as the local level of a
  field and form an area-weighted *average* of the tract values the catchment
  overlaps — a blend, not a sum.

This is exactly why one weight cannot serve both. **Extensive variables are
summed with a weight that allocates a fraction of each tract amount; intensive
variables are averaged with a weight that is a share and sums to one.** The
package encodes this split as a variable's *estimand family*; the two weights are
the two ways the same intersection area enters an aggregate.

**When the assumption fails.** It is most strained where population is
concentrated on a small part of a large tract. Suppose 80% of a tract's *area*
is uninhabited woodland and its residents are packed into the remaining 20%. A
catchment that clips only the woodland edge gets a coverage weight near $0.8$,
and area weighting then credits the catchment with ~80% of the tract's
population — when the truth is closer to 0%. The error is largest where land use
is most heterogeneous within a tract: the urban–rural fringe, coastlines, large
rural tracts straddling a town and its hinterland. The reserved
`weight_method = "population"` mode is designed to relax exactly this assumption
by substituting an ancillary population surface for area as the allocation basis
(a *dasymetric* refinement); in the current release area is the only basis, and
the assumption is stated plainly rather than hidden. Everything below concerns the
implemented area path, whose honest caveat is this paragraph.

## Coverage weight versus normalized mean weight

Both weights are built from the same ingredient, the intersection area
$\lvert I_s \cap T_j \rvert$. They differ only in what they divide it by, and that
single difference is the difference between a total and an average.

**The coverage weight.** From the areal-interpolation identity, the share of
tract $j$'s amount that the catchment captures is the fraction of the tract's area
inside the catchment,

$$
w^{cov}_{sj} = \frac{\lvert I_s \cap T_j \rvert}{\lvert T_j \rvert}
\;\in\; [0, 1].
$$

It is bounded by one tract at a time: $w^{cov}_{sj} = 1$ when $T_j \subseteq I_s$
(the tract is wholly inside the catchment) and $w^{cov}_{sj} = 0$ when the two are
disjoint. Crucially, the coverage weights of the tracts a catchment touches do
**not** sum to one. Summing over $j$,

$$
\sum_j w^{cov}_{sj}
= \sum_j \frac{\lvert I_s \cap T_j \rvert}{\lvert T_j \rvert}
\;\neq\; 1
$$

in general, because each term divides by a *different* tract area $\lvert T_j
\rvert$. Its value is interpretable: it is the *effective number of
tract-equivalents* the catchment covers. A catchment that wholly contained two
tracts and half of a third would have $\sum_j w^{cov}_{sj} = 2.5$; three tracts
each half-covered would sum to about $1.5$. Because the catchment area
$\lvert I_s \rvert$ equals $\sum_j \lvert I_s \cap T_j \rvert$ when $I_s$ lies
entirely within the tract layer, the coverage sum is the catchment's area
re-expressed in units of "average tract," not a probability that must normalize.
The package surfaces this sum directly as the `weight_sum` column, and a value
near a small integer means the catchment captured roughly that many whole tracts.

**The normalized mean weight.** An intensive variable needs a *share*: each tract
should contribute to the catchment average in proportion to how much of the
catchment it accounts for, with the contributions summing to one. Dividing the
intersection area by the *total* intersection area over the catchment gives

$$
w^{mean}_{sj}
= \frac{\lvert I_s \cap T_j \rvert}{\sum_k \lvert I_s \cap T_k \rvert},
$$

and this is normalized by construction,

$$
\sum_j w^{mean}_{sj}
= \frac{\sum_j \lvert I_s \cap T_j \rvert}{\sum_k \lvert I_s \cap T_k \rvert}
= 1,
$$

because the numerator summed over $j$ is exactly the denominator. The sum-to-one
property is structural, not coincidental. The two weights are proportional within
a catchment — $w^{mean}_{sj} \propto w^{cov}_{sj}\, \lvert T_j \rvert$ — but they
are renormalized against different denominators, and that is what separates a
quantity that *scales with the catchment* from one that *stays on the per-unit
scale of the variable*.

**Why the families split this way.** Put the two estimators side by side.

- A **count** is a sum of amounts, so it is aggregated with the coverage weight:

  $$
  \widehat{Y}_s = \sum_j w^{cov}_{sj}\, Y_j .
  $$

  If the catchment captures more area it captures more people, and $\widehat{Y}_s$
  grows accordingly. Using a normalized weight here would force the answer to a
  per-tract average and destroy the dependence on catchment size — a silent error.

- A **scalar** is a per-unit level, so it is aggregated with the normalized mean
  weight as a convex combination of the tract values:

  $$
  \widehat{\bar X}_s = \sum_j w^{mean}_{sj}\, X_j ,
  \qquad \sum_j w^{mean}_{sj} = 1 .
  $$

  Because the weights sum to one and are non-negative, $\widehat{\bar X}_s$ lies
  between $\min_j X_j$ and $\max_j X_j$: a catchment average per-capita income
  cannot exceed every tract's per-capita income. Using the coverage weight here
  would let an intensive value accumulate past any tract's level — again a silent
  error, not a runtime failure.

- A **derived rate** is a ratio of two extensive quantities. Each of the
  numerator and denominator is aggregated as its own coverage-weighted total, and
  the ratio is formed last:

  $$
  \widehat{R}_s
  = \frac{\sum_j w^{cov}_{sj}\, A_j}{\sum_j w^{cov}_{sj}\, B_j} .
  $$

  The rate inherits the coverage path because both of its parts are counts; it is
  not a weighted average of tract rates, which would weight a tiny-denominator
  tract the same as a large one.

**Neither misuse raises a warning.** Apply the coverage weight to an intensive
variable and you sum quantities that should average, producing a dollar figure
inflated past any tract's value — an income no one earns. Apply the mean weight to
a count and you average quantities that should accumulate, producing a
per-tract-ish number disconnected from how many people the catchment actually
contains. The arithmetic runs; it just estimates the wrong thing. This is the
silent estimand error the family classifier exists to prevent, and the reason the
package refuses to apply one universal weight to every variable.

The numerical check at the end of this article exhibits both sums on real
fixture geometry: $\sum_j w^{cov}_{sj}$ coming out to a non-integer, and
$\sum_j w^{mean}_{sj}$ coming out to exactly one.

## The six estimand families

A variable's *estimand family* is the rule that selects its weight and its
aggregation formula. The package classifies every requested ACS variable into one
of six families up front, then dispatches the matching arithmetic, so applying the
wrong weight to a variable cannot happen quietly. The table gives the point
estimate and the weight for each; the margin-of-error column is deliberately
omitted, as the MOE formulas are derived in the companion article.

| Estimand family | Representative variable | Point-estimate aggregation | Weight |
|---|---|---|---|
| `spatial_total` | total population; persons in poverty (`B17001_002`) | $\widehat{Y}_s = \sum_j w^{cov}_{sj} Y_j$ | coverage $w^{cov}_{sj}$ |
| `derived_rate` | poverty rate; SNAP rate | $\widehat{R}_s = \dfrac{\sum_j w^{cov}_{sj} A_j}{\sum_j w^{cov}_{sj} B_j}$ | coverage, numerator and denominator separately |
| `area_weighted_scalar_proxy` | per-capita income (`B19301_001`) | $\widehat{\bar X}_s = \sum_j w^{mean}_{sj} X_j$ | normalized mean $w^{mean}_{sj}$ |
| `population_weighted_scalar_proxy` *(reserved)* | scalar proxy under population weighting | $\widehat{\bar X}_s = \sum_j w^{mean}_{sj} X_j$ | population-normalized mean |
| `median_proxy` | tract median household income (`B19013_001`) | weighted blend of tract medians, **not** the true catchment median | normalized mean $w^{mean}_{sj}$ |
| `metadata_only` | provenance fields | no numeric aggregation | none |

A few families need a word beyond the table.

**`spatial_total` and `derived_rate`** are the two coverage-weighted families.
The total is the direct application of $\widehat{Y}_s = \sum_j w^{cov}_{sj} Y_j$.
The rate preserves the coverage-weighted numerator total $\sum_j w^{cov}_{sj} A_j$
and denominator total $\sum_j w^{cov}_{sj} B_j$ as separate carriers and divides
them last, so it is a ratio of catchment totals rather than an average of tract
ratios.

**`area_weighted_scalar_proxy`** is the active intensive family in this release:
a convex combination of tract values under the normalized mean weight, on the same
per-unit scale as the inputs. Its point estimate is, by definition,
$\widehat{\bar X}_s = \sum_j w^{mean}_{sj}\, X_j$. The exact per-cell
normalization the package applies when a single requested variable is missing in
some intersecting tracts is the implementation's responsibility, documented in
`?cacs_intersect_weight`; the definitional blend above always lands between the
contributing tracts' minimum and maximum.

**`population_weighted_scalar_proxy`** is a *reserved* family. It would replace
the area-based mean weight with a population-based one, allocating by an ancillary
population surface instead of by area to relax the uniform-within-tract assumption
where population is concentrated. Conceptually this is dasymetric refinement: a
finer ancillary layer (block-group population) is used to redistribute the coarser
tract quantity, so the allocation follows people rather than land. In the current
release this family is inert — `weight_method = "population"` stops with a clear
error at the orchestrator boundary — and the population weight is not yet
propagated through to estimates.

**`median_proxy`** carries an explicit caveat. A median is not additive and not
averageable: the median of a pooled population is generally *not* any weighted
average of the sub-populations' medians, which would require the underlying income
distribution the ACS does not publish at the tract level. The package reports a
mean-weighted blend of the tract medians under $w^{mean}_{sj}$ as a **proxy**, a
convenient summary on the right scale, and labels it as such throughout. It is
*not* the true median household income of the catchment population. The proxy is
honest about being a proxy.

**`metadata_only`** variables are provenance fields that carry through without any
numeric aggregation.

The classifier and the per-family dispatch are documented in the "Estimand
families and weights" section of `?cacs_intersect_weight`; this section is the
derivation behind that documentation.

## Why EPSG:5070

Both weights are ratios in which the intersection area appears. A ratio of areas
is only meaningful if numerator and denominator are measured in the *same* metric,
and only *unbiased* across the catchment if that metric preserves area
everywhere — so that "fraction of the tract" behaves like a fraction wherever the
tract falls on the map. That is precisely the property of an **equal-area**
projection, and it is why the choice of coordinate reference system is part of the
estimator's correctness, not a display detail.

On a geographic CRS (longitude/latitude), area is either dimensionless
(degrees-squared, which is not an area) or, with great-circle integration, an
ellipsoidal figure that is not equal-area. Web Mercator inflates area away from
the equator dramatically. None preserves the proportionality the weights rely on,
so the same physical hectare would be measured as more land in one part of the map
than another, and $w^{cov}_{sj}$ would silently over- or under-credit tracts by
where they sit, not by how much the catchment covers them. The package therefore
transforms validated inputs **once**, before measuring any area, to **EPSG:5070
(NAD83 / CONUS Albers Equal Area)**, an Albers equal-area projection whose
standard parallels are placed so that areal distortion across the contiguous
United States is held to a fraction of a percent — and across the Southeast,
between the standard parallels, it is effectively nil. With areas measured on
EPSG:5070, a square meter of intersection contributes the same weight in north and
south, $\lvert T_j \rvert$ and $\lvert I_s \cap T_j \rvert$ are evaluated in the
same $\text{m}^2$, and the coverage and mean weights mean what their formulas say.
The choice also follows entrenched domain practice: equal-area CONUS Albers is the
de facto standard for tract-level area work, which keeps these results comparable
to neighboring analyses.

This is also the reason the current release is scoped to the **contiguous United
States plus DC**: EPSG:5070 is the authoritative equal-area default there, where
Albers distortion is negligible. Regions outside that footprint — Alaska, Hawaii,
Puerto Rico — would need a different equal-area CRS and are deferred to a future
release. The single up-front transform happens before the per-site loop, and no
re-projection occurs inside it, so every area in a run is comparable on the same
equal-area metric.

## Computation

The mathematics above is per-tract, but a run involves many sites and many
tracts, and the implementation strategy is a deliberate choice rather than an
incidental one. The package evaluates the estimator with a **per-site loop**:
each site (and drive-time band) is processed in a closed body that subsets only
*that* isochrone, prefilters candidate tracts with a spatial index, cuts the exact
intersection geometry against only those candidates, measures intersection area
and divides by the pre-computed tract area to get $w^{cov}_{sj}$, builds both
weight columns, and aggregates by family — before moving to the next. The
alternative — one large vectorized spatial join over the full site-by-tract
cross-product — is rejected for two reasons that matter at scale.

- **Memory predictability.** A single join would hold every intersection polygon
  in memory at once, so peak memory grows with the total number of intersections
  across all sites. The per-site loop's working set is one catchment and its
  handful of overlapping tracts, so peak memory is independent of the number of
  sites and the largest study areas stay tractable.
- **Debugging and error isolation.** Spatial intersection is sensitive to invalid
  geometry, and in a single join one bad polygon can abort the whole batch with a
  stack trace that does not say which input was at fault. In the loop, a failure is
  attributable to a specific site, and a site whose geometry cannot be repaired is
  marked with a missing result while the rest of the batch proceeds.

Two robustness concerns are handled at the level of the individual site.

**Geometry repair.** Tract polygons and provider isochrones often arrive with
self-intersections, slivers, or holes that make an exact intersection throw a
topology error. The package applies a tiered repair before intersecting: a quick
validity check passes clean geometry through untouched; otherwise a standard
make-valid step fixes the large majority of cases; and a last-resort zero-width
buffer handles the remainder. Geometry that survives all three tiers still invalid
is treated as a failure for that site rather than silently producing a wrong area.

**Edge cases.** A handful of geometric situations are handled explicitly, each
isolated to the site where it occurs.

- *Empty intersection.* A catchment that overlaps no tract — outside ACS coverage,
  or a missing isochrone — yields a result row marked as covering zero tracts, so
  the caller can detect it rather than mistake it for a real estimate of zero.
- *Sliver fragments.* A tract clipped along a coastline or river can intersect in
  a fragment so thin that its weight is floating-point noise. A small `min_weight`
  threshold drops these negligible fragments before they enter an aggregate, so
  they cannot perturb a total.
- *Self-intersections.* These are the geometry-repair case above, resolved by the
  tiered repair and, failing that, by isolating the site.

The conceptual point is that the loop turns batch-level fragility into site-level
diagnosability, and the equal-area weights are computed identically inside it. The
full parameter contract, the default `min_weight`, the audit attribute used below,
and the per-row provenance live in `?cacs_intersect_weight`.

## Numerical verification: the formulas match the package

This is the section that earns a methodologist's trust. We take **one** site and
**one** ring from the fixtures, pull the per-tract audit the package exposes, and
**recompute the coverage weights and a spatial-total aggregation entirely by
hand** from the audit's two area columns. Then we show the by-hand figures equal
the package's reported figures to machine precision. If the formulas on this page
were not the formulas the code runs, this check would fail.

We deliberately pick a catchment whose coverage is *partial and uneven* — not a
trivial fully-enclosed case — so the area-fraction formula is genuinely exercised
and the coverage and mean weights are genuinely different. The synthetic site
`AL_SITE_17` at a 10-minute drive time intersects three tracts: one it fully
contains and two it clips at the edge.

First, intersect and keep the per-tract audit:

```{r}
#| label: vf-intersect
#| eval: true

# These fixtures are exactly the shapes pipeline stages 1-2 would have returned,
# frozen for offline reproducibility.
sites <- readRDS(system.file(
  "extdata", "legacy_2025_sites.rds", package = "catchmentACS"
))
iso <- readRDS(system.file(
  "extdata", "legacy_2025_isochrones.rds", package = "catchmentACS"
))
acs <- readRDS(system.file(
  "extdata", "sample_alabama_subset.rds", package = "catchmentACS"
))

site_id    <- "AL_SITE_17"
drive_time <- 10L

iso_one <- iso[
  iso$site_id == site_id & iso$drive_time_min == drive_time, ,
  drop = FALSE
]

weighted <- cacs_intersect_weight(
  iso_sf           = iso_one,
  acs_sf           = acs,
  weight_method    = "area",
  keep_tract_audit = TRUE,
  verbose          = FALSE
)
```

The audit attribute holds, for each tract the catchment touched, the
intersection area `int_area_m2` and the full tract area `tract_area_m2` — the two
ingredients of $w^{cov}_{sj}$ — both measured on EPSG:5070:

```{r}
#| label: vf-audit
#| eval: true

audit <- attr(weighted, "cacs_tract_audit") |>
  select(GEOID, int_area_m2, tract_area_m2) |>
  arrange(desc(int_area_m2))

audit
```

### Step 1 — recompute both weights by hand

From those two columns we form the coverage weight (each tract's own denominator)
and the normalized mean weight (the common catchment denominator), straight from
the formulas above:

```{r}
#| label: vf-weights
#| eval: true

hand <- audit |>
  mutate(
    w_cov  = int_area_m2 / tract_area_m2,        # |I_s n T_j| / |T_j|
    w_mean = int_area_m2 / sum(int_area_m2)      # |I_s n T_j| / sum_k |I_s n T_k|
  )

hand |> select(GEOID, w_cov, w_mean)

c(
  sum_w_cov  = sum(hand$w_cov),    # NOT 1: each term has its own denominator
  sum_w_mean = sum(hand$w_mean)    # exactly 1 by construction
)
```

The three coverage weights are `1.000`, `0.1125`, and `0.1125`: one tract fully
enclosed, two clipped to about a ninth of their area. Their sum, about `1.225`,
is the effective number of whole tracts the catchment covers — emphatically not
one, exactly as the derivation said. The mean weights, by contrast, sum to
exactly `1`. That structural contrast is the entire difference between a *total*
and an *average*, here in fixture numbers.

### Step 2 — recompute a spatial total by hand

Now a count. *Persons in poverty* (`B17001_002`) is a `spatial_total`, so the
package reports $\widehat{Y}_s = \sum_j w^{cov}_{sj}\, Y_j$. We pull the three
tracts' published counts and form that sum by hand:

```{r}
#| label: vf-total
#| eval: true

counts <- acs |>
  sf::st_drop_geometry() |>
  filter(GEOID %in% hand$GEOID, variable == "B17001_002") |>
  select(GEOID, Y = estimate)

hand_total <- hand |>
  select(GEOID, w_cov) |>
  left_join(counts, by = "GEOID")

hand_total

hand_Y_hat       <- sum(hand_total$w_cov * hand_total$Y)   # sum_j w_cov * Y_j
hand_weight_sum  <- sum(hand_total$w_cov)                  # sum_j w_cov

c(hand_Y_hat = hand_Y_hat, hand_weight_sum = hand_weight_sum)
```

### Step 3 — compare against the package

The package's own row for the same variable reports the estimate and the weight
sum it computed internally:

```{r}
#| label: vf-pkg-row
#| eval: true

pkg_row <- weighted |>
  filter(variable == "B17001_002") |>
  select(variable, estimate, weight_sum, n_tracts,
         estimand_family, weight_basis)

pkg_row
```

`estimand_family` is `spatial_total` and `weight_basis` is `coverage`,
confirming the package routed this count through the coverage path. Now the
equality check — by-hand against package, on both the estimate and the weight
sum:

```{r}
#| label: vf-equality
#| eval: true

c(
  estimate_match      = isTRUE(all.equal(hand_Y_hat,      pkg_row$estimate)),
  weight_sum_match    = isTRUE(all.equal(hand_weight_sum, pkg_row$weight_sum)),
  estimate_abs_diff   = abs(hand_Y_hat      - pkg_row$estimate),
  weight_sum_abs_diff = abs(hand_weight_sum - pkg_row$weight_sum)
)
```

Both match. The hand-computed spatial total — about `1237` persons in poverty —
equals the package's `estimate`, and the hand-computed $\sum_j w^{cov}_{sj}$
equals the package's `weight_sum`, with absolute differences at the
floating-point floor (on the order of $10^{-12}$ and $10^{-15}$). The
`weight_sum` column the package reports *is* the not-equal-to-one coverage sum
from Step 1, and the `estimate` *is* the coverage-weighted count we built from
the audit. The formulas in this article and the arithmetic in the code are the
same arithmetic.

The mean weight's defining property is equally direct to confirm from the same
audit:

```{r}
#| label: vf-mean-sum
#| eval: true

# w_mean sums to exactly one within this (site, drive_time) cell.
isTRUE(all.equal(sum(hand$w_mean), 1))
```

### The intensive path, by definition

A `spatial_total` count and an `area_weighted_scalar_proxy` scalar reuse this
*identical* intersection geometry; only the denominator of the weight differs —
$\lvert T_j \rvert$ for the total just verified, $\sum_k \lvert I_s \cap T_k
\rvert$ for the mean. By definition the scalar proxy is the mean-weighted blend

$$
\widehat{\bar X}_s = \sum_j w^{mean}_{sj}\, X_j ,
$$

a convex combination that lands strictly between $\min_j X_j$ and $\max_j X_j$.
Because the mean weights just summed to one, this is the structural guarantee
that a catchment "average per-capita income" cannot exceed any tract's
per-capita income — the exact opposite of what summing an intensive level with
the coverage weight would do, which inflates the figure past every tract's value
into a number no one earns. That is the silent estimand error the family split
exists to enforce, and it is why the package routes `B19301_001` through the
mean path and `B17001_002` through the coverage path automatically.

The package's reported proxy `estimate` for a scalar is on this per-unit scale by
construction. Its exact per-cell normalization — in particular how it handles a
requested variable that is missing in some intersecting tracts — is a bookkeeping
detail that is the package's responsibility, documented in the "Estimand families
and weights" section of `?cacs_intersect_weight`. We therefore frame the proxy
here at the level of its definition rather than printing a specific proxy estimate
number, so this self-contained check stays anchored on the **coverage-weighted
`spatial_total`** path it verifies exactly, and on the
$\sum_j w^{mean}_{sj} = 1$ identity, both confirmed above to machine precision.

To summarize the check: the coverage weight, its non-unit sum, the package's
`weight_sum`, and the coverage-weighted `spatial_total` all coincide with the
hand computation to the floating-point floor; the mean weights sum to exactly
one; and the scalar proxy is the definitional mean-weighted blend on the
per-unit scale, between each tract's minimum and maximum.

## Where this sits

This article derived and verified the spatial half of the estimator. Its two
companions cover the uncertainty half, and the methodology overview ties all
three together; each reuses the notation fixed in the overview.

- **The estimator overview**, with the shared notation, the six-family table,
  and the count-vs-average worked example:
  `vignette("methodology", package = "catchmentACS")`. This article is the
  derivation behind its weight and family sections.
- **MOE propagation** — the published tract MOEs are carried through these same
  weights by `cacs_propagate_moe()`; the four formula families, the
  $M = 1.645\,SE$ conversion, the ratio-variance forms, and the independence
  assumptions are derived in
  `vignette("theory-moe-propagation", package = "catchmentACS")` and
  `?cacs_propagate_moe`.
- **Derived rates** — the coverage-weighted numerator and denominator totals
  built here are divided into the curated catchment rates in
  `vignette("theory-derived-rates", package = "catchmentACS")` and
  `?cacs_derive_rates`.
- **Function reference** — the locked signature, the per-site loop, the edge
  cases, the tract audit attribute used above, the `min_weight` sliver filter,
  the per-row provenance, and the "Estimand families and weights" section:
  `?cacs_intersect_weight`.

For the applied, run-it path that produces reporting tables from these same
fixtures, see `vignette("getting-started", package = "catchmentACS")` and
`vignette("alabama-tutorial", package = "catchmentACS")`.
