---
title: "Visual walkthrough: one Birmingham catchment, four stages"
vignette: >
  %\VignetteIndexEntry{Visual walkthrough}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 2
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| eval: true

library(catchmentACS)
library(sf)

fixture <- readRDS(system.file("extdata", "visual_walkthrough_fixture.rds",
                                package = "catchmentACS"))
ANCHOR <- fixture$anchor_site
ANCHOR
```

## Why a visual walkthrough

`cacs_run()` hands you a tidy table: one row per site per rate, with an
estimate and a margin of error. That table tells you *what* a site's
neighborhood looks like, but it hides *how* the package got there — the
drive-time polygon it drew, the Census tracts it clipped, and the weights it
used to blend them. When a reviewer or co-author asks "how exactly did you
turn one address into a poverty rate?", a table is a hard thing to point at.

This article answers that question with maps. It opens up that one line of
code and follows a single, real site — the **Vulcan statue in Birmingham,
Alabama** (site `AL_BHM_01`, `Birmingham_Vulcan`, on Red Mountain just south
of downtown, the same site exercised in the package's verified v0.3.0
workflow run) — through the four spatial stages that turn one point on a map
into neighborhood statistics:

1. **Isochrone** — draw the area a family can reach from the site in ten
   minutes of driving.
2. **Intersection** — find every Census tract that area touches, and how much
   of each one falls inside it.
3. **Weighted variable** — pool one ACS variable across those tracts,
   weighting each tract by its overlap.
4. **Rates** — collapse the whole catchment into five headline rates, each
   with a margin of error.

Each stage is drawn as one interactive `leaflet` map, so you can *watch* the
computation happen for a place you can picture. Where
`vignette("getting-started", package = "catchmentACS")` shows the ergonomics
of the one-line call, this article traces the geometry inside it. Every map is
paired with a short **"what to look for"** note that names the exact spatial
feature to read off the picture. We stay at the map level here; the math
behind area weighting and MOE propagation lives in `?cacs_propagate_moe` and
`vignette("methodology", package = "catchmentACS")`.

::: {.callout-note}
## The maps are a frozen real-data snapshot

Every map below is built from one small bundled file,
`visual_walkthrough_fixture.rds`, so this article renders fully **offline** —
no Census, OSRM, ORS, or other network call is made. The snapshot holds *real*
data: a real ten-minute OSRM drive-time isochrone around the Vulcan, plus the
real surrounding Birmingham-area Census tracts and American Community Survey
(ACS) estimates it overlaps, captured once and frozen so the figures reproduce
exactly. It is one recognizable catchment, not a synthetic toy — but it is
also a single moment in time, a teaching snapshot rather than a live policy
estimate. A real analysis of your own sites needs a fresh live run against a
routing provider and the Census API (see
`vignette("providers", package = "catchmentACS")`). To reuse these maps on
your own data, hand a `cacs_run()` result and its inputs to the same five
plotting functions in place of `fixture`.
:::

## Stage 1 — the isochrone

The first stage answers the most basic question: *what, spatially, is this
site's catchment?* An **isochrone** is the drive-time catchment polygon: the
area you can reach from the site within a set number of minutes. For the Vulcan
we use a single **ten-minute** drive-time ring, so the map reads as one
connected reachable area rather than a set of nested bands. The red marker is
the site; click it to read its `site_id`, name, and coordinates. The shaded
polygon is everywhere a car can get to in ten minutes on the road network. (In
a live run, OSRM is the default routing provider, ORS is available with an API
key, and Mapbox and r5r are reserved fail-loud names — see
`vignette("providers", package = "catchmentACS")`.)

```{r}
#| label: stage1
#| eval: !expr requireNamespace("leaflet", quietly = TRUE)

cacs_plot_site_isochrone(
  site_id  = ANCHOR,
  iso_sf   = fixture$iso_sf,
  sites_df = fixture$sites_df
)
```

**What to look for.** Read the *shape and reach* of the ten-minute polygon.
Notice that it is not a tidy circle: it stretches along the interstates and
arterials that radiate from Red Mountain and pinches in where the terrain or
the street grid slows a driver down. That asymmetry is the whole point of a
drive-time catchment — it follows the road network, not a compass, so it
captures who can actually get to the site rather than who happens to live
within a fixed radius. The polygon's footprint is the geographic area this site
is, in effect, responsible for, and it is what the next three stages fill in
with population data.

## Stage 2 — intersection and area weight

Stage 2 answers *how much of each Census tract falls inside the catchment?* The
ten-minute polygon is overlaid on every tract it touches and intersected with
them in an equal-area projection (EPSG:5070). Each tract is then shaded by its
**area weight**,

$$
\text{area\_wt}_j \;=\;
\frac{\text{area}\!\left(\text{catchment} \cap \text{tract}_j\right)}
     {\text{area}\!\left(\text{tract}_j\right)},
$$

the fraction of that tract's land that sits inside the catchment, running from
`0` (no overlap) to `1` (the tract is entirely inside). The shading uses a
viridis scale locked to that fixed `[0, 1]` domain, so a given weight always
reads as the same color. This weight is exactly the lever the next stage pulls
when it blends tracts; its derivation lives in
`vignette("methodology", package = "catchmentACS")`.

```{r}
#| label: stage2
#| eval: !expr requireNamespace("leaflet", quietly = TRUE)

cacs_plot_site_intersection(
  site_id  = ANCHOR,
  iso_sf   = fixture$iso_sf,
  tract_sf = fixture$tract_sf,
  sites_df = fixture$sites_df
)
```

**What to look for.** Dark tracts carry a high area weight: most or all of the
tract lies inside the catchment, so it pulls hard on every downstream number.
The interior tracts near the Vulcan should read dark — close to `1.0`, fully
enclosed — while the **pale tracts trace the catchment boundary**: those are
*slivers*, tracts the ten-minute polygon clips only at the edge, contributing
just a fraction of their residents. Hover or click any tract to read its
`GEOID` and exact `area_wt` in the popup. The contrast between the dark core
and the pale rim is the visual fingerprint of area weighting, and it is what
makes catchment estimates more honest than a crude "is the tract's centroid
inside?" rule: a tract that is 18% inside is counted as 18% inside, not
all-or-nothing.

## Stage 3 — a weighted ACS variable

Stage 3 answers *how is one ACS indicator distributed across the catchment?*
Here we map `B17001_002`, the count of people with income below the poverty
level. Because it is a count, `cacs_intersect_weight()` aggregates it as an
area-weighted **sum**: a tract that is half inside the catchment contributes
half of its poverty count. Each tract is drawn as a choropleth of its raw ACS
`estimate`, with the ten-minute outline laid on top for reference. Each tract
popup reports its `GEOID`, the variable code, the `estimate`, the margin of
error (MOE — the half-width of the 90% confidence band the Census publishes
with every ACS figure), the tract's `area_wt`, and `area_wt × estimate`, which
is that tract's contribution to the site total.

```{r}
#| label: stage3
#| eval: !expr requireNamespace("leaflet", quietly = TRUE)

cacs_plot_site_weighted(
  site_id  = ANCHOR,
  iso_sf   = fixture$iso_sf,
  tract_sf = fixture$tract_sf,
  acs_sf   = fixture$acs_sf,
  variable = "B17001_002",
  sites_df = fixture$sites_df
)
```

**What to look for.** Read the *spread* of color across the tracts inside the
outline. An even shade would mean poverty is roughly uniform across the
catchment; a sharp dark-to-light gradient flags internal heterogeneity that a
single catchment average would paper over. Cross-check the deepest-shaded
tracts against Stage 2: a tract that is dark in *both* maps — a high poverty
count **and** a high area weight — is the one driving the catchment poverty
total hardest. That `area_wt × estimate` line in each popup is exactly the
quantity Stage 4 sums up.

::: {.callout-tip}
## Mapping a different indicator

Swap the `variable` argument for any code in your ACS data, and set
`variable_family` to match so the palette and popup units stay correct. For
example, `variable = "B19013_001"` with `variable_family = "median_proxy"` for
median household income, or `variable = "B23025_005"` (the unemployed count)
with the default `variable_family = "spatial_total"` for a count. The family
controls the ColorBrewer palette and how the popup labels the value, so match
it to the kind of measure: `"spatial_total"` for counts,
`"area_weighted_scalar_proxy"` for scalar proxies, `"median_proxy"` for
medians.
:::

## Stage 4 — the headline rates

Stage 4 compresses the whole tract-level panel into one site and one popup.
Click the site marker to read the five curated rates for this Birmingham
catchment — `poverty_rate`, `snap_rate`, `ssi_rate`, `unemp_rate`, and
`labor_force_participation` — each as an `estimate ± 90% MOE` band, alongside
the count of contributing tracts for the numerator and denominator
(`n_tracts_num`, `n_tracts_den`). The exact figures live in the popup; read
them straight from the marker.

```{r}
#| label: stage4
#| eval: !expr requireNamespace("leaflet", quietly = TRUE)

cacs_plot_site_rates(
  site_id    = ANCHOR,
  iso_sf     = fixture$iso_sf,
  run_result = fixture$run_result,
  sites_df   = fixture$sites_df
)
```

**What to look for.** Read each rate together with the *width* of its `± MOE`
band, which tells you how precisely the estimate is pinned down — a tight band
is a confident number, a wide one is a number to quote carefully. Two flags in
the popup mark the limits of reliability. First, if a rate's MOE had to be
computed with the conservative *fallback* formula rather than the requested
one, its name is prefixed with an asterisk. Second, if an estimate could not be
formed at all — say a carrier variable failed to resolve on one side of the
ratio — the cell comes back `NA`. For this clean Birmingham snapshot all five
rates resolve normally, with no asterisks and no `NA`s, but those markers are
the first things to scan for on a real run. The full NA-propagation rule is in
the "Failure propagation" section of `?cacs_run`.

## All four in one call: `cacs_plot_site_pipeline()`

You rarely build the four maps by hand. `cacs_plot_site_pipeline()` runs all
four stage functions above with one shared set of site and styling arguments
and returns a single `cacs_site_plot_pipeline` object — a classed list of four
`leaflet` widgets. Printing it does **not** draw anything; its `print()` method
just lists the accessors, so you can render exactly the stage you want, at the
console or in a Quarto chunk.

```{r}
#| label: pipeline
#| eval: !expr requireNamespace("leaflet", quietly = TRUE)

maps <- cacs_plot_site_pipeline(
  site_id    = ANCHOR,
  iso_sf     = fixture$iso_sf,
  tract_sf   = fixture$tract_sf,
  acs_sf     = fixture$acs_sf,
  run_result = fixture$run_result,
  sites_df   = fixture$sites_df,
  variable   = "B17001_002"
)
print(maps)
```

```{r}
#| label: pipeline-display
#| eval: !expr requireNamespace("leaflet", quietly = TRUE)

# Display any single stage by its accessor. We show the rate-summary stage
# here because its one popup pulls together the entire upstream pipeline:
# isochrone + intersection + area weighting + MOE propagation.
maps$rates
```

All four accessors — `maps$isochrone`, `maps$intersection`, `maps$weighted`,
and `maps$rates` — are ordinary `leaflet` HTML widgets, so you can save any of
them to a standalone HTML report with `htmlwidgets::saveWidget()` or drop one
straight into a Shiny `leafletOutput()` slot. See `?cacs_plot_site_pipeline`
for the full argument list.

## Interpretation guide: which question each stage answers

The value of splitting the pipeline into four maps is that each one answers a
different question — exactly how you want to narrate it in a policy report,
because each map maps to a different reviewer concern.

- **Stage 1 (isochrone)** — *"How far does my site's catchment reach?"* The
  **accessibility** question. The shape and reach of the ten-minute polygon
  show a decision-maker the geographic area the site is responsible for, and
  whether the setting is dense-urban or sprawling.
- **Stage 2 (intersection / area weight)** — *"How do my catchment and the
  Census tracts overlap?"* The **methodological-transparency** question.
  Showing the weight itself lets a reviewer verify, tract by tract, how much
  each one feeds the downstream estimate, and see that edge slivers are
  down-weighted rather than counted whole — no hidden all-or-nothing
  assignment.
- **Stage 3 (weighted variable)** — *"How is one indicator distributed inside
  my catchment?"* The **spatial-heterogeneity** question. The tract choropleth
  exposes internal disparities that a single catchment average would mask, and
  points to the neighborhoods driving the total.
- **Stage 4 (rates summary)** — *"What are my site's five headline rates, in
  one line?"* The **final-reporting** question. Estimates ± MOE and
  contributing tract counts fit in one popup and transfer cleanly to a report
  table, while the fallback asterisk and any `NA` make the limits of each
  estimate visible at the same moment.

## Where to next

- **Your first run in ten minutes**, the whole pipeline in one offline call:
  `vignette("getting-started", package = "catchmentACS")`.
- **A full reporting workflow** on Alabama sites, from the raw pipeline to
  export-ready tables: `vignette("alabama-tutorial", package = "catchmentACS")`.
- **The math** behind area weighting and MOE propagation:
  `?cacs_propagate_moe` and `vignette("methodology", package = "catchmentACS")`.
- **The four-stage builder in detail**: `?cacs_plot_site_pipeline`.
- **Routing providers and credentials** — OSRM (default), ORS (optional, needs
  a key), and the reserved Mapbox/r5r surfaces, plus how to take a walkthrough
  like this one live: `vignette("providers", package = "catchmentACS")`.

Bug reports and feature requests:
<https://github.com/joonho112/catchmentACS/issues>.
