---
title: "Mapping each step for one Birmingham site"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    math_method: mathml
vignette: >
  %\VignetteIndexEntry{Mapping each step for one Birmingham site}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: knitr-options
#| include: false

knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

```{r}
#| label: setup
#| eval: true

library(catchmentACS)
```

```{r}
#| label: setup-cache
#| include: false

# Compute every result in this article instead of reading saved ones; the
# option is restored at the end of the article.
old_options <- options(catchmentACS.cache_enabled = FALSE)
```

catchmentACS has functions that map, for one site, how `cacs_run()` turns
American Community Survey (ACS) estimates for census tracts into estimates for
the site's drive-time area. This article draws their four maps for a site in
Birmingham, Alabama:

1. `cacs_plot_site_isochrone()`: the site and its drive-time area (isochrone),
   the area within a given driving time of the site.
2. `cacs_plot_site_intersection()`: the census tracts that overlap the
   drive-time area, shaded by the share of each tract's area that lies inside
   it.
3. `cacs_plot_site_weighted()`: the estimates of one ACS variable for those
   tracts.
4. `cacs_plot_site_rates()`: the five rates for the drive-time area, with their
   margins of error.

`cacs_plot_site_pipeline()` builds all four maps in one call. The maps are
interactive and need the leaflet package.
`vignette("getting-started", package = "catchmentACS")` shows how to run
`cacs_run()` and read its result.

The maps are drawn from `visual_walkthrough_fixture.rds`, a file installed with
the package that holds data for a real place. The site, `AL_BHM_01`, is a point
near the Vulcan statue on Red Mountain, south of downtown Birmingham. Its
10-minute drive-time area was computed on May 27, 2026, by the Open Source
Routing Machine (OSRM) with the default settings of `cacs_isochrone()`. The
requests went to the public OSRM demo server, and the osrm package drew the
area from travel times to a grid of 30 by 30 points around the site
(`res = 30`). The server routes on OpenStreetMap road data, © OpenStreetMap
contributors, available under the Open Database License
(<https://www.openstreetmap.org/copyright>); the date of the data it used is not
recorded. An area computed later, or with another `res`, can differ, and so can
the estimates computed from it.

The file also holds the census tracts within 2 km of the drive-time area: their
boundaries and their ACS 5-year estimates for 2019–2023, with margins of error,
for the variables in `cacs_acs_default_vars`, as downloaded by
`cacs_acs_prefetch()`. It also holds the result of `cacs_run()` for the site,
computed from these inputs. The map functions only draw such results; they do
not contact a routing service or the Census Bureau.

```{r}
#| label: read-data
#| eval: true

walkthrough <- readRDS(system.file("extdata", "visual_walkthrough_fixture.rds",
                                   package = "catchmentACS"))
names(walkthrough)
one_site <- walkthrough$anchor_site
one_site
```

`iso_sf` holds the drive-time area and `sites_df` the site's point. `tract_sf`
and `acs_sf` are the same table of tract boundaries and estimates, with one row
for each tract and variable; the map functions take the boundaries and the
estimates through separate arguments, which can be the same object.
`run_result` is the result of `cacs_run()`, and `anchor_site` gives the
`site_id` used below as `one_site`. For your own sites, the maps take your site
points and the results of `cacs_isochrone()`, `cacs_acs_prefetch()`, and
`cacs_run()` in place of these elements.

## Stage 1: the drive-time area

`cacs_plot_site_isochrone()` draws the drive-time areas of a site, in one color
for each drive time, and marks the site with a red point. Clicking the point
shows the site's `site_id`, name, and coordinates.

```{r}
#| label: stage1
#| eval: !expr requireNamespace("leaflet", quietly = TRUE)

cacs_plot_site_isochrone(
  site_id  = one_site,
  iso_sf   = walkthrough$iso_sf,
  sites_df = walkthrough$sites_df
)
```

The example has one drive-time area, for 10 minutes. The area is built from
travel times on the road network, and its edge lies much farther from the site
in some directions than in others.

## Stage 2: the tracts and their weights

`cacs_plot_site_intersection()` overlays the drive-time area on the census
tracts. It shades the part of each tract that lies inside the area by the
tract's coverage weight, labeled `area_wt` on the map,

$$
\text{area\_wt}_j \;=\;
\frac{\text{area}\!\left(\text{drive-time area} \cap \text{tract}_j\right)}
     {\text{area}\!\left(\text{tract}_j\right)},
$$

the share of the tract's area that lies inside the drive-time area. The weight
is 0 when the two do not overlap and 1 when the whole tract is inside. The
function measures both areas on the equal-area projection EPSG:5070, as
`cacs_intersect_weight()` does, and a tract's area includes any water inside
its boundary. The colors run from 0 to 1 on every map that this function
draws, so a weight has the same color on all of them.

```{r}
#| label: stage2
#| eval: !expr requireNamespace("leaflet", quietly = TRUE)

cacs_plot_site_intersection(
  site_id  = one_site,
  iso_sf   = walkthrough$iso_sf,
  tract_sf = walkthrough$tract_sf,
  sites_df = walkthrough$sites_df
)
```

The tracts that lie wholly inside the drive-time area have a weight of 1 and
are yellow. Along the edge of the area, the tracts cut by it have smaller
weights and darker colors, down to dark purple for tracts that lie almost
entirely outside. Hovering over a tract shows its `GEOID` and weight; clicking
it also shows `intersection_km2`, the area of its part inside the drive-time
area in square kilometers.

Counts, including the two counts of each rate, are added up with these weights.
A tract with 18 percent of its area inside the drive-time area adds 18 percent
of its count, such as its number of people below the poverty level, to the
count for the area. This assumes that whatever the count counts is spread
evenly over the tract's area: for the number of people below the poverty
level, that those people are spread evenly, and not only the population as a
whole;
`vignette("theory-spatial-aggregation", package = "catchmentACS")` discusses
this assumption and when it fails. Medians and per-person values are averaged
with other weights, the area shares, which
`vignette("methodology", package = "catchmentACS")` describes.

## Stage 3: one ACS variable

`cacs_plot_site_weighted()` shades the same parts of the tracts by each tract's
published estimate for one ACS variable, before any weighting, and draws the
outline of the drive-time area on top. The map below shows `B17001_002`, the
number of people whose income in the past 12 months was below the poverty
level. Clicking a tract shows the same values as the Stage 2 map and also the
variable code, the tract's estimate, its margin of error (MOE), and the
estimate multiplied by the coverage weight, labeled `area_wt x estimate`. The
MOE is the half-width of the 90 percent confidence interval published with the
estimate.

```{r}
#| label: stage3
#| eval: !expr requireNamespace("leaflet", quietly = TRUE)

cacs_plot_site_weighted(
  site_id  = one_site,
  iso_sf   = walkthrough$iso_sf,
  tract_sf = walkthrough$tract_sf,
  acs_sf   = walkthrough$acs_sf,
  variable = "B17001_002",
  sites_df = walkthrough$sites_df
)
```

Darker greens mark larger counts. A count describes a whole tract and grows
with the tract's population, so a dark tract need not have a high poverty rate.
For a count with no missing tract estimates, as here, the products in the
pop-ups add up to the estimate that `cacs_run()` gives for the drive-time area,
the numerator of the poverty rate in Stage 4. A tract adds much to that total
only when both its count and its weight are large. Here the two tracts with the
largest counts lie mostly outside the drive-time area, as their weights in the
Stage 2 map show. The map shows only small parts of them, and they add only
small parts of their counts.

Another variable in `acs_sf` is mapped by changing `variable`. The argument
`variable_family` only chooses the colors; its values are those of the
`estimand_family` column of a `cacs_run()` result: `"spatial_total"` (a count,
the default), `"median_proxy"` (a median), `"area_weighted_scalar_proxy"` (a
per-person value), and `"derived_rate"` (a rate). The function does not check
it against `variable`. Median household income, for example, is mapped with
`variable = "B19013_001"` and `variable_family = "median_proxy"`.

For a median or a per-person value, the estimate that `cacs_run()` gives for
the drive-time area is an average of the tract estimates weighted by area
shares, so the products in the pop-ups do not add up to it. The column
`estimand_family` of the result shows how each variable was combined;
`vignette("methodology", package = "catchmentACS")` lists the ACS tables that
are read as medians or per-person values and what happens to a median from
another table.

## Stage 4: the five rates

`cacs_plot_site_rates()` draws the site and the outline of its drive-time
areas. Clicking the site opens a pop-up with the site's values in `run_result`
for the five rates of `cacs_acs_default_rates`: `poverty_rate`, `snap_rate`,
`ssi_rate`, `unemp_rate`, and `labor_force_participation`. For each rate, the
pop-up shows the estimate and its MOE as proportions and the numbers of tracts
combined for the numerator and the denominator. It also shows the formula used
for the MOE and whether the chosen formula could not be used (`moe_fallback`).

```{r}
#| label: stage4
#| eval: !expr requireNamespace("leaflet", quietly = TRUE)

cacs_plot_site_rates(
  site_id    = one_site,
  iso_sf     = walkthrough$iso_sf,
  run_result = walkthrough$run_result,
  sites_df   = walkthrough$sites_df
)
```

Here the numerator and the denominator of each rate combine the same tracts,
and no rate has a fallback. The MOE of every rate comes from the ratio formula
(`general_ratio_conservative`), the formula for a ratio whose numerator is not
part of its denominator. `cacs_run()` uses it for all five rates by default,
although the numerator of each is part of its denominator.
`vignette("theory-derived-rates", package = "catchmentACS")` describes the
formula for that case, the proportion formula, and the argument
`formula_dispatch` that chooses between the two. Like every margin of error
from catchmentACS, the margins of error of the rates treat the tract estimates
as independent and include no error from area weighting or from the drive-time
area itself (see the Limitations section of
`vignette("methodology", package = "catchmentACS")`).

A rate with `moe_fallback = TRUE` has an asterisk before its name in the
pop-up, and a value that is `NA` shows as `n/a`. The help page of
`cacs_plot_site_rates()` says when the asterisk appears, and that of
`cacs_derive_rates()` lists the cases in which a rate is `NA`. When
`run_result` has rates for several drive times of the site, the pop-up lists
each rate once for each drive time and does not show the drive times. A
`run_result` narrowed to one drive time, such as
`run_result[run_result$drive_time_min == 10, ]`, gives one entry for each rate.

## All four maps in one call

`cacs_plot_site_pipeline()` calls the four functions above with the same site
and map settings and returns the four maps in a list of class
`cacs_site_plot_pipeline`:

```{r}
#| label: pipeline
#| eval: !expr requireNamespace("leaflet", quietly = TRUE)

maps <- cacs_plot_site_pipeline(
  site_id    = one_site,
  iso_sf     = walkthrough$iso_sf,
  tract_sf   = walkthrough$tract_sf,
  acs_sf     = walkthrough$acs_sf,
  run_result = walkthrough$run_result,
  sites_df   = walkthrough$sites_df,
  variable   = "B17001_002"
)
names(maps)
```

Printing the list at the console gives a short summary and draws no map. Each
element is a leaflet map, which is drawn when the element is printed:

```{r}
#| label: pipeline-display
#| eval: !expr requireNamespace("leaflet", quietly = TRUE)

# The map of cacs_plot_site_rates(), as in Stage 4
maps$rates
```

The elements are HTML widgets, so `htmlwidgets::saveWidget()` can save one as
an HTML file, and a Shiny app can show one with `leaflet::leafletOutput()` and
`leaflet::renderLeaflet()`. `?cacs_plot_site_pipeline` describes the
arguments, among them `drive_time_min`, which chooses the drive time of the
second and third maps for a site with several drive-time areas.

```{r}
#| label: restore-options
#| include: false

options(old_options)
rm(old_options)
```
